### 2016-12-04 treelapse.Rmd
Main report doc

---
title: "Interactive Visualization of Hierarchically Structured Data"
author: "Kris Sankaran"
date: "`r Sys.Date()`"
output:
	bookdown::pdf_book:
bibliography: ["bibliography.bib"]
link-citations: yes
---

```{r set-chunk-opts}
knitr::opts_chunk$set(fig.align = "center", out.width = "200px")
```

## outline ##

This is an outline for the treelapse article I want to write.

Introduction

- Give core theses
    + Using principles from the data visualization literature, we have designed
      a tool that we hope will be useful for analyzing tree structured data,
      especially hierarchically structured collections of time series.
    + Focus / context and linking are central to each of our visualizations
    + We hope our implementation makes this specific tool more widely
      accessible
- Describe paper structure

Problem Motivation

- Microbiome motivation
      + The meeting of ideas from medicine and ecology
      + The differential abundance problem: treatment vs. control in
        experiments
      + The ecological dynamics problem
             * Are there niches?
             * Is there competition?
- Abstraction
      + Random variables at nodes
	  + In differential abundance, we have a vector of counts, one coordinate
        per group
	  + For microbial dynamics, we have a time series at nodes
	  + In these applications, we define values at nodes as linear combinations
        of descendants (sum or average). This sort of constraint is not
		necessary though
- Problems with related structure
      + Spatial hierarchy when working with time series
      + Creating trees
             * Hierarchical clustering
             * CART
             * When variables don't have intrinsic order, but have comparable
               ranges, can interpret time series as parallel coordinates

Solution Principles / Background Literature

- Focus and Context (summary then examples)
      + Degree of interest -- a way of navigating graph structures. Modularize
        the problem of graph layout and interaction. One part is determination
        of a degree of interest for nodes. The other is determining a layout
        given a particular degree of interest distribution. Blocks are trimmed
        away depending on their average degree of interest.
      + Timeboxes -- A way of querying the types of shapes that appear in a
        given collection of time series. You draw boxes, and it highlights
        series that pass through them. The focus is on highlighted series, the
        context is the collection of series in the background.
      + Visual scent -- Determine pan and zoom based on a smaller version of
        the larger figure. Similar to interactively adjusting bounds
        numerically, but more intuitive. Focus is on the part of the scent that
        is highlighted, context is the rest of the scent.
- Linking (summary then examples)
      + Linked scatterplot brushes: Let you see many dimensions for the same
        few points. Conditional on having high values for one series, what are
        values for other variables? This is analogous to timeboxes on parallel
        coordinates plots.
      + Visual queries / conditional interpretation: Power is in querying one
        variable and seeing values in the other.
      + XGobi: Historical example of linking views, grounded in
        multidimensional scaling

- Participation in broader R software ecosystem
    + Can iterate through many interactive visualizations (like in ggvis)
    + Limited scope, but easier to maintain and more likely to remain relevant
    + Rely heavily on htmlwidgets
         * Also note similarity with networkD3: though, we have focused more on
           principles for a specific problem setup, rather than constructing
           foundational building blocks (in a way, this is more applied)

Specific Proposals

- DOI Trees
     + definition and (microbiome) use case: for a single sample, where are
       most bacteria?
     + principles it relies on: just DOI tree.
     + This is the only visualization that is completely thought through from
       before (but which as far as we can tell hasn't had easily accessible
       implementations).
     + well not exactly...we are also encoding abundance in edge widths / node
       sizes
- DOI Sankeys
     + Definition and use case: treatment vs. control in tree structured data,
       but using DOI principle
     + Exploratory analog of structSSI
     + Minor note: we use the colorbrewer palette for simpler readability
- Timebox Trees
     + Combines a few more ideas: linking and focus + context
     + definition and explanation of how these ideas being implemented
              * timebox recap: we're able to highlight lines passing through
                intersection of boxes (this is focus and context)
              * Link the display between tree and time series. Interpretation:
                Position on tree conditional on time series
                shape. Alternatively, database query on one field, returning
                over all of them.
              * Focus and context / pan and zoom using visual scent [so we have
                two focus and context ideas, and one linking idea, in this one
                display]
      + Primary use case this is designed for -- we notice a pattern in how a
        clump of microbial series look like, and we try to figure out which
        subtrees have this behavior (or, if that behavior is appear at
        different places on the tree).
- Treeboxes
       + Describe as the natural analog of timebox trees
       + Intentionally avoided queries on both tree and time series
       + Example use case: we can see how different parts of the tree have
         different series.
               * Or, we can look at all the phyla / orders at once
               * Or, we can observe if there is any competition within a
                 particular clade
       + Conditional on coming from a part of a tree, what are the observed
         series?

Case Studies

- Introduction
      + Practically, want to provide example output and how to interpret it
      + This is only a partial goal (there are many vignettes that describe
        this in more detail and are explicitly designed as tutorials) -- more
        of the focus here is on explaining how the visual principles described
        above can substantively improve data analysis in this problem domain
      + Purpose is to motivate potential workflows that incorporate the
        treelapse package. Show how it can be incorporated into any existing R
        data analysis workflow (this was all done in a single R document)
      + Argue that the problem of navigating hierarchical collections of time
        series is actually more generic than you might think

- Antibiotics
      + Study background: (check out abstract to confirm) How does microbiota
        respond to a time course of antibiotics. Especially relevant given
        growing understanding of importance of microbes in *health* (not just
        disease)
      + Framing as a timebox / treebox problem
      + Initial view: can clearly discern decrease across all microbes
      + Selecting some series, notice that the Firmicutes are more strongly
        affected
      + Can zoom into period around perturbation, using the scent idea
      + Can rerun using node averaged data -- this is only possible because we
        have linked it with R (it would not be possible in a standard d3 Block,
        for example)
      + Ideally: comparison with gPCA? Show how it complements.
- DOI Sankey: Preterm births
      + Basic goal of the study -- are certain microbiome signatures associated
        with preterm births?
      + Formal inferential approach was taken by the original publication,
        testing each microbe for an association with term / preterm
      + Also had an unsupervised analysis: cluster into CSTs
                - We can follow-up this unsupervised work using treelapse (this
                  is analogous to gPCA followup)
                - They used a heatmap, which gets the main points across, but
                  without the tree context (or allowing interactive focus +
                  context)
      + We immediately note that CST 4 has much more diversity
      + navigating down the tree, we notice that the main way the other CSTs
        are different is in the different types of Lactobacillus
- Housing prices
      + A non-microbial example, but with relatively obvious hierarchical
        (spatial) structure
      + Basic goal: How were home prices affected by recession, and how do
        prices and trends in prices differ across regions?
      + Findings from timebox / treebox:
                 * Can see quickly which neighborhoods are richer than others
                 * Can also see that poorer areas were more strongly affected
                   by the recession
                 * Can also see the variation in home prices within individual
                   counties
      + Limitations: We never actually see the spatial locations for the
        series. Can easily imagine a linking between the map and the original
        series (cite xgobi), but this is out of scope (at least for now)
- Bikesharing
      + A non-microbial example, and with non-obvious hierarchical structure
      + To induce hierarchical structure, we perform classification using a
        supplementary data table
                * We take advantage of the fact that this is essentially a two
                  table problem (one whose rows are different time series,
                  another giving the supplemental data)
                * Perform classification using classification trees (similar to
                  Kaggle, but not quite -- we treat a whole series as a sample,
                  not just a single time point).
       + Observations
                * Quickly notice difference between weekend and weekday
                * Also more subtle differences though -- temperature and year
                  effects, for example
                * The 3AM people on warm weekends (importance of focus)
       + Ideal scenario: Develop model using ideas from here, relate model to
         hierarchical structure again. Alternatively, fit a tree on correct and
         incorrect cases from another algorithm, inspect what are features of
         incorrectly classified cases, build these into new features.
- Potentially: genetics and hierarchical clustering?
- Summary: Tree structured data and collections of related time series are
  common than you might suspect, the principles of data visualization actually
  facilitate navigation of high dimensional data, and since it's in R, it's
  easy to alternate between interactive visualization, data regeneration /
  preparation, and statistical modeling.

Conclusion

- Potential future work
   + Arrangement
        * Cognostics for arranging views
        * Faceting timebox / treebox views
        * Linking with DPCOA / gPCA? In practice, microbiome literature cares a
          lot about these ordinations, we can try to explicitly tie them
          together in one visualization (rather than iterating, like we do
          now).
   + Incorporating the DOI idea into the timebox trees?
   + Other layouts? Graphs? DAGs?
   + Grammar
        * Is it possible to modularize the key ideas, so we can create similar
          kinds of linked and arranged views without so much effort?
- Summary
    + Developed an R package based on ideas from the visualization community
    + Think there is power in interactive visualization for hands on data
      analysis, not just polished public tools
    + Think there is a relationship between statistical modeling and data
      visualization: it can be more useful to visualize raw data in relation to
      models
    + Focused on solving a very specific methodological problem, rather than
      building a generic system. But this methodological problem appears in a
      variety of domains.

## text ##

## Introduction ##

In this work, we describe methods for visualization of data structured along
trees, especially hierarchically structured collections of time
series. Influenced by Degree of Interest trees, linked scatterplot brushing,
and timeboxes, we have adapted the fundamental visualization principles of
"focus + context" and "linking" to study questions that arise when analyzing
tree structured data. Our motivating application is to the analysis of
microbial time series, where an evolutionary tree relating microbes is
available a priori. However, we have identified common problem types where, if
a tree is not directly available, it can be constructed from data and then
studied using our techniques.

We have implemented our visualizations in d3, but encapsulated in an R package,
to facilitate rapid turnover from data preparation and modeling to interactive
exploration, and vice versa. We hope this package encourages data analysts to
work at this border of data modeling and visualization, and more generally
empowers a wider audience to apply less widely known, but powerful,
visualization ideas.

The paper is organized as follows. First, we describe our microbiome
application, and then abstract away the essential analysis tasks. This
abstraction suggests applications to standard problems outside of the
microbiome, some of which we articulate here. Next, we review the underlying
visualization principles behind our work, and provide references to the
software systems that originally implemented them. We then connect these
principles to analysis tasks we identified earlier, describing in detail the
visualization methods we have implemented in treelapse. We close with several
case studies using publicly available data.

### Problem Motivation ###

Microbiomics is the study of the communities of microbes living in given
environments, for example, ocean water or the human gut. Progress in the field
has been rapidly accelerated by the advent of genomic technologies, which
enable detailed quantification of microbial community structure, and as the
relationship between the microbiome and human health has become more and more
apparent, research into the human microbiome has garnered widespread
attention. Being concerned with both microbial community structure and human
health, the field exists at the border between ecology and medicine;
consequently, papers in the area often apply a blend of exploratory data
analysis and formal statistical inference ideas.

The two essential microbiome analysis problems that motivated our work are the
tree structured differential abundance and microbial dynamics problems. In the
differential abundance problem, we attempt to compare the abundances of
individual microbes across experimental conditions -- for example, treatment
vs. control or healthy vs. diseased. This is the microbiome analog of
differential expression analysis in genomics. We prepend the description
"tree-structured" because, in practice, researchers generate interpretations
about middling taxonomic orders of microbes -- it is more interesting to
discover novel behavior taxonomic levels intermediate between high-order phyla
and low-level sequence variants. Hence, we frame the tree-structured
differential abundance problem as the question of identifying the largest
taxonomic subtree whose associated microbes are differentially abundant.

In the tree structured microbial dynamics problem, the goal is to describe
changes in microbial abundances in an environment over time. As in the
differential abundance problem, it is useful if these descriptions can be given
at the highest subtree at which the pattern appears. Specific questions of
interest often have an ecological flavor. For example, researchers are often
interested in understanding how microbial populations respond to sudden or
gradual environmental changes or how microbes fill, drop out from, or compete
for certain environmental niches. Medically, these questions are important for
illuminating the impact of antibiotic time courses or the potential for direct
manipulation, for example.

### Problem Abstraction ###

An abstraction that unifies thinking through both the tree structured
differential abundance and microbial dynamic problems is to identify the data
with a collection of random variables indexed by nodes in a prespecified tree
structure. In the differential abundance problem, each random variable itself
lives in $\mathbb{R}^{G}$ where $G$ is the number of groups being
compared. Each coordinate represents the average value across one of the
groups, and the node exhibits differential abundance when the coordinates are
not drawn from the same distribution. In the microbial dynamics problem, each
random variable is instead a time series, living in $\mathbb{R}^{T}$.

In both these applications, we constrain the values of parent nodes according
to the value of the children nodes: we define the value at each node to be
either the sum or average of all descendant tips, but it is possible to imagine
situations where the internal nodes are drawn from their own distribution,
unconstrained by descendants. In general, analysis in this abstraction focuses
on describing the distribution of these random variables as a function of their
position in specified subsets of the tree. The essential difficulty in these
problems is high-dimensionality -- there are many tree nodes, each holding a
vector-valued random variable. Even simply navigating across the tree and
comparing coordinates in the observed variables is a challenge; ideally we
could construct a succinct representation of the essential covariation across
subtrees and coordinates.

This framework suggests other potential application areas, not all of which
have a priori known tree structures. For example, collections of spatially
indexed time series are frequently encountered in practice -- consider energy
consumption, product sales, or high school dropout rates across regional
districts. This type of data has an implicit tree structure -- at the top level
are different states, while at the bottom are individual census tracts,
say. Analysis here revolves around the question of how variation across series
is related to their geographic position.  An example using this type of data is
given in the Zillow home prices case study in Section (todo: insert section
number).

Alternatively, if this type of hierarchical contextual information is not
available, a tree structure can be learned directly from the data. This can be
most directly achieved by learning a hierarchical clustering on the original
series. If there is contextual information, but it is not hierarchical, it is
possible to setup a supervised problem that uses context to predict features of
the time series. We can retrieve a tree by using a regression tree directly
(cite: rpart) or extracting a regression tree from a more complex supervised
model (cite: regression tree extraction methods). Analysis then focuses on how
different partitions of the covariate (contextual) space relate to observed
time series. This approach is described in the UCI bikesharing demand case
study below.

Finally, note that, while we have focused on time series valued nodes, all of
this discussion could be translated to studying high dimensional data via
parallel coordinates (todo: cite parallel coordinates). The usual parallel
coordinates challenges remain, mainly selecting scales for and ordering the
different coordinates, but the linking and focus + context can still be
employed this setting.

## Background Literature and Solution Principles ##

Now that we have specified the essential questions of interest, we survey some
ideas from the visualization literature that can be applied to answer
them. Again, the core difficulty is one of high dimensionality, so it should be
no surprise that the techniques we adapt come from the literature on
high-dimensional data visualization, namely, work on the focus + context and
linking.

The focus + context principle is that it is possible to study large collections
of visual elements at multiple scales, and these scales can be juxtaposed by
"focusing" on a few elements of interest while maintaining the "context" of the
coarser-scale background. A simple example of this idea is to include a search
box that highlights samples corresponding to matching terms (the focus) while
muting those that don't (the context). Two more sophisticated methods anchored
in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are central
to the proposals in treelapse \@cite(hochheiser2004dynamic,
heer2004doitrees). In timeboxes, a collection of time series are graphically
queried using interactive brushes. Series that pass through all of the
user-specified brushes are highlighted, and the rest are faded into the
background. Hence, time series meeting the constraints imposed by the brushes
are focused, while the remainder remain present, though de-emphasized, as
context. This method can be interpreted programmatically as the visual version
of a database query, or probabilistically as the conditional distribution for
the full series, given it passes through certain bounds.

In DOI Trees, the viewer's attention is focused on a collection of
high-interest nodes, while the remaining lower-interest nodes are left on the
fringes (or entirely hidden) as context. The implementation is modularized into
two tasks -- the determination of a DOI distribution over nodes in the tree and
visual layout of a tree given DOI assignments. The DOI distribution used in
\@cite(heer2004doitrees) places maximal interest on nodes that the user clicks,
along with all ancestors. The DOI for all other nodes is defined as the
distance to the closest maximal interest node. The layout proposed here trimmed
low-interest subtrees until the remaining nodes all fit within the screen
size. By changing the minimal DOI below which nodes are hidden, the user can
transition between node-specific and full-tree scales.

In linking, alternative representations of the same samples are placed side by
side in a way that suggests covariation across the representations. A canonical
application is to linked scatterplot brushing \@cite(becker1987brushing). Here,
a scatterplot matrix is constructed displaying the relationship between all
pairs of variables.  Points brushed in one scatterplot are then highlighted in
all others. For example, this helps the user determine whether an outlier in
one dimension is an outlier in others. Another instance of this idea is in
linking the results of dimensionality reduction methods to displays of the raw
data, as implemented by XGobi and Cranvas, for example. As in timeboxes,
linking ca ben interpreted as database queries or conditional probabilities:
given a subset of the series after conditioning on the values for one set of
features, what are the values for a second set \@cite(buja1996interactive)?

Finally, while not exactly a fundamental visualization principle, we note that
our work is deliberately grounded in the R software ecosystem. Not only does
this make these methods more broadly accessible, we hope to encourage a more
animated exchange between data preparation and modeling with interactive
visualization. Moreover, our tools are intentionally limited in scope --
designed to facilitate this dialog for a specific class of problems, rather
than providing a toolbox for generic types of visualization design. We believe
that this narrow context within a broad ecosystem strikes a balance between
problem-specificity and ease-of-use.

## Specific Proposals ##

Our first proposed visualization technique is a minor modification of the DOI
Tree idea. The standard DOI Tree definition does not have any notion of data
defined at nodes, it is only used a device for navigating tree structures. A
trivial extension can encode scalar data at nodes: make the node radius reflect
the associated scalar value. To reinforce this effect, we also encode this
scalar value in the width of the parent edge. When parent nodes have values
equal to the sum of their children, this creates the effect of values "flowing"
from the root to leaves. To help viewers make use of their domain knowledge, we
have included a search box that highlights paths to matching node terms. Edges
are ordered from widest on the left to narrowest on the right. While this
method can only be used to visualize a single scalar-value per tree node, it
suggests an approach to the tree-structured differential abundance problem,
which we call the DOI Sankey.

In the DOI Sankey, we split each edge in the DOI Tree across different
groups. For example, suppose we have the average counts for treatment and
control groups at each tree node. Every edge in the tree is split into two
colors^[We use the colorbrewer palette to facilitate readability], with the
relative widths of the different colors reflecting the difference in sizes for
the two groups. The overall width of each edge is represents the sum of values
across all groups.

This display is designed to facilitate investigation of the tree structured
differential abundance question. For example, for a single node and a single
group, we can compute the average abundance at that node among all samples in
that group. This will give the width for that group's color on the edge leading
to the specified node. Differentially abundant subtrees then correspond to
subtrees where some colors occupy more space than others. That is, this
representation makes it easier to identify points where the "flows" for
different groups diverge -- the colors begin to separate. The DOI principle
assists navigation across the tree structure, allowing focus on individual flow
structures without losing broader tree context.

Our third display is directed at the microbial dynamics question instead. It
ignores the DOI principle, placing emphasis rather on linking and the
representation of full time series. Here, two panels are arranged one over the
other; one displays all time series, while the other displays all tree nodes,
with node sizes reflecting the average time series value at that node. For this
reason, we call the display, Timebox Trees. In the time series panel, we have
directly implemented the timeboxes idea. Our link is between the two panels:
when a set of series is highlighted by the timeboxes, the associated tree nodes
are also highlighted. For example, timeboxes can be used to focus on a set of
series with specific shape -- increased abundance after an ecological shock,
for example -- and then identify along what subtrees this pattern is
present. To further focus attention, a pan-zoom scented widget is provided in
the top right [@willett2007scented]. The widget is a miniature version of
the full time series panel, equipped with a single brush whose extent specifies
the limits in the main time series panel. As in the DOI trees and Sankeys, a
search bar can be used to highlight of interest a priori.

The final display currently implemented in the package is the natural converse
of the Timebox Trees display. Rather than defining visual queries in terms of
time series, we define queries using nodes in the tree. For this reason, we
call the display Treeboxes. Rather than focusing on the intersection of
brushes, as in Timebox Trees, we focus on the union of brushed over nodes. This
allows us to highlight the series associated with nodes on distant
subtrees. This display is also suited for the microbial dynamics problem. For
example, by highlighting all nodes at one taxonomic level in the tree, we can
easily summarize the time series pattern for all the taxa at that
level. Alternatively, focusing on all the children below a single node makes it
possible to see how much correlation and competition there is between
taxonomically similar microbes. As in the Timebox Trees display, a search box
and pan-zoom scented widget is provided.

## Case Studies ##

We now delve into concrete examples of applying treelapse to real
data. Our goals are is to illustrate potential workflows that
incorporate treelapse, describe the formulation of questions that can
be naturally investigated with our methods, and provide example
interpretations for specific treelapse output. Our examples are also
chosen to reflect the range of problem domains to which the package
can be applied -- though it was motivated by applications to the
microbiome, it is not tied to it. More importantly, we argue that the
visualization principles described earlier can substantively improve
the practice of data analysis in the class of problems we have focused
attention on.

### Microbial Dynamics of Antibiotics Time Courses ###

[@dethlefsen2008pervasive] investigates the effect of antibiotics on microbial
community composition from an ecological perspective. The study tracks the
microbiome of *x* patients across *y* weeks (todo: give the numbers), with
antibiotic two time courses introduced during * and *, respectively. Discerning
the variation in resilience across microbes is important, considering the
community's increasing acceptance of the idea that microbes are central to human
health and not just disease.

We approach the data using the linked time and treebox views, after first
filtering low variance taxa and taking an $asinh$ transformation. An initial
view, Figure \@ref(fig:antibioticoverview), reveals two dramatic drops in the
overall microbial abundance time series at the time of the antibiotics time
courses. Two more subtle effects are also suggested from this view,

* The second antibiotic treatment seems to have a more lasting effect, as the
series take longer to return to their original values.
* Some high level taxa appear relatively unaffected by the first
antibiotic treatment.  By more closely inspecting the display, we are
able to identify these as members of the Bacteroidetes phylum, see
Figure \@ref(fig:antibioticbacteroidetes).

```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
knitr::include_graphics("antibiotic_overview.png")
```

```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
knitr::include_graphics("antibiotic_bacteroidetes.png")
```

Next, using the scented widget, we focus on the window around the
second antibiotic treatment. We apply the treebox display to compare
then behavior of different families of Firmicutes, Lachnospiraceae and
Ruminoccocus. We suspect these taxa are associated with the delayed
recovery after the second time course. To inspect this, we these
family names in the search box to isolate their positions on the tree,
and then we apply brushes to highlight the series that contribute to
these higher-level families. The resulting view is given by Figure
\@ref(fig:antibioticfirmicutes)

```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
knitr::include_graphics("antibiotic_firmicutes.png")
```

Alternatively, we can summarize each node by the average of its descendants --
this brings attention to individual microbes that may be underlying some of the
broader taxonomic patterns we have noted when studying the subtree sums. For
example, in Figure \@ref(fig:ruminococcus), we highlight all families below the
Ruminococcus, suggesting that the decrease due to antibiotics occurs uniformly
across almost all families. A point that was not evident in the earlier
sum-across-descendants view is that, after the second treatment of antibiotics,
a few of the Ruminoccocus families recover more rapidly than the rest, for
example the Unc095d3 (highlighted in red) are only briefly affected. In
contrast, most families seem to recover in unison after the first treatment.

```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders, with the except of "}
knitr::include_graphics("ruminococcus.png")
```

Further, note that in this subtree averages view, the tree display has
changed. This is because, at each branching point, we place the node
with larger average value on the left. This places the
Verrucomicrobiae at the far left, which seem to have large abundance
over many time points. This phylum had been previously obfuscated,
because there are not many leaves associated with this phylum, hence
the sum was small. Interestingly, the abundance of these microbes
seems to *increase* after the first antibiotics treatment.  Be
cautious, however, that the average over only a few Verrucomicrobiae
species will be a more variable estimate than the averages over the
more prevalent phyla.

```{r verrucomicrobiae, fig.cap = "The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant. In particular, they seem to increase after the first antibiotic time course."}
knitr::include_graphics("verrucomicrobiae.png")
```

### Differential Microbial Abundance and Preterm Births ###

The study [@digiulio2015temporal] tracked the abundance of microbes in the
during vaginal microbiome during pregnancy in an effort to study relationships
between microbial community composition and preterm birth. Ideally, it would be
possible to develop clear microbial signatures associated with preterm births.

Unlike the antibiotics study, we have measurements across more individuals than
we could reasonably inspect one at a time. While we could average across all
individuals, we will take our cue from [@digiulio2015temporal] and place each
sample into one of five Community State Types (CSTs), identified via k-medoids.
Using a linear model, one of these CSTs (CST 4) was found to be more diverse,
further it appeared associated with preterm births. Here, we corroborate this
finding using exploratory views.

Therefore, our focus here is on the differential abundance question,
rather than dynamics.  We would like to provide visual representations
of differential abundance across CSTs and also between preterm and
non-preterm births. [@digiulio2015temporal] interpreted its CSTs using
a heatmap, with microbes ordered according to a hierarchical
clustering.  By using the DOI Sankey instead, we can interpret the
CSTs in their taxonomic context and at multiple scales of taxonomic
resolution. Further, while [@digiulio2015temporal] focused on
identifying associations between preterm births and CSTs -- presumably
because testing individual microbes loses power -- we can compare
microbe abundances between preterm and non-preterm samples along
subtrees, without requiring CSTs as an intermediary.

In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
their values along the subtree. Specifically, we took the average of
all samples within each CST to define values at the (microbe-level)
leaf nodes, and then aggregated the counts up to the root. It is
immediately clear that samples from CST 4 have much more taxonomic
diversity. Further, focusing on the Lactobacillaceae family, we note
that the differential abundnace of these microbes distinguishes th
remaining CSTs, see Figure \@ref(fig:pretermcstslacto).

Alternatively, in Figure \@ref(fig:preterm), we avoid working with
CSTs, displaying instead averages among samples associated with either
preterm or term births. The green edges are associated with preterm
births -- we see that they contribute more weight to Phyla outside the
Firmicutes. This is consistent with the claim that CST 4, the most
diverse of the CSTs, is associated with preterm births.

```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
knitr::include_graphics("preterm_csts.png")
```

```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
knitr::include_graphics("preterm_csts_lacto.png")
```

```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
knitr::include_graphics("preterm_preterm.png")
```

### Dynamics in Housing Prices ###

We next consider an application unrelated to the microbiome, but with
relatively clear hierarchical structure. Our data are downloaded from
Zillow^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
and give the Zillow Home Value Indexes at the neighborhood level,
across the country, given monthly between 1996 and 2016. In our
display, we have taken the base 10 log of these indexes. As our
hierarchical structure, we use each neighborhood's assignment to
state, regional, county, and city levels. We represent each of these
coarser spatial categories using the average of all neighborhoods
contained in them. We have filtered down to the 890 neighborhoods in
California; rendering more neighborhoods while keeping all 246
timepoints causes the interface to lag^[See the Future Work section
for potential optimizations, however.].

Our basic analysis revolves around geographic and temporal variation
in home prices. We are especially interested in the effect of the 2008
recession and any variation in the lead-up or recovery. These
questions can be naturally framed using timebox trees and treeboxes.

We first study variation in prices at the city level. These series can
be highlighted using a single brush in treeboxes, since each spatial
level is displayed at the same height in the tree. THe resulting
display is given in Figure \@ref(fig:zillowcities). From a high level,
the housing bubble, decrease in prices due to the recession, and
subsequent recovery are readily apparent. The city-level view makes it
clear that not all neighborhoods were equally affected by the
recession -- richer cities plateaued at their 2008 prices,
middle-income cities saw moderate decreases, and poorer cities saw the
most significant declines in home prices. These more significant
declines tended to occur in cities that had recently seen rapid
increases during the housing bubble. Finally, we note that the range
in home prices seems to have increased after the recession, speaking
to the differential long-term effect of the recession on prices.

Alternatively, we can consider the trajectories of home prices among
neighborhoods, conditional on their being that were middle-income
before the recession.  We generate the sequence of views in Figures
\@ref(fig:zillowmiddlepre), \@ref(fig:zillowmiddleup), and
\@ref(fig:zillowmiddledown) in this vein. The first of these figures
isolates neighborhoods with middle incomes before the recession, using
a single timebox. Since there appears to be a divergence in
trajectories after the recession, we introduce a second post-recession
timebox, dragging it over series with higher and lower incomes during
this second time period. This is the content of Figures
\@ref(fig:zillowmiddleup) and \@ref(fig:zillowmiddledown). Inspecting
the highlighted tree nodes associated with these series, we find that
most of the middle-income series that increased after the recession
are associated with middle-income neighborhoods within the coastal
Southern California counties. In constrast, those middle income series
that saw decreases were mostly located in Central California and
Oakland.

The previous analysis highlights the fact that, within even narrow geographic
regions, there can be substantial variation in prices. We can study this
directly using treeboxes. In Figure \@ref(fig:zillowsf) we have highlighted all
series in San Francisco County. We see that, in 2016, prices range from around
$10^{5.6} \approx \$400,000$ to $10^{6.7} \approx \$5 \text{million}$. So, while
all these neighborhoods tend to be among the more expensive ones in California,
prices can vary in a non-smooth way across geographic space.

We conclude this example with a caveat that the Zillow data are not
representative of all neighborhoods in California, only those with
enough listings on the site, so should be supplemented by other data
sources for more substantial decision-making.

```{r zillowcities, fig.cap = "California home prices at the city level, between 1996 and 2016. The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
knitr::include_graphics("zillow_cities.png")
```

```{r zillowmiddlepre}
knitr::include_graphics("zillow_middle_pre.png")
```

```{r zillowmiddleup}
knitr::include_graphics("zillow_middle_up.png")
```

```{r zillowmiddledown}
knitr::include_graphics("zillow_middle_down.png")
```

```{r zillowsf}
knitr::include_graphics("zillow_sf.png")
```

### Sources of Variation in Bikesharing Demand ###

Our next example is a study in bikesharing demand, included as an
example of analyzing collections of time series when there is no
obvious hierarchical structure a priori. The data are downloaded from the
[UCI Repository](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset) and
were originally collected by a Washington D.C.-based bikesharing
system for use in a Kaggle prediction competition. The data are hourly
measurements of bike demand, aggregated across all bikesharing
stations, over two years, along with some supplemental weather
data. In the competition, participants were asked to predict the
hourly demand on a held-out test set. Here, we adopt a descriptive
view instead, attempting to characterize factors associated with
variation in bikesharing demand.

Like the Zillow home prices application, we study this problem as one
of describing a large collection of related time series. Here, we
consider the demand during a single day to be one time series; this is
a natural choice considering the daily periodicity of bike demand. To
relate these daily series according to an interpretable tree
structure, we apply a regression tree relating the supplemental data
to the bikesharing demand (J.  Friedman, Hastie, and Tibshirani
2001). In more detail, we build this tree by noting the “two table”
structure of this problem: one describes bike demand, the other holds
the supplemental data. In both, the rows index days, while the columns
index hours and supplemental features for the two tables,
respectively. Our tree is the trained regression tree after predicting
demand at 8AM based on the supplemental data. We choose this response
because (1) we need a univariate response in order to apply regression
trees and (2) the more straightforwards reduction to
daily-average-demand fails to distinguish between weekdays and
weekends, whose series appear qualitatively very different from each
other.

Given this response, the first split in the regression tree is
(unsurprisingly) the difference between weekends and weekdays. This is
emphasized in Figures 14 and 15, respectively; using timeboxes to
isolate the two types of series highlight the left and right sides of
the tree, respectively. For a mroe subtle effect, we select the
internal nodes associated with the first split below the weekday
vs. weekend split; these are given in Figures 16 and 17. This suggests
that more commuters became aware or interested in bikesharing during
the second year of data collection.

To contrast with these general questions about daily demand, we ask a
more granular question about demand at night: what characterizes days
on which there is larger than average demand after midnight? We can
select these series after first zooming into this time window. Figure
18 reveals that the highlighted series are associated with the
warm-weekend split, which seems quite reasonable in retrospect.

Finally, we can study the behavior of the regression tree itself using
the DOI Sankey. Here, we group samples according to their quintile of
8AM demand and then count the abundance of the groups flowing down
different branches. We find that the quintiles are each rather
strongly separated after descending even a few steps down the
regression tree – for example, in Figure ?? focuses on 2011 vs. 2012
split among weekday samples, showing that this split distinguishes
between samples falling in the second and third quintiles of 8AM
demand.

This interactive representation of regression trees is potentially
more useful on larger trees that cannot be easily parsed in a single
view; in this sense the bikesharing tree is relatively
simple. Further, in light of research on approximating more complex
supervised learning algorithms using decision trees (Boz 2002, Saito
and Nakano (2002)), this approach of modeling and interpreting
two-table data could be used in conjunction with more sophisticated
regression methods. In our ideas data analysis workflow, we imagine
the analyst applying interactive visualization and modeling techniques
in an iterative, nonlinear fashion.

## Conclusion and Future Work ##

Here, we have reviewed some fundamental principles of data visualization and described their implementation
in a new treelapse package. Further, we have provided examples of the practical usefulness of these
principles in real-world data analysis situations.

This package has only developed some of the most basic ideas, and
there are a number of potentially useful extensions worth
exploring. For example, we have not considered the principle of
arrangement in our visualizations (Buja, Cook, and Swayne 1996),
though many of our conclusions were based on comparing alternative
selections of the same display. We could imagine faceting our displays
across groups to make these types of comparisons more
accessible. Further, we have only worked with the DOI distribution
described in (Heer and Card 2004). It would be interesting to define a
more statistical notion of interest along nodes, based on cognostics,
for example (Hafen et al. 2013, J. H. Friedman and Stuetzle
(2002)). Further, it could be helpful to link the visualization-based
workflow described in this work with more traditional multivariate and
inferential statistics workflows – it is easy to imagine an
interactive version of ordinations that display representations of the
raw data upon selection. A simple extension could be to allow graph
layouts instead of trees in time and treebox displays, for data that
cannot be coerced into a hierarchical structure. Finally, if these
ideas turn out to be useful in practice, it would be valuable to
modularize the basic visualization layouts and relationships into a
library, allowing the wider community to construct linked, interactive
graphics with minimal effort.

All these potential extensions aside, we have built an easily
accessible R package for visualization techniques in a very specific
methodology problem – analysis of differential abundance and dynamics
in hierarchical structured data – that appears in a variety of
application domains. We have leveraged (Vaidyanathan et al.  2014) to
create exploratory visualizations while still deep in data analysis;
in this way our work is a departure from the culture of polished,
journalistic visualizations prioritized by the D3 community. Finally,
we have given a series of examples to demonstrate how the general
visualization techniques of focus + context and linked brushing can be
practically incorporated into a range of practical analysis workflows,
from studying the impact of microbes on human health to better
allocating units in commuter bikesharing systems.

## References ##

### 2016-12-04 treelapse.Rmd
echo = FALSE, and revisions for first few paragraphs

@@
 ```{r set-chunk-opts}
-knitr::opts_chunk$set(fig.align = "center", out.width = "200px")
+knitr::opts_chunk$set(fig.align = "center", out.width = "200px", echo = FALSE)
 ```
@@
 
-In this work, we describe methods for visualization of data structured along
-trees, especially hierarchically structured collections of time
-series. Influenced by Degree of Interest trees, linked scatterplot brushing,
-and timeboxes, we have adapted the fundamental visualization principles of
-"focus + context" and "linking" to study questions that arise when analyzing
-tree structured data. Our motivating application is to the analysis of
-microbial time series, where an evolutionary tree relating microbes is
-available a priori. However, we have identified common problem types where, if
-a tree is not directly available, it can be constructed from data and then
-studied using our techniques.
+(todo: sentence about why we should care, and a high level summary
+before diving into degree-of-interest linking jargon)
 
-We have implemented our visualizations in d3, but encapsulated in an R package,
-to facilitate rapid turnover from data preparation and modeling to interactive
-exploration, and vice versa. We hope this package encourages data analysts to
-work at this border of data modeling and visualization, and more generally
-empowers a wider audience to apply less widely known, but powerful,
-visualization ideas.
+In this work, we describe methods for visualization of data structured
+along trees, especially hierarchically structured collections of time
+series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
+linked scatterplot brushing [@becker1987brushing], and timeboxes
+[@hochheiser2004dynamic], we have adapted the fundamental
+visualization principles of "focus + context" and "linking" to study
+questions that arise when analyzing tree structured data. Our
+motivating application is to the analysis of microbial time series,
+where an evolutionary tree relating microbes is available a
+priori. However, we have identified common problem types where, if a
+tree is not directly available, it can be constructed from data and
+then studied using our techniques.
+
+We have implemented our visualizations in d3, but encapsulated in an R
+package, to facilitate rapid turnover from data preparation and
+modeling to interactive exploration, and vice versa. We hope this
+package encourages data analysts to work at the border between data
+modeling and visualization, and more generally empowers a wider
+audience to apply less widely known, but powerful, visualization
+ideas.
 

### 2016-12-04 treelapse.Rmd
Revisions to problem motivation

@@
 We have implemented our visualizations in d3, but encapsulated in an R
-package, to facilitate rapid turnover from data preparation and
-modeling to interactive exploration, and vice versa. We hope this
-package encourages data analysts to work at the border between data
-modeling and visualization, and more generally empowers a wider
-audience to apply less widely known, but powerful, visualization
-ideas.
+package, called treelapse, to facilitate rapid turnover from data
+preparation and modeling to interactive exploration, and vice
+versa. We hope this package encourages data analysts to work at the
+border between data modeling and visualization, and more generally
+empowers a wider audience to apply less widely known, but powerful,
+visualization ideas.
 
-The paper is organized as follows. First, we describe our microbiome
-application, and then abstract away the essential analysis tasks. This
-abstraction suggests applications to standard problems outside of the
-microbiome, some of which we articulate here. Next, we review the underlying
-visualization principles behind our work, and provide references to the
-software systems that originally implemented them. We then connect these
-principles to analysis tasks we identified earlier, describing in detail the
-visualization methods we have implemented in treelapse. We close with several
-case studies using publicly available data.
+The paper is organized as follows. First, we describe our motivating
+application to the microbiome and the associated generic analysis
+tasks. Next, we review the underlying visualization principles behind
+our contributions. Then we then connect these principles to analysis
+tasks we identified earlier, describing in detail the visualization
+methods we have implemented in treelapse. We close with several case
+studies using publicly available data across both microbiome and
+non-microbiome related applications.
 
@@
 
-Microbiomics is the study of the communities of microbes living in given
-environments, for example, ocean water or the human gut. Progress in the field
-has been rapidly accelerated by the advent of genomic technologies, which
-enable detailed quantification of microbial community structure, and as the
-relationship between the microbiome and human health has become more and more
-apparent, research into the human microbiome has garnered widespread
-attention. Being concerned with both microbial community structure and human
-health, the field exists at the border between ecology and medicine;
-consequently, papers in the area often apply a blend of exploratory data
-analysis and formal statistical inference ideas.
+A microbiome is a community of microbes living in given environments,
+for example, ocean water or the human gut [@human2012structure,
+@human2012structure, @cho2012human]. Progress in the field has been
+rapidly accelerated by the advent of genomic technologies, which
+enable detailed quantification of microbial ecological structure and
+its influence in human and environmental health. Being concerned with
+both microbial community structure and human health, the field exists
+at the border between ecology and medicine; consequently, papers in
+the area often apply a blend of exploratory data analysis and formal
+statistical inference.
 
-The two essential microbiome analysis problems that motivated our work are the
-tree structured differential abundance and microbial dynamics problems. In the
-differential abundance problem, we attempt to compare the abundances of
-individual microbes across experimental conditions -- for example, treatment
-vs. control or healthy vs. diseased. This is the microbiome analog of
-differential expression analysis in genomics. We prepend the description
-"tree-structured" because, in practice, researchers generate interpretations
-about middling taxonomic orders of microbes -- it is more interesting to
-discover novel behavior taxonomic levels intermediate between high-order phyla
-and low-level sequence variants. Hence, we frame the tree-structured
-differential abundance problem as the question of identifying the largest
-taxonomic subtree whose associated microbes are differentially abundant.
+The two essential microbiome analysis problems that motivated our work
+are the tree-structured differential abundance and differential
+dynamics problems. In the differential abundance problem, we attempt
+to compare the abundances of individual microbes across experimental
+conditions -- for example, treatment vs. control or healthy
+vs. diseased. This is the microbiome analog of differential expression
+analysis in genomics [@anders2010differential]. We prepend the
+description "tree-structured" because, in practice, researchers
+generate interpretations about intermediate taxonomic orders of
+microbes -- it is more interesting to discover novel behavior
+taxonomic levels between high-order phyla and low-level
+species. Hence, we frame the tree-structured differential abundance
+problem as the question of identifying the largest taxonomic subtree
+whose associated microbes are differentially abundant.
 
-In the tree structured microbial dynamics problem, the goal is to describe
-changes in microbial abundances in an environment over time. As in the
-differential abundance problem, it is useful if these descriptions can be given
-at the highest subtree at which the pattern appears. Specific questions of
-interest often have an ecological flavor. For example, researchers are often
-interested in understanding how microbial populations respond to sudden or
-gradual environmental changes or how microbes fill, drop out from, or compete
-for certain environmental niches. Medically, these questions are important for
-illuminating the impact of antibiotic time courses or the potential for direct
-manipulation, for example.
+In the tree structured microbial dynamics problem, the goal is to
+describe changes in microbial abundances in an environment over
+time. As in the differential abundance problem, it is useful if these
+descriptions can be given at the highest subtree at which the pattern
+appears. Specific questions of interest often have an ecological
+flavor. For example, researchers are often interested in understanding
+how microbial populations respond to sudden or gradual environmental
+changes or how microbes fill, drop out from, or compete for
+environmental niches. Medically, these questions are important for
+illuminating the impact of antibiotic time courses or diet changes,
+for example.
 

### 2016-12-04 treelapse.Rmd
Revisions to problem background

@@
 - Timebox Trees
-     + Combines a few more ideas: linking and focus + context
+     + Combines a few more ideas: linking and focus-plus-context
      + definition and explanation of how these ideas being implemented
@@
 [@hochheiser2004dynamic], we have adapted the fundamental
-visualization principles of "focus + context" and "linking" to study
+visualization principles of "focus-plus-context" and "linking" to study
 questions that arise when analyzing tree structured data. Our
@@
 
-An abstraction that unifies thinking through both the tree structured
-differential abundance and microbial dynamic problems is to identify the data
-with a collection of random variables indexed by nodes in a prespecified tree
-structure. In the differential abundance problem, each random variable itself
-lives in $\mathbb{R}^{G}$ where $G$ is the number of groups being
-compared. Each coordinate represents the average value across one of the
-groups, and the node exhibits differential abundance when the coordinates are
-not drawn from the same distribution. In the microbial dynamics problem, each
-random variable is instead a time series, living in $\mathbb{R}^{T}$.
+To unify the tree-structured differential abundance and microbial
+dynamic problems, we identify the data with a collection of random
+variables indexed by nodes in a prespecified tree structure. In the
+differential abundance problem, each random variable lives in
+$\mathbb{R}^{G}$ where $G$ is the number of groups being
+compared. Each coordinate represents the abundance for that group, and
+a node exhibits differential abundance when the coordinates are drawn
+from different distributions. On the other hand, in the microbial
+dynamics problem, each random variable is a time series, living in
+$\mathbb{R}^{T}$.
 
-In both these applications, we constrain the values of parent nodes according
-to the value of the children nodes: we define the value at each node to be
-either the sum or average of all descendant tips, but it is possible to imagine
-situations where the internal nodes are drawn from their own distribution,
-unconstrained by descendants. In general, analysis in this abstraction focuses
-on describing the distribution of these random variables as a function of their
-position in specified subsets of the tree. The essential difficulty in these
-problems is high-dimensionality -- there are many tree nodes, each holding a
-vector-valued random variable. Even simply navigating across the tree and
-comparing coordinates in the observed variables is a challenge; ideally we
-could construct a succinct representation of the essential covariation across
-subtrees and coordinates.
+In both of these applications, we constrain the values of parent nodes
+according to the value of the children nodes: we define the value at
+each node to be either the sum or average of all descendant
+tips. However, it is possible to imagine situations where the internal
+nodes are drawn from their own distribution, unconstrained by
+descendants. In general, analysis in this abstraction focuses on
+describing the distribution of these random variables as a function of
+their position across subsets of the tree. The essential difficulty in
+these problems is high-dimensionality -- there are many tree nodes,
+each holding a vector-valued random variable. Even simply navigating
+across the tree and comparing coordinates in the observed variables is
+a challenge; ideally we could construct a succinct representation of
+the essential covariation across subtrees and coordinates.
 
-This framework suggests other potential application areas, not all of which
-have a priori known tree structures. For example, collections of spatially
-indexed time series are frequently encountered in practice -- consider energy
-consumption, product sales, or high school dropout rates across regional
-districts. This type of data has an implicit tree structure -- at the top level
-are different states, while at the bottom are individual census tracts,
-say. Analysis here revolves around the question of how variation across series
-is related to their geographic position.  An example using this type of data is
-given in the Zillow home prices case study in Section (todo: insert section
-number).
+This framework suggests other potential application areas, not all of
+which have a priori known tree structures. For example, collections of
+spatially-indexed time series are frequently encountered in practice
+-- consider energy consumption, product sales, or high school dropout
+rates across regional districts. This type of data has an implicit
+tree structure -- at the top level are different states, while at the
+bottom are individual census tracts, say. Analysis here revolves
+around the question of how variation across time series is related to
+their geographic position.  A case study to this type of data is given
+in Section \@ref(zillow-study).
 
-Alternatively, if this type of hierarchical contextual information is not
-available, a tree structure can be learned directly from the data. This can be
-most directly achieved by learning a hierarchical clustering on the original
-series. If there is contextual information, but it is not hierarchical, it is
-possible to setup a supervised problem that uses context to predict features of
-the time series. We can retrieve a tree by using a regression tree directly
-(cite: rpart) or extracting a regression tree from a more complex supervised
-model (cite: regression tree extraction methods). Analysis then focuses on how
-different partitions of the covariate (contextual) space relate to observed
-time series. This approach is described in the UCI bikesharing demand case
-study below.
+Alternatively, if this type of hierarchical contextual information is
+not directly available, a tree structure can be learned from the
+data. This could be achieved by learning a hierarchical clustering on
+the original series. Further, if contextual information is available,
+but it is not hierarchical, it is possible to setup a supervised
+problem that uses context to predict features of the time series. We
+can construct a tree by applying a tree-based classifier
+[@breiman1984classification] or extracting a regression tree from a
+more complex supervised model [@boz2002extracting,
+@saito2002extracting]. Analysis then focuses on how different
+partitions of the contextual, covariate space relate to observed time
+series. This approach is described in Section \@ref(bikesharing-study).
 
-Finally, note that, while we have focused on time series valued nodes, all of
-this discussion could be translated to studying high dimensional data via
-parallel coordinates (todo: cite parallel coordinates). The usual parallel
-coordinates challenges remain, mainly selecting scales for and ordering the
-different coordinates, but the linking and focus + context can still be
+Finally, note that, while we have focused on time series valued nodes,
+all of this discussion could be translated to studying
+high-dimensional data via parallel coordinates
+[@inselberg1991parallel]. The usual parallel coordinates challenges
+remain, mainly selecting scales for and ordering across the different
+coordinates, but the linking and focus-plus-context can still be
 employed this setting.
@@
 no surprise that the techniques we adapt come from the literature on
-high-dimensional data visualization, namely, work on the focus + context and
+high-dimensional data visualization, namely, work on the focus-plus-context and
 linking.
 
-The focus + context principle is that it is possible to study large collections
+The focus-plus-context principle is that it is possible to study large collections
 of visual elements at multiple scales, and these scales can be juxtaposed by
@@
 
-### Dynamics in Housing Prices ###
+### Dynamics in Housing Prices {#zillow-study}
 
@@
 
-### Sources of Variation in Bikesharing Demand ###
+### Sources of Variation in Bikesharing Demand {#bikesharing-study}
 
@@
 we have given a series of examples to demonstrate how the general
-visualization techniques of focus + context and linked brushing can be
+visualization techniques of focus-plus-context and linked brushing can be
 practically incorporated into a range of practical analysis workflows,

### 2016-12-05 treelapse.Rmd

@@
 
-Now that we have specified the essential questions of interest, we survey some
-ideas from the visualization literature that can be applied to answer
-them. Again, the core difficulty is one of high dimensionality, so it should be
-no surprise that the techniques we adapt come from the literature on
-high-dimensional data visualization, namely, work on the focus-plus-context and
-linking.
+Now that we have specified the essential questions of interest, we
+survey some ideas from the visualization literature that can be
+applied to answer them. As the core difficulty is high-dimensionality,
+so it should be no surprise that the techniques we adapt come from the
+literature on high-dimensional data visualization, namely,
+focus-plus-context and linking.
 
-The focus-plus-context principle is that it is possible to study large collections
-of visual elements at multiple scales, and these scales can be juxtaposed by
-"focusing" on a few elements of interest while maintaining the "context" of the
-coarser-scale background. A simple example of this idea is to include a search
-box that highlights samples corresponding to matching terms (the focus) while
-muting those that don't (the context). Two more sophisticated methods anchored
-in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are central
-to the proposals in treelapse \@cite(hochheiser2004dynamic,
-heer2004doitrees). In timeboxes, a collection of time series are graphically
-queried using interactive brushes. Series that pass through all of the
-user-specified brushes are highlighted, and the rest are faded into the
-background. Hence, time series meeting the constraints imposed by the brushes
-are focused, while the remainder remain present, though de-emphasized, as
-context. This method can be interpreted programmatically as the visual version
-of a database query, or probabilistically as the conditional distribution for
-the full series, given it passes through certain bounds.
+The focus-plus-context principle is that large collections of visual
+elements can be studied at multiple scales, by simultaneously
+focusing" on a few elements of interest and maintaining the "context"
+of the coarser-scale background. A simple example of this idea is to
+include a search box that highlights matching samples (focus) and
+mutes the rest (context). Two more sophisticated methods anchored in
+this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
+central to the proposals in treelapse [@hochheiser2004dynamic,
+heer2004doitrees]. In timeboxes, a collection of time series are
+graphically queried using interactive brushes. Series that pass
+through all of the user-specified brushes are highlighted, and the
+rest are faded to the background. Hence, time series meeting the
+constraints imposed by the brushes are focused, while the remainder
+are de-emphasized, though they remain present as context. This method
+can be interpreted programmatically as the visual analog of a database
+query, or probabilistically as the conditional distribution for the
+full series, given it passes through certain bounds.
 
-In DOI Trees, the viewer's attention is focused on a collection of
-high-interest nodes, while the remaining lower-interest nodes are left on the
-fringes (or entirely hidden) as context. The implementation is modularized into
-two tasks -- the determination of a DOI distribution over nodes in the tree and
-visual layout of a tree given DOI assignments. The DOI distribution used in
-\@cite(heer2004doitrees) places maximal interest on nodes that the user clicks,
-along with all ancestors. The DOI for all other nodes is defined as the
-distance to the closest maximal interest node. The layout proposed here trimmed
-low-interest subtrees until the remaining nodes all fit within the screen
-size. By changing the minimal DOI below which nodes are hidden, the user can
-transition between node-specific and full-tree scales.
+In DOI trees, the viewer's attention is focused on a collection of
+high-interest nodes, while the remaining lower-interest nodes are left
+on the fringes as context. The implementation is modularized into two
+tasks -- the determination of a DOI distribution over nodes in the
+tree and visual layout of a tree given DOI assignments. The DOI
+distribution used in [@heer2004doitrees] places maximal interest on
+the node that the user had most recently clicked, along with all
+ancestors. The DOI for all other nodes is defined as the distance to
+the closest maximal interest node. The layout step then trims
+low-interest subtrees until the remaining nodes fit within a given
+screen size. By adjusting the minimal DOI below which nodes are
+hidden, the user can transition between node-specific and full-tree
+scales.
 
-In linking, alternative representations of the same samples are placed side by
-side in a way that suggests covariation across the representations. A canonical
-application is to linked scatterplot brushing \@cite(becker1987brushing). Here,
-a scatterplot matrix is constructed displaying the relationship between all
-pairs of variables.  Points brushed in one scatterplot are then highlighted in
-all others. For example, this helps the user determine whether an outlier in
-one dimension is an outlier in others. Another instance of this idea is in
-linking the results of dimensionality reduction methods to displays of the raw
-data, as implemented by XGobi and Cranvas, for example. As in timeboxes,
-linking ca ben interpreted as database queries or conditional probabilities:
-given a subset of the series after conditioning on the values for one set of
-features, what are the values for a second set \@cite(buja1996interactive)?
+In linking, alternative representations of the same samples are placed
+side-by-side in order to display covariation across views. A canonical
+application is to linked scatterplot brushing
+[@becker1987brushing]. Here, a scatterplot matrix gives the
+relationship between all pairs of variables.  Points brushed in one
+scatterplot are then highlighted in all others. For example, this
+helps the user determine whether an outlier in one dimension is an
+outlier in others. Another instance of this idea links the results of
+dimensionality reduction methods to displays of the raw data, as
+implemented by XGobi and Cranvas, for example [@xie2013cranvas,
+@swayne1998xgobi]. As in timeboxes, linking can be interpreted as
+database queries or conditional probabilities: given a subset of the
+series after conditioning on the values for one set of features, what
+are the values for a second set [@buja1996interactive]?
 
-Finally, while not exactly a fundamental visualization principle, we note that
-our work is deliberately grounded in the R software ecosystem. Not only does
-this make these methods more broadly accessible, we hope to encourage a more
-animated exchange between data preparation and modeling with interactive
-visualization. Moreover, our tools are intentionally limited in scope --
-designed to facilitate this dialog for a specific class of problems, rather
-than providing a toolbox for generic types of visualization design. We believe
-that this narrow context within a broad ecosystem strikes a balance between
-problem-specificity and ease-of-use.
+Finally, unrelated to established visualization principles, we note
+that our work is deliberately grounded in the R software
+ecosystem. Not only does this make these methods more broadly
+accessible, we hope to facilitate exchange between data modeling and
+interactive visualization. Moreover, our tools are intentionally
+limited in scope -- designed to facilitate this dialog for a specific
+class of problems, rather than providing a toolbox for generic types
+of visualization design. We believe that this narrow context within a
+broad ecosystem strikes a balance between problem-specificity and
+ease-of-use.
 
@@
 
-Our first proposed visualization technique is a minor modification of the DOI
-Tree idea. The standard DOI Tree definition does not have any notion of data
-defined at nodes, it is only used a device for navigating tree structures. A
-trivial extension can encode scalar data at nodes: make the node radius reflect
-the associated scalar value. To reinforce this effect, we also encode this
-scalar value in the width of the parent edge. When parent nodes have values
-equal to the sum of their children, this creates the effect of values "flowing"
-from the root to leaves. To help viewers make use of their domain knowledge, we
-have included a search box that highlights paths to matching node terms. Edges
-are ordered from widest on the left to narrowest on the right. While this
-method can only be used to visualize a single scalar-value per tree node, it
-suggests an approach to the tree-structured differential abundance problem,
-which we call the DOI Sankey.
+Our first proposed visualization technique is a minor modification of
+the DOI tree. The standard DOI tree definition does not have any
+notion of data defined at nodes, it is only used a device for
+navigating tree structures. A trivial extension can encode scalar data
+at nodes: have the node radius reflect the associated scalar value. To
+reinforce this effect, we can adjust the width of the parent
+edge. When parent nodes have values equal to the sum of their
+children, this creates the effect of values "flowing" from the root to
+leaves. To help viewers make use of their domain knowledge, we have
+included a search box that highlights paths to nodes with matching
+terms. Edges are ordered from widest on the left to narrowest on the
+right. While this method can only represent a single scalar-value per
+tree node, it suggests an approach to the tree-structured differential
+abundance problem, which we call the DOI sankey.
 
-In the DOI Sankey, we split each edge in the DOI Tree across different
-groups. For example, suppose we have the average counts for treatment and
-control groups at each tree node. Every edge in the tree is split into two
-colors^[We use the colorbrewer palette to facilitate readability], with the
-relative widths of the different colors reflecting the difference in sizes for
-the two groups. The overall width of each edge is represents the sum of values
-across all groups.
+In the DOI sankey, we split each edge in the DOI Tree across different
+groups. For example, suppose we have the average counts for treatment
+and control groups at each tree node. Every edge in the tree is split
+into two colors^[We use the colorbrewer palette to facilitate
+readability], with relative widths of the different colors reflecting
+differences in sizes for the two groups. The overall width of each
+edge represents the sum of values across all groups.
 
-This display is designed to facilitate investigation of the tree structured
-differential abundance question. For example, for a single node and a single
-group, we can compute the average abundance at that node among all samples in
-that group. This will give the width for that group's color on the edge leading
-to the specified node. Differentially abundant subtrees then correspond to
-subtrees where some colors occupy more space than others. That is, this
-representation makes it easier to identify points where the "flows" for
-different groups diverge -- the colors begin to separate. The DOI principle
-assists navigation across the tree structure, allowing focus on individual flow
-structures without losing broader tree context.
+This display is designed to facilitate investigation of the tree
+structured differential abundance question. For example, for a single
+node and a single group, first compute the average abundance at that
+node among all samples in that group. This will give the width for
+that group's color on the edge leading to the specified
+node. Differentially abundant subtrees then correspond to subtrees
+where some colors occupy more space than others. That is, this
+representation makes it easier to identify points where the "flows"
+for different groups diverge -- the colors begin to separate. The DOI
+principle assists navigation across the tree structure, allowing focus
+on individual flow structures without losing broader tree context.
 
-Our third display is directed at the microbial dynamics question instead. It
-ignores the DOI principle, placing emphasis rather on linking and the
-representation of full time series. Here, two panels are arranged one over the
-other; one displays all time series, while the other displays all tree nodes,
-with node sizes reflecting the average time series value at that node. For this
-reason, we call the display, Timebox Trees. In the time series panel, we have
-directly implemented the timeboxes idea. Our link is between the two panels:
-when a set of series is highlighted by the timeboxes, the associated tree nodes
-are also highlighted. For example, timeboxes can be used to focus on a set of
-series with specific shape -- increased abundance after an ecological shock,
-for example -- and then identify along what subtrees this pattern is
-present. To further focus attention, a pan-zoom scented widget is provided in
-the top right [@willett2007scented]. The widget is a miniature version of
-the full time series panel, equipped with a single brush whose extent specifies
-the limits in the main time series panel. As in the DOI trees and Sankeys, a
-search bar can be used to highlight of interest a priori.
+Our third display is directed at the microbial dynamics
+question. Here, two panels are arranged one over the other; one
+displays all time series, while the other displays all tree nodes,
+with node sizes reflecting the time series value at that node. For
+this reason, we call the display, Timebox Trees. In the time series
+panel, we have directly implemented the timeboxes idea. We then link
+the panels: when a set of series is highlighted by the timeboxes, the
+associated tree nodes are also highlighted. For example, timeboxes can
+be used to focus on a set of series with specific shape -- increased
+abundance after an ecological shock, for example -- and identify along
+what subtrees this pattern is present. To further focus on specific
+elements, a pan-zoom scented widget is provided
+[@willett2007scented]. The widget is a miniature version of the full
+time series panel, equipped with a single brush whose extent specifies
+the limits in the main time series panel. As in the DOI trees and
+sankeys, a search bar can be used to highlight those series of
+interest a priori.
 
-The final display currently implemented in the package is the natural converse
-of the Timebox Trees display. Rather than defining visual queries in terms of
-time series, we define queries using nodes in the tree. For this reason, we
-call the display Treeboxes. Rather than focusing on the intersection of
-brushes, as in Timebox Trees, we focus on the union of brushed over nodes. This
-allows us to highlight the series associated with nodes on distant
-subtrees. This display is also suited for the microbial dynamics problem. For
-example, by highlighting all nodes at one taxonomic level in the tree, we can
-easily summarize the time series pattern for all the taxa at that
-level. Alternatively, focusing on all the children below a single node makes it
-possible to see how much correlation and competition there is between
-taxonomically similar microbes. As in the Timebox Trees display, a search box
-and pan-zoom scented widget is provided.
+The final display currently implemented in the package is the natural
+converse of the Timebox Trees display. Rather than defining visual
+queries in terms of time series, it defines queries using nodes in the
+tree. For this reason, we call the display Treeboxes. Rather than
+focusing on the intersection of brushes, as in Timebox Trees, we focus
+on the union of brushed over nodes. This allows us to highlight series
+associated with nodes on distant subtrees. This display is also suited
+for the microbial dynamics problem. For example, by highlighting all
+nodes at one taxonomic level in the tree, we can easily summarize the
+time series pattern for all the taxa at that level. Alternatively,
+focusing on all the children below a single node makes it possible to
+see how much correlation and competition there is between
+taxonomically similar microbes. As in the Timebox Trees display, a
+search box and pan-zoom scented widget are provided.
 
@@
 a heatmap, with microbes ordered according to a hierarchical
-clustering.  By using the DOI Sankey instead, we can interpret the
+clustering.  By using the DOI sankey instead, we can interpret the
 CSTs in their taxonomic context and at multiple scales of taxonomic
@@
 Finally, we can study the behavior of the regression tree itself using
-the DOI Sankey. Here, we group samples according to their quintile of
+the DOI sankey. Here, we group samples according to their quintile of
 8AM demand and then count the abundance of the groups flowing down

### 2016-12-06 treelapse.Rmd
Revisions to case studies and conclusion

@@
   bookdown::pdf_book:
+    keep_tex: yes
 bibliography: ["bibliography.bib"]
@@
 into two colors^[We use the colorbrewer palette to facilitate
-readability], with relative widths of the different colors reflecting
-differences in sizes for the two groups. The overall width of each
-edge represents the sum of values across all groups.
+readability [@brewer2003colorbrewer]], with relative widths of the
+different colors reflecting differences in sizes for the two
+groups. The overall width of each edge represents the sum of values
+across all groups.
 
@@
 The final display currently implemented in the package is the natural
-converse of the Timebox Trees display. Rather than defining visual
+converse of the timebox trees display. Rather than defining visual
 queries in terms of time series, it defines queries using nodes in the
 tree. For this reason, we call the display Treeboxes. Rather than
-focusing on the intersection of brushes, as in Timebox Trees, we focus
+focusing on the intersection of brushes, as in timebox trees, we focus
 on the union of brushed over nodes. This allows us to highlight series
@@
 see how much correlation and competition there is between
-taxonomically similar microbes. As in the Timebox Trees display, a
+taxonomically similar microbes. As in the timebox trees display, a
 search box and pan-zoom scented widget are provided.
@@
 
-We now delve into concrete examples of applying treelapse to real
-data. Our goals are is to illustrate potential workflows that
-incorporate treelapse, describe the formulation of questions that can
-be naturally investigated with our methods, and provide example
-interpretations for specific treelapse output. Our examples are also
-chosen to reflect the range of problem domains to which the package
-can be applied -- though it was motivated by applications to the
-microbiome, it is not tied to it. More importantly, we argue that the
-visualization principles described earlier can substantively improve
-the practice of data analysis in the class of problems we have focused
-attention on.
+We now delve into applications on real data. Our goals are to
+illustrate potential workflows that incorporate treelapse, describe
+the formulation of questions that can be naturally investigated with
+our methods, and provide example interpretations on treelapse
+output. Our examples are also chosen to reflect the range of problem
+domains to which the package can be applied -- though it was motivated
+by applications to the microbiome, it is not tied to it. More
+importantly, we argue that the visualization principles reviewed above
+can substantively improve the practice of data analysis in the class
+of problems to which we have limited ourselves.
 
@@
 
-[@dethlefsen2008pervasive] investigates the effect of antibiotics on microbial
-community composition from an ecological perspective. The study tracks the
-microbiome of *x* patients across *y* weeks (todo: give the numbers), with
-antibiotic two time courses introduced during * and *, respectively. Discerning
-the variation in resilience across microbes is important, considering the
-community's increasing acceptance of the idea that microbes are central to human
-health and not just disease.
+[@dethlefsen2008pervasive] investigates the effect of antibiotics on
+microbial community composition from an ecological perspective. The
+study tracks the microbiome of three patients across ten months, with
+two five-day antibiotic time courses separated by 6 months. Discerning
+the variation in resilience across microbes is important, considering
+the the role of microbes in health and not just disease.
 
-We approach the data using the linked time and treebox views, after first
-filtering low variance taxa and taking an $asinh$ transformation. An initial
-view, Figure \@ref(fig:antibioticoverview), reveals two dramatic drops in the
-overall microbial abundance time series at the time of the antibiotics time
-courses. Two more subtle effects are also suggested from this view,
+We approach the data using the linked time and treebox views, after
+first filtering low variance taxa and taking an $\text{asinh}$
+transformation. An initial view, Figure \@ref(fig:antibioticoverview),
+reveals two dramatic drops in the overall microbial abundance time
+series during the antibiotics time courses. Two more subtle effects
+are also suggested from this view,
 
@@
 then behavior of different families of Firmicutes, Lachnospiraceae and
-Ruminoccocus. We suspect these taxa are associated with the delayed
-recovery after the second time course. To inspect this, we these
-family names in the search box to isolate their positions on the tree,
-and then we apply brushes to highlight the series that contribute to
-these higher-level families. The resulting view is given by Figure
-\@ref(fig:antibioticfirmicutes)
+Ruminoccocus. We suspect that these taxa are associated with the
+delayed recovery after the second time course. To investigate this, we
+input these family names in the search box to isolate their positions
+on the tree; then we apply brushes to highlight the series that
+contribute to these higher-level families. The resulting view is given
+by Figure \@ref(fig:antibioticfirmicutes)
 
@@
 
-Alternatively, we can summarize each node by the average of its descendants --
-this brings attention to individual microbes that may be underlying some of the
-broader taxonomic patterns we have noted when studying the subtree sums. For
-example, in Figure \@ref(fig:ruminococcus), we highlight all families below the
-Ruminococcus, suggesting that the decrease due to antibiotics occurs uniformly
-across almost all families. A point that was not evident in the earlier
-sum-across-descendants view is that, after the second treatment of antibiotics,
-a few of the Ruminoccocus families recover more rapidly than the rest, for
-example the Unc095d3 (highlighted in red) are only briefly affected. In
-contrast, most families seem to recover in unison after the first treatment.
+Alternatively, we can summarize each node by the average across its
+descendants -- this brings attention to individual microbes that may
+be underlying some of the broader taxonomic patterns we have noted
+when studying the subtree sums. For example, in Figure
+\@ref(fig:ruminococcus), we highlight all families below order
+Ruminococcus, suggesting that the decrease due to antibiotics occurs
+uniformly across almost all families. A point that was not evident in
+the earlier sum-across-descendants view is that, after the second
+treatment of antibiotics, a few of the Ruminoccocus families recover
+more rapidly than the rest, for example the Unc095d3 (highlighted in
+red) are only briefly affected. In contrast, most families seem to
+recover in unison after the first treatment.
 
@@
 Verrucomicrobiae at the far left, which seem to have large abundance
-over many time points. This phylum had been previously obfuscated,
-because there are not many leaves associated with this phylum, hence
-the sum was small. Interestingly, the abundance of these microbes
-seems to *increase* after the first antibiotics treatment.  Be
-cautious, however, that the average over only a few Verrucomicrobiae
-species will be a more variable estimate than the averages over the
-more prevalent phyla.
+over many time points. This phylum had been previously obfuscated --
+because there are not many leaves associated with this phylum, the sum
+was small. Interestingly, the abundance of these microbes seems to
+*increase* after the first antibiotics treatment.  Be cautious,
+however, that the average over only a few Verrucomicrobiae species
+will be a more variable estimate than the averages over the more
+prevalent phyla.
 
@@
 
-The study [@digiulio2015temporal] tracked the abundance of microbes in the
-during vaginal microbiome during pregnancy in an effort to study relationships
-between microbial community composition and preterm birth. Ideally, it would be
-possible to develop clear microbial signatures associated with preterm births.
+The study [@digiulio2015temporal] tracked the abundance of microbes in
+the vaginal microbiome during pregnancy in an effort to study
+relationships between microbial community composition and preterm
+birth. Ideally, it would be possible to develop clear microbial
+signatures associated with preterm births.
 
-Unlike the antibiotics study, we have measurements across more individuals than
-we could reasonably inspect one at a time. While we could average across all
-individuals, we will take our cue from [@digiulio2015temporal] and place each
-sample into one of five Community State Types (CSTs), identified via k-medoids.
-Using a linear model, one of these CSTs (CST 4) was found to be more diverse,
-further it appeared associated with preterm births. Here, we corroborate this
-finding using exploratory views.
+Unlike the antibiotics study, we have measurements across more
+individuals than we could reasonably inspect one at a time. While we
+could average across all individuals, we will take our cue from
+[@digiulio2015temporal] and place each sample into one of five
+Community State Types (CSTs), identified via k-medoids.  In that
+study, a linear model identified one of these CSTs (CST 4) as
+significantly more diverse, further it appeared associated with
+preterm births. Here, we corroborate this finding using exploratory
+views.
 
@@
 of differential abundance across CSTs and also between preterm and
-non-preterm births. [@digiulio2015temporal] interpreted its CSTs using
+non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
 a heatmap, with microbes ordered according to a hierarchical
@@
 all samples within each CST to define values at the (microbe-level)
-leaf nodes, and then aggregated the counts up to the root. It is
+leaf nodes, and then aggregated the averages up to the root. It is
 immediately clear that samples from CST 4 have much more taxonomic
 diversity. Further, focusing on the Lactobacillaceae family, we note
-that the differential abundnace of these microbes distinguishes th
+that the differential abundance of these microbes distinguishes the
 remaining CSTs, see Figure \@ref(fig:pretermcstslacto).
 
-Alternatively, in Figure \@ref(fig:preterm), we avoid working with
+Alternatively, in Figure \@ref(fig:pretermcsts), we avoid working with
 CSTs, displaying instead averages among samples associated with either
 preterm or term births. The green edges are associated with preterm
-births -- we see that they contribute more weight to Phyla outside the
+births -- we see that they contribute more weight to phyla outside the
 Firmicutes. This is consistent with the claim that CST 4, the most
@@
 relatively clear hierarchical structure. Our data are downloaded from
-Zillow^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
+Zillow[^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)]],
 and give the Zillow Home Value Indexes at the neighborhood level,
-across the country, given monthly between 1996 and 2016. In our
+across the country, computed monthly between 1996 and 2016. In our
 display, we have taken the base 10 log of these indexes. As our
@@
 California; rendering more neighborhoods while keeping all 246
-timepoints causes the interface to lag^[See the Future Work section
-for potential optimizations, however.].
+timepoints causes the interface to lag^[See the Section
+\@ref(conclusion) for potential optimizations, however.].
 
@@
 in home prices. We are especially interested in the effect of the 2008
-recession and any variation in the lead-up or recovery. These
-questions can be naturally framed using timebox trees and treeboxes.
+recession and any variation in the lead-up to or recovery from this
+event. These questions can be naturally framed using timebox trees and
+treeboxes.
 
@@
 be highlighted using a single brush in treeboxes, since each spatial
-level is displayed at the same height in the tree. THe resulting
+level is displayed at the same height in the tree. The resulting
 display is given in Figure \@ref(fig:zillowcities). From a high level,
@@
 
-Alternatively, we can consider the trajectories of home prices among
-neighborhoods, conditional on their being that were middle-income
-before the recession.  We generate the sequence of views in Figures
+Alternatively, we can study the trajectories of home prices among
+neighborhoods, conditional on their being middle-income before the
+recession. We generate the sequence of views in Figures
 \@ref(fig:zillowmiddlepre), \@ref(fig:zillowmiddleup), and
-\@ref(fig:zillowmiddledown) in this vein. The first of these figures
+\@ref(fig:zillowmiddledown) to this end. The first of these figures
 isolates neighborhoods with middle incomes before the recession, using
@@
 series in San Francisco County. We see that, in 2016, prices range from around
-$10^{5.6} \approx \$400,000$ to $10^{6.7} \approx \$5 \text{million}$. So, while
+$10^{5.6} \approx \$400,000$ to $10^{6.7} \approx \$5 \text{ million}$. So, while
 all these neighborhoods tend to be among the more expensive ones in California,
@@
 enough listings on the site, so should be supplemented by other data
-sources for more substantial decision-making.
+sources for any substantial decision-making.
 
@@
 example of analyzing collections of time series when there is no
-obvious hierarchical structure a priori. The data are downloaded from the
-[UCI Repository](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset) and
-were originally collected by a Washington D.C.-based bikesharing
+obvious hierarchical structure a priori. The data are available at
+the [UCI
+Repository][^[(https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
+and were originally collected by a Washington D.C.-based bikesharing
 system for use in a Kaggle prediction competition. The data are hourly
 measurements of bike demand, aggregated across all bikesharing
-stations, over two years, along with some supplemental weather
-data. In the competition, participants were asked to predict the
-hourly demand on a held-out test set. Here, we adopt a descriptive
-view instead, attempting to characterize factors associated with
-variation in bikesharing demand.
+stations, over two years, along with supplemental weather data. In the
+competition, participants were asked to predict the hourly demand on a
+held-out test set. Here, we adopt a descriptive view instead,
+attempting to characterize factors associated with variation in
+bikesharing demand.
 
@@
 a natural choice considering the daily periodicity of bike demand. To
-relate these daily series according to an interpretable tree
-structure, we apply a regression tree relating the supplemental data
-to the bikesharing demand (J.  Friedman, Hastie, and Tibshirani
-2001). In more detail, we build this tree by noting the “two table”
-structure of this problem: one describes bike demand, the other holds
-the supplemental data. In both, the rows index days, while the columns
-index hours and supplemental features for the two tables,
-respectively. Our tree is the trained regression tree after predicting
-demand at 8AM based on the supplemental data. We choose this response
-because (1) we need a univariate response in order to apply regression
-trees and (2) the more straightforwards reduction to
+arrange these daily series along an interpretable tree structure, we
+apply a regression tree relating the supplemental data to the
+bikesharing demand [@breiman1984classification]. In more detail, we
+built this tree by noting the “two table” structure of this problem:
+one describes bike demand, the other holds the supplemental data. In
+both, the rows index days, while the columns index either hours or
+supplemental features. Our tree is the trained regression tree after
+predicting demand at 8AM based on the supplemental data. We choose
+this response because (1) we need a univariate response in order to
+apply regression trees and (2) the more straightforwards reduction to
 daily-average-demand fails to distinguish between weekdays and
@@
 isolate the two types of series highlight the left and right sides of
-the tree, respectively. For a mroe subtle effect, we select the
+the tree, respectively. For a more subtle effect, we select the
 internal nodes associated with the first split below the weekday
 vs. weekend split; these are given in Figures 16 and 17. This suggests
-that more commuters became aware or interested in bikesharing during
-the second year of data collection.
+that weekday demand increased during the second year.
 
-To contrast with these general questions about daily demand, we ask a
-more granular question about demand at night: what characterizes days
-on which there is larger than average demand after midnight? We can
-select these series after first zooming into this time window. Figure
-18 reveals that the highlighted series are associated with the
-warm-weekend split, which seems quite reasonable in retrospect.
+In contrast to these general questions about daily demand, we could
+ask a more granular question about specific time windows. For example,
+what characterizes days on which there is larger than average demand
+after midnight? We can select these series after first zooming into
+this time window. Figure 18 reveals that the highlighted series are
+associated with the warm-weekend split, which seems quite reasonable
+in retrospect.
 
@@
 strongly separated after descending even a few steps down the
-regression tree – for example, in Figure ?? focuses on 2011 vs. 2012
-split among weekday samples, showing that this split distinguishes
-between samples falling in the second and third quintiles of 8AM
-demand.
+regression tree – for example, Figures \@ref(fig:weekday_2011) and
+\@ref(fig:weekday_2012) focuse on 2011 vs. 2012 split among weekday
+samples, showing that this split distinguishes between samples falling
+in the second and third quintiles of 8AM demand.
 
@@
 two-table data could be used in conjunction with more sophisticated
-regression methods. In our ideas data analysis workflow, we imagine
+regression methods. In our ideal data analysis workflow, we imagine
 the analyst applying interactive visualization and modeling techniques
@@
 
-## Conclusion and Future Work ##
+```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
+knitr::include_graphics("working.png")
+```
 
-Here, we have reviewed some fundamental principles of data visualization and described their implementation
-in a new treelapse package. Further, we have provided examples of the practical usefulness of these
-principles in real-world data analysis situations.
+```{r weekend, fig.cap = "Unlike weekday demand, weekend demand is unimodal.
+The few weekday series with unimodal series seem to be associated with holidays.
+This is the case for New Years' Eve, which is currently hovered over in the tree."}
+knitr::include_graphics("weekend.png")
+```
 
-This package has only developed some of the most basic ideas, and
-there are a number of potentially useful extensions worth
-exploring. For example, we have not considered the principle of
-arrangement in our visualizations (Buja, Cook, and Swayne 1996),
-though many of our conclusions were based on comparing alternative
-selections of the same display. We could imagine faceting our displays
-across groups to make these types of comparisons more
-accessible. Further, we have only worked with the DOI distribution
-described in (Heer and Card 2004). It would be interesting to define a
-more statistical notion of interest along nodes, based on cognostics,
-for example (Hafen et al. 2013, J. H. Friedman and Stuetzle
-(2002)). Further, it could be helpful to link the visualization-based
-workflow described in this work with more traditional multivariate and
-inferential statistics workflows – it is easy to imagine an
-interactive version of ordinations that display representations of the
-raw data upon selection. A simple extension could be to allow graph
-layouts instead of trees in time and treebox displays, for data that
-cannot be coerced into a hierarchical structure. Finally, if these
-ideas turn out to be useful in practice, it would be valuable to
+```{r weekday_2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
+knitr::include_graphics("weekday_2011.png")
+```
+
+```{r weekday_2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
+knitr::include_graphics("weekday_2012.png")
+```
+
+```{r warm_weekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
+knitr::include_graphics("warm_weekend.png")
+```
+
+```{r bike_sankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
+knitr::include_graphics("bike_sankey.png")
+```
+
+## Conclusion and Future Work {#conclusion}
+
+Here, we have reviewed some fundamental principles of data
+visualization and described their implementation in a new treelapse
+package. Further, we have provided examples of the practical
+usefulness of these principles in real-world data analysis situations.
+
+This package has only developed basic ideas, and there are a number of
+potentially useful extensions worth exploring. For example, we have
+not considered the principle of arrangement in our visualizations
+(Buja, Cook, and Swayne 1996), though many of our conclusions were
+based on comparing alternative selections of the same display. We
+could imagine faceting our displays across groups to make these types
+of comparisons more accessible. Further, we have only worked with the
+DOI distribution described in (Heer and Card 2004). It would be
+interesting to define a more statistical notion of interest along
+nodes, based on cognostics, for example (Hafen et al. 2013,
+J. H. Friedman and Stuetzle (2002)). A simple extension could be to
+allow graph layouts instead of trees in time and treebox displays, for
+data that cannot be coerced into a hierarchical structure. Finally, if
+these ideas turn out to be useful in practice, it would be valuable to
 modularize the basic visualization layouts and relationships into a
-library, allowing the wider community to construct linked, interactive
-graphics with minimal effort.
+library, allowing the wider community to construct novel linked,
+interactive graphics with minimal effort.
 
-All these potential extensions aside, we have built an easily
-accessible R package for visualization techniques in a very specific
-methodology problem – analysis of differential abundance and dynamics
-in hierarchical structured data – that appears in a variety of
-application domains. We have leveraged (Vaidyanathan et al.  2014) to
-create exploratory visualizations while still deep in data analysis;
-in this way our work is a departure from the culture of polished,
-journalistic visualizations prioritized by the D3 community. Finally,
-we have given a series of examples to demonstrate how the general
-visualization techniques of focus-plus-context and linked brushing can be
+In summary, we have built an easily accessible R package for
+visualization techniques in a very specific methodology problem –
+analysis of differential abundance and dynamics in hierarchically
+structured data – that appears in a variety of application domains. We
+have leveraged (Vaidyanathan et al.  2014) to create exploratory
+visualizations while still deep in data analysis; in this way our work
+is a departure from the culture of polished, journalistic
+visualizations prioritized by the D3 community. Finally, we have given
+a series of examples to demonstrate how the general visualization
+techniques of focus-plus-context and linked brushing can be
 practically incorporated into a range of practical analysis workflows,

### 2016-12-06 treelapse.Rmd

@@
 title: "Interactive Visualization of Hierarchically Structured Data"
-author: "Kris Sankaran"
+author: "Kris Sankaran and Susan Holmes"
 date: "`r Sys.Date()`"
@@
     bookdown::pdf_book:
-    keep_tex: yes
+        toc: false
 bibliography: ["bibliography.bib"]
+biblio-style: "agsm"
 link-citations: yes
@@
 
-```{r set-chunk-opts}
-knitr::opts_chunk$set(fig.align = "center", out.width = "200px", echo = FALSE)
+## Introduction ##
+
+```{r set_chunk_opts, echo = FALSE}
+knitr::opts_chunk$set(
+  fig.align = "center",
+  out.width = "250px",
+  echo = FALSE,
+  cache = TRUE
+)
 ```
-
-## outline ##
-
-This is an outline for the treelapse article I want to write.
-
-Introduction
-
-- Give core theses
-    + Using principles from the data visualization literature, we have designed
-      a tool that we hope will be useful for analyzing tree structured data,
-      especially hierarchically structured collections of time series.
-    + Focus / context and linking are central to each of our visualizations
-    + We hope our implementation makes this specific tool more widely
-      accessible
-- Describe paper structure
-
-Problem Motivation
-
-- Microbiome motivation
-      + The meeting of ideas from medicine and ecology
-      + The differential abundance problem: treatment vs. control in
-        experiments
-      + The ecological dynamics problem
-             * Are there niches?
-             * Is there competition?
-- Abstraction
-      + Random variables at nodes
-	  + In differential abundance, we have a vector of counts, one coordinate
-        per group
-	  + For microbial dynamics, we have a time series at nodes
-	  + In these applications, we define values at nodes as linear combinations
-        of descendants (sum or average). This sort of constraint is not
-		necessary though
-- Problems with related structure
-      + Spatial hierarchy when working with time series
-      + Creating trees
-             * Hierarchical clustering
-             * CART
-             * When variables don't have intrinsic order, but have comparable
-               ranges, can interpret time series as parallel coordinates
-
-Solution Principles / Background Literature
-
-- Focus and Context (summary then examples)
-      + Degree of interest -- a way of navigating graph structures. Modularize
-        the problem of graph layout and interaction. One part is determination
-        of a degree of interest for nodes. The other is determining a layout
-        given a particular degree of interest distribution. Blocks are trimmed
-        away depending on their average degree of interest.
-      + Timeboxes -- A way of querying the types of shapes that appear in a
-        given collection of time series. You draw boxes, and it highlights
-        series that pass through them. The focus is on highlighted series, the
-        context is the collection of series in the background.
-      + Visual scent -- Determine pan and zoom based on a smaller version of
-        the larger figure. Similar to interactively adjusting bounds
-        numerically, but more intuitive. Focus is on the part of the scent that
-        is highlighted, context is the rest of the scent.
-- Linking (summary then examples)
-      + Linked scatterplot brushes: Let you see many dimensions for the same
-        few points. Conditional on having high values for one series, what are
-        values for other variables? This is analogous to timeboxes on parallel
-        coordinates plots.
-      + Visual queries / conditional interpretation: Power is in querying one
-        variable and seeing values in the other.
-      + XGobi: Historical example of linking views, grounded in
-        multidimensional scaling
-
-- Participation in broader R software ecosystem
-    + Can iterate through many interactive visualizations (like in ggvis)
-    + Limited scope, but easier to maintain and more likely to remain relevant
-    + Rely heavily on htmlwidgets
-         * Also note similarity with networkD3: though, we have focused more on
-           principles for a specific problem setup, rather than constructing
-           foundational building blocks (in a way, this is more applied)
-
-Specific Proposals
-
-- DOI Trees
-     + definition and (microbiome) use case: for a single sample, where are
-       most bacteria?
-     + principles it relies on: just DOI tree.
-     + This is the only visualization that is completely thought through from
-       before (but which as far as we can tell hasn't had easily accessible
-       implementations).
-     + well not exactly...we are also encoding abundance in edge widths / node
-       sizes
-- DOI Sankeys
-     + Definition and use case: treatment vs. control in tree structured data,
-       but using DOI principle
-     + Exploratory analog of structSSI
-     + Minor note: we use the colorbrewer palette for simpler readability
-- Timebox Trees
-     + Combines a few more ideas: linking and focus-plus-context
-     + definition and explanation of how these ideas being implemented
-              * timebox recap: we're able to highlight lines passing through
-                intersection of boxes (this is focus and context)
-              * Link the display between tree and time series. Interpretation:
-                Position on tree conditional on time series
-                shape. Alternatively, database query on one field, returning
-                over all of them.
-              * Focus and context / pan and zoom using visual scent [so we have
-                two focus and context ideas, and one linking idea, in this one
-                display]
-      + Primary use case this is designed for -- we notice a pattern in how a
-        clump of microbial series look like, and we try to figure out which
-        subtrees have this behavior (or, if that behavior is appear at
-        different places on the tree).
-- Treeboxes
-       + Describe as the natural analog of timebox trees
-       + Intentionally avoided queries on both tree and time series
-       + Example use case: we can see how different parts of the tree have
-         different series.
-               * Or, we can look at all the phyla / orders at once
-               * Or, we can observe if there is any competition within a
-                 particular clade
-       + Conditional on coming from a part of a tree, what are the observed
-         series?
-
-Case Studies
-
-- Introduction
-      + Practically, want to provide example output and how to interpret it
-      + This is only a partial goal (there are many vignettes that describe
-        this in more detail and are explicitly designed as tutorials) -- more
-        of the focus here is on explaining how the visual principles described
-        above can substantively improve data analysis in this problem domain
-      + Purpose is to motivate potential workflows that incorporate the
-        treelapse package. Show how it can be incorporated into any existing R
-        data analysis workflow (this was all done in a single R document)
-      + Argue that the problem of navigating hierarchical collections of time
-        series is actually more generic than you might think
-
-- Antibiotics
-      + Study background: (check out abstract to confirm) How does microbiota
-        respond to a time course of antibiotics. Especially relevant given
-        growing understanding of importance of microbes in *health* (not just
-        disease)
-      + Framing as a timebox / treebox problem
-      + Initial view: can clearly discern decrease across all microbes
-      + Selecting some series, notice that the Firmicutes are more strongly
-        affected
-      + Can zoom into period around perturbation, using the scent idea
-      + Can rerun using node averaged data -- this is only possible because we
-        have linked it with R (it would not be possible in a standard d3 Block,
-        for example)
-      + Ideally: comparison with gPCA? Show how it complements.
-- DOI Sankey: Preterm births
-      + Basic goal of the study -- are certain microbiome signatures associated
-        with preterm births?
-      + Formal inferential approach was taken by the original publication,
-        testing each microbe for an association with term / preterm
-      + Also had an unsupervised analysis: cluster into CSTs
-                - We can follow-up this unsupervised work using treelapse (this
-                  is analogous to gPCA followup)
-                - They used a heatmap, which gets the main points across, but
-                  without the tree context (or allowing interactive focus +
-                  context)
-      + We immediately note that CST 4 has much more diversity
-      + navigating down the tree, we notice that the main way the other CSTs
-        are different is in the different types of Lactobacillus
-- Housing prices
-      + A non-microbial example, but with relatively obvious hierarchical
-        (spatial) structure
-      + Basic goal: How were home prices affected by recession, and how do
-        prices and trends in prices differ across regions?
-      + Findings from timebox / treebox:
-                 * Can see quickly which neighborhoods are richer than others
-                 * Can also see that poorer areas were more strongly affected
-                   by the recession
-                 * Can also see the variation in home prices within individual
-                   counties
-      + Limitations: We never actually see the spatial locations for the
-        series. Can easily imagine a linking between the map and the original
-        series (cite xgobi), but this is out of scope (at least for now)
-- Bikesharing
-      + A non-microbial example, and with non-obvious hierarchical structure
-      + To induce hierarchical structure, we perform classification using a
-        supplementary data table
-                * We take advantage of the fact that this is essentially a two
-                  table problem (one whose rows are different time series,
-                  another giving the supplemental data)
-                * Perform classification using classification trees (similar to
-                  Kaggle, but not quite -- we treat a whole series as a sample,
-                  not just a single time point).
-       + Observations
-                * Quickly notice difference between weekend and weekday
-                * Also more subtle differences though -- temperature and year
-                  effects, for example
-                * The 3AM people on warm weekends (importance of focus)
-       + Ideal scenario: Develop model using ideas from here, relate model to
-         hierarchical structure again. Alternatively, fit a tree on correct and
-         incorrect cases from another algorithm, inspect what are features of
-         incorrectly classified cases, build these into new features.
-- Potentially: genetics and hierarchical clustering?
-- Summary: Tree structured data and collections of related time series are
-  common than you might suspect, the principles of data visualization actually
-  facilitate navigation of high dimensional data, and since it's in R, it's
-  easy to alternate between interactive visualization, data regeneration /
-  preparation, and statistical modeling.
-
-Conclusion
-
-- Potential future work
-   + Arrangement
-        * Cognostics for arranging views
-        * Faceting timebox / treebox views
-        * Linking with DPCOA / gPCA? In practice, microbiome literature cares a
-          lot about these ordinations, we can try to explicitly tie them
-          together in one visualization (rather than iterating, like we do
-          now).
-   + Incorporating the DOI idea into the timebox trees?
-   + Other layouts? Graphs? DAGs?
-   + Grammar
-        * Is it possible to modularize the key ideas, so we can create similar
-          kinds of linked and arranged views without so much effort?
-- Summary
-    + Developed an R package based on ideas from the visualization community
-    + Think there is power in interactive visualization for hands on data
-      analysis, not just polished public tools
-    + Think there is a relationship between statistical modeling and data
-      visualization: it can be more useful to visualize raw data in relation to
-      models
-    + Focused on solving a very specific methodological problem, rather than
-      building a generic system. But this methodological problem appears in a
-      variety of domains.
-
-## text ##
-
-## Introduction ##
 
@@
 [@hochheiser2004dynamic], we have adapted the fundamental
-visualization principles of "focus-plus-context" and "linking" to study
+visualization principles of focus-plus-context and linking to study
 questions that arise when analyzing tree structured data. Our
@@
 central to the proposals in treelapse [@hochheiser2004dynamic,
-heer2004doitrees]. In timeboxes, a collection of time series are
+@heer2004doitrees]. In timeboxes, a collection of time series are
 graphically queried using interactive brushes. Series that pass
@@
 strongly separated after descending even a few steps down the
-regression tree – for example, Figures \@ref(fig:weekday_2011) and
-\@ref(fig:weekday_2012) focuse on 2011 vs. 2012 split among weekday
+regression tree – for example, Figures \@ref(fig:weekday2011) and
+\@ref(fig:weekday2012) focuse on 2011 vs. 2012 split among weekday
 samples, showing that this split distinguishes between samples falling
@@
 more useful on larger trees that cannot be easily parsed in a single
-view; in this sense the bikesharing tree is relatively
-simple. Further, in light of research on approximating more complex
-supervised learning algorithms using decision trees (Boz 2002, Saito
-and Nakano (2002)), this approach of modeling and interpreting
-two-table data could be used in conjunction with more sophisticated
-regression methods. In our ideal data analysis workflow, we imagine
-the analyst applying interactive visualization and modeling techniques
-in an iterative, nonlinear fashion.
+view; in this sense the bikesharing tree is relatively simple. In our
+ideal data analysis workflow, we imagine the analyst applying
+interactive visualization and modeling techniques in an iterative,
+nonlinear fashion.
 
@@
 
-```{r weekday_2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
+```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
 knitr::include_graphics("weekday_2011.png")
@@
 
-```{r weekday_2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
+```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
 knitr::include_graphics("weekday_2012.png")
@@
 
-```{r warm_weekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
+```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
 knitr::include_graphics("warm_weekend.png")
@@
 
-```{r bike_sankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
+```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
 knitr::include_graphics("bike_sankey.png")
@@
 not considered the principle of arrangement in our visualizations
-(Buja, Cook, and Swayne 1996), though many of our conclusions were
-based on comparing alternative selections of the same display. We
-could imagine faceting our displays across groups to make these types
-of comparisons more accessible. Further, we have only worked with the
-DOI distribution described in (Heer and Card 2004). It would be
-interesting to define a more statistical notion of interest along
-nodes, based on cognostics, for example (Hafen et al. 2013,
-J. H. Friedman and Stuetzle (2002)). A simple extension could be to
-allow graph layouts instead of trees in time and treebox displays, for
-data that cannot be coerced into a hierarchical structure. Finally, if
-these ideas turn out to be useful in practice, it would be valuable to
-modularize the basic visualization layouts and relationships into a
-library, allowing the wider community to construct novel linked,
-interactive graphics with minimal effort.
+[@buja1996interactive], though many of our conclusions were based on
+comparing alternative selections of the same display. We could imagine
+faceting our displays across groups to make these types of comparisons
+more accessible. Further, we have only worked with the DOI
+distribution described in [@heer2004doitrees]. It would be interesting
+to define a more statistical notion of interest along nodes, based on
+cognostics, for example [@hafen2013trelliscope, @friedman2002john]. A
+simple extension could be to allow graph layouts instead of trees in
+time and treebox displays, for data that cannot be coerced into a
+hierarchical structure. Finally, if these ideas turn out to be useful
+in practice, it would be valuable to modularize the basic
+visualization layouts and relationships into a library, allowing the
+wider community to construct novel linked, interactive graphics with
+minimal effort.
 
@@
 structured data – that appears in a variety of application domains. We
-have leveraged (Vaidyanathan et al.  2014) to create exploratory
-visualizations while still deep in data analysis; in this way our work
-is a departure from the culture of polished, journalistic
-visualizations prioritized by the D3 community. Finally, we have given
-a series of examples to demonstrate how the general visualization
-techniques of focus-plus-context and linked brushing can be
-practically incorporated into a range of practical analysis workflows,
-from studying the impact of microbes on human health to better
-allocating units in commuter bikesharing systems.
+have leveraged a link between R and D3 [@vaidyanathan2014htmlwidgets]
+to create exploratory visualizations while still deep in data
+analysis; in this way our work is a departure from the culture of
+polished, journalistic visualizations prioritized by the D3
+community. Finally, we have given a series of examples to demonstrate
+how the general visualization techniques of focus-plus-context and
+linked brushing can be practically incorporated into a range of
+practical analysis workflows, from studying the impact of microbes on
+human health to better allocating units in commuter bikesharing
+systems.
 

### 2016-12-06 treelapse.Rmd
Make yaml compatible with jcgs template

@@
 ---
-title: "Interactive Visualization of Hierarchically Structured Data"
 author: "Kris Sankaran and Susan Holmes"
 date: "`r Sys.Date()`"
+standalone: true
+documentclass: article
+fontsize: 12pt
 output:
   bookdown::pdf_book:
+    keep_tex: yes
     toc: false
+    includes:
+      in_header: preamble.tex
+      before_body: before_body.tex
 bibliography: ["bibliography.bib"]

### 2016-12-06 treelapse.Rmd
Remove title elements in yaml

--- 2016-12-06 treelapse.Rmd
@@
 ---
-title: "Interactive Visualization of Hierarchically Structured Data"
-author: "Kris Sankaran and Susan Holmes"
-date: "`r Sys.Date()`"
+standalone: true
+documentclass: article
+fontsize: 12pt
 output:
   bookdown::pdf_book:
+    keep_tex: yes
     toc: false
+    includes:
+      in_header: preamble.tex
+      before_body: before_body.tex
 bibliography: ["bibliography.bib"]

### 2016-12-06 before_body.tex

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Interactive Visualization of Hierarchically Structured Data}
  \author{Kris Sankaran\thanks{
    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
    Department of Statistics, Stanford University\\
    and \\
    Susan Holmes \\
    Department of Statistics, Stanford University}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract.  200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  3 to 6 keywords, that do not appear in the title
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

### 2016-12-06 treelapse.Rmd
Add more motivation in introduction

@@
 
-(todo: sentence about why we should care, and a high level summary
-before diving into degree-of-interest linking jargon)
+We introduce methods for visualization of data structured along trees,
+especially hierarchically structured collections of time series. We
+hope both to characterize generically useful techniques for
+interactively visualizing hierarchical data and to offer practical
+tools for implementing such displays. To this end, we identify
+questions that often emerge when working with hierarchical data and
+provide an R package to simplify their investigation [@ripley2001r].
 
-In this work, we describe methods for visualization of data structured
-along trees, especially hierarchically structured collections of time
-series. Influenced by Degree-of-Interest trees [@heer2004doitrees],
-linked scatterplot brushing [@becker1987brushing], and timeboxes
-[@hochheiser2004dynamic], we have adapted the fundamental
-visualization principles of focus-plus-context and linking to study
-questions that arise when analyzing tree structured data. Our
-motivating application is to the analysis of microbial time series,
-where an evolutionary tree relating microbes is available a
-priori. However, we have identified common problem types where, if a
-tree is not directly available, it can be constructed from data and
-then studied using our techniques.
+Our key contribution is the adaptation of the visualization principles
+of focus-plus-context and linking to the study of tree-structured data
+[@buja1996interactive, @becker1987brushing]. Our motivating
+application is to the analysis of microbial time series, where an
+evolutionary tree relating microbes is available a priori. However, we
+have identified common problem types where, if a tree is not directly
+available, it can be constructed from data and then studied using our
+techniques.
 
-We have implemented our visualizations in d3, but encapsulated in an R
+We have implemented our visualizations in D3, but encapsulated in an R
 package, called treelapse, to facilitate rapid turnover from data
@@
 that our work is deliberately grounded in the R software
-ecosystem. Not only does this make these methods more broadly
-accessible, we hope to facilitate exchange between data modeling and
-interactive visualization. Moreover, our tools are intentionally
-limited in scope -- designed to facilitate this dialog for a specific
-class of problems, rather than providing a toolbox for generic types
-of visualization design. We believe that this narrow context within a
+ecosystem. This connection is made using the htmlwidgets package
+[@vaidyanathan2014htmlwidgets]. Not only does linking R with D3 make
+these visualization methods more broadly accessible, we hope to
+facilitate exchange between data modeling and interactive
+visualization. Moreover, our tools are intentionally limited in scope
+-- designed to facilitate this dialog for a specific class of
+problems, rather than providing a toolbox for generic types of
+visualization design. We believe that this narrow context within a
 broad ecosystem strikes a balance between problem-specificity and
@@
 relatively clear hierarchical structure. Our data are downloaded from
-Zillow[^[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)]],
+Zillow^[[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
 and give the Zillow Home Value Indexes at the neighborhood level,
@@
 
-```{r zillowmiddlepre}
+```{r zillowmiddlepre, fig.cap = "Neighborhoods with mid-range home prices before the recession are selected here.
+Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
 knitr::include_graphics("zillow_middle_pre.png")
@@
 
-```{r zillowmiddleup}
+```{r zillowmiddleup, fig.cap = "Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly.
+These appear to be located mainly in Los Angeles and San Diego counties."}
 knitr::include_graphics("zillow_middle_up.png")
@@
 
-```{r zillowmiddledown}
+```{r zillowmiddledown, fig.cap = "Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession.
+These seem mostly to be located in Central California and the East San Francisco Bay Area."}
 knitr::include_graphics("zillow_middle_down.png")
@@
 
-```{r zillowsf}
+```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
 knitr::include_graphics("zillow_sf.png")
@@
 obvious hierarchical structure a priori. The data are available at
-the [UCI
-Repository][^[(https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
+the UCI
+Repository^[[https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
 and were originally collected by a Washington D.C.-based bikesharing
@@
 regression tree – for example, Figures \@ref(fig:weekday2011) and
-\@ref(fig:weekday2012) focuse on 2011 vs. 2012 split among weekday
+\@ref(fig:weekday2012) focus on 2011 vs. 2012 split among weekday
 samples, showing that this split distinguishes between samples falling
@@
 
+Potential sections to write
+
+* describe extension from svg to canvas
+* give example of parallel coordinates / hierarchical clustering?
+* explain incorporation into / give example of shiny app?
+
 ## References ##

### 2016-12-06 before_body.tex
Write an abstract

@@
   \author{Kris Sankaran\thanks{
-    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
+      This research was supported in part by an NIH training grant (5T32GM096982-03).}\hspace{.2cm}\\
     Department of Statistics, Stanford University\\
@@
 \begin{abstract}
-The text of your abstract.  200 or fewer words.
+We introduce methods for visualization of data structured along trees,
+especially hierarchically structured collections of time series.  To
+this end, we identify questions that often emerge when working with
+hierarchical data and provide an R package to simplify their
+investigation. Our key contribution is the adaptation of the
+visualization principles of focus-plus-context and linking to the
+study of tree-structured data.
+
+Our motivating application is to the analysis of microbial time
+series, where an evolutionary tree relating microbes is available a
+priori. However, we have identified common problem types where, if a
+tree is not directly available, it can be constructed from data and
+then studied using our techniques. We perform detailed case studies to
+describe the alternative use cases, interpretations, and utility of
+the proposed visualization methods.
+
 \end{abstract}
@@
 \noindent%
-{\it Keywords:}  3 to 6 keywords, that do not appear in the title
+{\it Keywords:}  focus-plus-context, linking, time-series, tree-structured, R, D3
 \vfill

### 2016-12-06 treelapse.Rmd
Use pdf figures for antibiotic and pregnancy examples

@@
 ```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
-knitr::include_graphics("antibiotic_overview.png")
+knitr::include_graphics("figure/antibiotic_overview.pdf")
 ```
@@
 ```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
-knitr::include_graphics("antibiotic_bacteroidetes.png")
+knitr::include_graphics("figure/antibiotic_bacteroidetes.pdf")
 ```
@@
 ```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
-knitr::include_graphics("antibiotic_firmicutes.png")
+knitr::include_graphics("figure/antibiotic_firmicutes.png")
 ```
@@
 ```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders, with the except of "}
-knitr::include_graphics("ruminococcus.png")
+knitr::include_graphics("figure/ruminococcus.pdf")
 ```
@@
 In particular, they seem to increase after the first antibiotic time course."}
-knitr::include_graphics("verrucomicrobiae.png")
+knitr::include_graphics("figure/verrucomicrobiae.pdf")
 ```
@@
 ```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
-knitr::include_graphics("preterm_csts.png")
+knitr::include_graphics("figure/preterm_csts.pdf")
 ```
@@
 ```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
-knitr::include_graphics("preterm_csts_lacto.png")
+knitr::include_graphics("figure/preterm_csts_lacto.png")
 ```

### 2016-12-06 treelapse.Rmd
Update references for zillow figures

@@
 The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
-knitr::include_graphics("zillow_cities.png")
+knitr::include_graphics("figure/zillow_cities.pdf")
 ```
@@
 Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
-knitr::include_graphics("zillow_middle_pre.png")
+knitr::include_graphics("figure/zillow_middle_pre.pdf")
 ```
@@
 These appear to be located mainly in Los Angeles and San Diego counties."}
-knitr::include_graphics("zillow_middle_up.png")
+knitr::include_graphics("figure/zillow_middle_up.pdf")
 ```
@@
 These seem mostly to be located in Central California and the East San Francisco Bay Area."}
-knitr::include_graphics("zillow_middle_down.png")
+knitr::include_graphics("figure/zillow_middle_down.pdf")
 ```
@@
 ```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
-knitr::include_graphics("zillow_sf.png")
+knitr::include_graphics("figure/zillow_sf.pdf")
 ```

### 2016-12-06 treelapse.Rmd
Update bike sharing figures

@@
 ```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
-knitr::include_graphics("preterm_preterm.png")
+#oknitr::include_graphics("figure/preterm_preterm.png")
 ```
@@
 ```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
-knitr::include_graphics("working.png")
+knitr::include_graphics("figure/working.pdf")
 ```
@@
 This is the case for New Years' Eve, which is currently hovered over in the tree."}
-knitr::include_graphics("weekend.png")
+knitr::include_graphics("figure/weekend.pdf")
 ```
@@
 ```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
-knitr::include_graphics("weekday_2011.png")
+knitr::include_graphics("figure/weekday_2011.pdf")
 ```
@@
 ```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
-knitr::include_graphics("weekday_2012.png")
+knitr::include_graphics("figure/weekday_2012.pdf")
 ```
@@
 ```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
-knitr::include_graphics("warm_weekend.png")
+knitr::include_graphics("figure/warm_weekend.pdf")
 ```

### 2016-12-07 treelapse.Rmd
New pdfs for doi sankey on pregnancy data

@@
 ```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
-knitr::include_graphics("figure/preterm_csts_lacto.png")
+knitr::include_graphics("figure/preterm_csts_lacto.pdf")
 ```
@@
 ```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
-#oknitr::include_graphics("figure/preterm_preterm.png")
+knitr::include_graphics("figure/preterm_preterm.pdf")
 ```

### 2016-12-07 treelapse.Rmd
Fix reference to bike sankey

@@
 ```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
-knitr::include_graphics("bike_sankey.png")
+knitr::include_graphics("figure/bike_sankey.pdf")
 ```

### 2017-02-15 treelapse.Rmd
Add a section about the iris data

--- 2016-12-07 treelapse.Rmd
@@
 ```{r set_chunk_opts, echo = FALSE}
-knitr::opts_chunk$set(
+library("knitr")
+opts_chunk$set(
   fig.align = "center",
@@
 ```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
-knitr::include_graphics("figure/antibiotic_overview.pdf")
+include_graphics("figure/antibiotic_overview.pdf")
 ```
@@
 ```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
-knitr::include_graphics("figure/antibiotic_bacteroidetes.pdf")
+include_graphics("figure/antibiotic_bacteroidetes.pdf")
 ```
@@
 ```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
-knitr::include_graphics("figure/antibiotic_firmicutes.png")
+include_graphics("figure/antibiotic_firmicutes.png")
 ```
@@
 
-```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders, with the except of "}
-knitr::include_graphics("figure/ruminococcus.pdf")
+```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders."}
+include_graphics("figure/ruminococcus.pdf")
 ```
@@
 In particular, they seem to increase after the first antibiotic time course."}
-knitr::include_graphics("figure/verrucomicrobiae.pdf")
+include_graphics("figure/verrucomicrobiae.pdf")
 ```
@@
 ```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
-knitr::include_graphics("figure/preterm_csts.pdf")
+include_graphics("figure/preterm_csts.pdf")
 ```
@@
 ```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
-knitr::include_graphics("figure/preterm_csts_lacto.pdf")
+include_graphics("figure/preterm_csts_lacto.pdf")
 ```
@@
 ```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
-knitr::include_graphics("figure/preterm_preterm.pdf")
+include_graphics("figure/preterm_preterm.pdf")
 ```
@@
 California; rendering more neighborhoods while keeping all 246
-timepoints causes the interface to lag^[See the Section
+timepoints causes the interface to lag^[See Section
 \@ref(conclusion) for potential optimizations, however.].
@@
 The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
-knitr::include_graphics("figure/zillow_cities.pdf")
+include_graphics("figure/zillow_cities.pdf")
 ```
@@
 Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
-knitr::include_graphics("figure/zillow_middle_pre.pdf")
+include_graphics("figure/zillow_middle_pre.pdf")
 ```
@@
 These appear to be located mainly in Los Angeles and San Diego counties."}
-knitr::include_graphics("figure/zillow_middle_up.pdf")
+include_graphics("figure/zillow_middle_up.pdf")
 ```
@@
 These seem mostly to be located in Central California and the East San Francisco Bay Area."}
-knitr::include_graphics("figure/zillow_middle_down.pdf")
+include_graphics("figure/zillow_middle_down.pdf")
 ```
@@
 ```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
-knitr::include_graphics("figure/zillow_sf.pdf")
+include_graphics("figure/zillow_sf.pdf")
 ```
@@
 ```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
-knitr::include_graphics("figure/working.pdf")
+include_graphics("figure/working.pdf")
 ```
@@
 This is the case for New Years' Eve, which is currently hovered over in the tree."}
-knitr::include_graphics("figure/weekend.pdf")
+include_graphics("figure/weekend.pdf")
 ```
@@
 ```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
-knitr::include_graphics("figure/weekday_2011.pdf")
+include_graphics("figure/weekday_2011.pdf")
 ```
@@
 ```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
-knitr::include_graphics("figure/weekday_2012.pdf")
+include_graphics("figure/weekday_2012.pdf")
 ```
@@
 ```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
-knitr::include_graphics("figure/warm_weekend.pdf")
+include_graphics("figure/warm_weekend.pdf")
 ```
@@
 ```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
-knitr::include_graphics("bike_sankey.png")
+include_graphics("figure/bike_sankey.pdf")
 ```
+
+### Hierarchically Clustering the Iris Data {#iris}
+
+Each of the timebox tree and treebox examples presented so far have focused on
+data with a clear time component. We note however that these techniques could
+alternatively be applied to high-dimensional data, via the use of parallel
+coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
+remain, namely selecteing scales for and an ordering across the different
+coordinates, but the linking and focus-plus-context ideas can still be employed
+in this setting. Here we provide a simple implementation of this idea, using the
+famous Iris data, inspired by the hierarchical clustering vignette in
+[@galili2015dendextend].
+
+The Iris data set is a collection of 150 measurements of four variables --
+widths and lengths of Iris sepals and petals [@fisher1936use]. There are 50
+samples for each of three species of Iris, \textit{setosa}, \textit{versicolor}
+and \textit{virgnica}. As in [@galili2015dendextend], we hierarchical cluster
+samples\footnote{We use complete linkage and the euclidean distance.}. Our
+contribution is to display the resulting hierarchy together with a parallel
+coordinates view of the raw data, see Figure \@ref(fig:iristimebox).
+
+```{r iristimebox, fig.cap = "A toy application to the Iris data gives an example of how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
+include_graphics("iris_timebox.pdf")
+```
+
+From Figure \@ref(fig:iristimebox), the fact that the highest split in the
+hierarchical clustering distinguishes those samples with small petal lengths
+from the rest. Upon inspecting the associated IDs, we find that this cluster
+includes all the samples from setosa. The similarity between versicolor and
+virginica is reflected in their proximity in the clustering. Further, applying a
+treebox view to scan the subtrees associated with these two species reveals the
+more subtle differences in sepal and petal lengths that can be used to
+distinguish them.
 
@@
 
-Potential sections to write
-
-* describe extension from svg to canvas
-* give example of parallel coordinates / hierarchical clustering?
-* explain incorporation into / give example of shiny app?
-
 ## References ##

### 2017-02-15 treelapse.Rmd
Add iris figure

@@
 ```{r iristimebox, fig.cap = "A toy application to the Iris data gives an example of how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
-include_graphics("iris_timebox.pdf")
+include_graphics("figure/iris_timebox.pdf")
 ```

### 2017-02-15 treelapse.Rmd
Mention the github
Use png instead of pdf images

--- 2017-02-15 treelapse.Rmd
@@
 
-We have implemented our visualizations in D3, but encapsulated in an R
-package, called treelapse, to facilitate rapid turnover from data
-preparation and modeling to interactive exploration, and vice
-versa. We hope this package encourages data analysts to work at the
-border between data modeling and visualization, and more generally
-empowers a wider audience to apply less widely known, but powerful,
-visualization ideas.
+We have implemented our visualizations in D3, but encapsulated in an R package,
+called treelapse, to facilitate rapid turnover from data preparation and
+modeling to interactive exploration, and vice versa. Our code is open-source,
+and available at \url{https://github.com/krisrs1128/treelapse}. We hope this
+package encourages data analysts to work at the border between data modeling and
+visualization, and more generally empowers a wider audience to apply less widely
+known, but powerful, visualization ideas.
 
@@
 ```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
-include_graphics("figure/antibiotic_overview.pdf")
+include_graphics("figure/antibiotic_overview.png")
 ```
@@
 ```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
-include_graphics("figure/antibiotic_bacteroidetes.pdf")
+include_graphics("figure/antibiotic_bacteroidetes.png")
 ```
@@
 ```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders."}
-include_graphics("figure/ruminococcus.pdf")
+include_graphics("figure/ruminococcus.png")
 ```
@@
 In particular, they seem to increase after the first antibiotic time course."}
-include_graphics("figure/verrucomicrobiae.pdf")
+include_graphics("figure/verrucomicrobiae.png")
 ```
@@
 ```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
-include_graphics("figure/preterm_csts.pdf")
+include_graphics("figure/preterm_csts.png")
 ```
@@
 ```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
-include_graphics("figure/preterm_csts_lacto.pdf")
+include_graphics("figure/preterm_csts_lacto.png")
 ```
@@
 ```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
-include_graphics("figure/preterm_preterm.pdf")
+include_graphics("figure/preterm_preterm.png")
 ```
@@
 The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
-include_graphics("figure/zillow_cities.pdf")
+include_graphics("figure/zillow_cities.png")
 ```
@@
 Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
-include_graphics("figure/zillow_middle_pre.pdf")
+include_graphics("figure/zillow_middle_pre.png")
 ```
@@
 These appear to be located mainly in Los Angeles and San Diego counties."}
-include_graphics("figure/zillow_middle_up.pdf")
+include_graphics("figure/zillow_middle_up.png")
 ```
@@
 These seem mostly to be located in Central California and the East San Francisco Bay Area."}
-include_graphics("figure/zillow_middle_down.pdf")
+include_graphics("figure/zillow_middle_down.png")
 ```
@@
 ```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
-include_graphics("figure/zillow_sf.pdf")
+include_graphics("figure/zillow_sf.png")
 ```
@@
 ```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
-include_graphics("figure/working.pdf")
+include_graphics("figure/working.png")
 ```
@@
 This is the case for New Years' Eve, which is currently hovered over in the tree."}
-include_graphics("figure/weekend.pdf")
+include_graphics("figure/weekend.png")
 ```
@@
 ```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
-include_graphics("figure/weekday_2011.pdf")
+include_graphics("figure/weekday_2011.png")
 ```
@@
 ```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
-include_graphics("figure/weekday_2012.pdf")
+include_graphics("figure/weekday_2012.png")
 ```
@@
 ```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
-include_graphics("figure/warm_weekend.pdf")
+include_graphics("figure/warm_weekend.png")
 ```
@@
 ```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
-include_graphics("figure/bike_sankey.pdf")
+include_graphics("figure/bike_sankey.png")
 ```
@@
 ```{r iristimebox, fig.cap = "A toy application to the Iris data gives an example of how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
-include_graphics("iris_timebox.pdf")
+include_graphics("figure/iris_timebox.png")
 ```

### 2017-02-15 before_body.tex
Add funding source

@@
     and \\
-    Susan Holmes \\
+    Susan Holmes\thanks{
+       This research was supported in part by NIH grant R01 AI112401.
+      }\hspace{0.2cm}\\
     Department of Statistics, Stanford University}
@@
 \noindent%
-{\it Keywords:}  focus-plus-context, linking, time-series, tree-structured, R, D3
+{\it Keywords:}  D3, focus-plus-context, linking, time-series, tree-structured, R
 \vfill

### 2017-03-15 treelapse.Rmd

@@
   fig.align = "center",
-  out.width = "250px",
+  out.width = "350px",
   echo = FALSE,
@@
 questions that often emerge when working with hierarchical data and
-provide an R package to simplify their investigation [@ripley2001r].
+provide an R package to simplify their investigation [@ihaka1996r].
 
@@
 modeling to interactive exploration, and vice versa. Our code is open-source,
-and available at \url{https://github.com/krisrs1128/treelapse}. We hope this
-package encourages data analysts to work at the border between data modeling and
-visualization, and more generally empowers a wider audience to apply less widely
-known, but powerful, visualization ideas.
+and available at \url{xxxxxxxxx}\footnote{Link removed in blinded manuscript,
+but available to editor.}. We hope this package encourages data analysts to work
+at the border between data modeling and visualization, and more generally
+empowers a wider audience to apply less widely known, but powerful,
+visualization ideas.
 
@@
 
-### Hierarchically Clustering the Iris Data {#iris}
+### Hierarchically Clustering the Global Patterns Data {#global_patterns}
 
@@
 coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
-remain, namely selecteing scales for and an ordering across the different
+remain, namely selecting scales for and an ordering across the different
 coordinates, but the linking and focus-plus-context ideas can still be employed
-in this setting. Here we provide a simple implementation of this idea, using the
-famous Iris data, inspired by the hierarchical clustering vignette in
-[@galili2015dendextend].
+in this setting. Here we provide an implementation of this idea on a dataset
+comparing microbiomes across various ecological environments
+[@caporaso2011global], which is publicly accessible through phyloseq
+[@mcmurdie2013phyloseq].
 
-The Iris data set is a collection of 150 measurements of four variables --
-widths and lengths of Iris sepals and petals [@fisher1936use]. There are 50
-samples for each of three species of Iris, \textit{setosa}, \textit{versicolor}
-and \textit{virgnica}. As in [@galili2015dendextend], we hierarchical cluster
-samples\footnote{We use complete linkage and the euclidean distance.}. Our
-contribution is to display the resulting hierarchy together with a parallel
-coordinates view of the raw data, see Figure \@ref(fig:iristimebox).
+The original Global Patterns data consists of 26 samples across 9 environments
+(for example, freshwater and soil). In each site, there are counts across 19216
+taxa -- to simplify visualization, we filter to the 500 most abundant taxa.
 
-```{r iristimebox, fig.cap = "A toy application to the Iris data gives an example of how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
-include_graphics("figure/iris_timebox.png")
+We hierarchically cluster these 26 samples based on the 500 most abundant taxa,
+using complete linkage on the UniFrac distance. Figure \@ref(gptimebox) displays
+the resulting hierarchy together with a parallel coordinates view of the
+$\text{asinh}$ transformed taxa.
+
+```{r gptimebox, fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
+include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
 ```
 
-From Figure \@ref(fig:iristimebox), the fact that the highest split in the
-hierarchical clustering distinguishes those samples with small petal lengths
-from the rest. Upon inspecting the associated IDs, we find that this cluster
-includes all the samples from setosa. The similarity between versicolor and
-virginica is reflected in their proximity in the clustering. Further, applying a
-treebox view to scan the subtrees associated with these two species reveals the
-more subtle differences in sepal and petal lengths that can be used to
-distinguish them.
+In Figure \@ref(fig:gptimebox), we compare two subclusters from the hierarchical
+clustering tree, after zooming to a few of the microbes that distinguish between
+the clusters. Upon revisiting the original data, it becomes clear that the
+samples highlighted on the left come from freshwater samples, while those on the
+right come from soil and skin, and looking up taxonomic groups associated with
+the distinguishing microbes confirms this. For example, many of the microbes
+with high abundances in the left figure come from order Oceanospirillales.
 

### 2017-03-18 treelapse.Rmd
Replace microbe with bacteria

@@
 [@buja1996interactive, @becker1987brushing]. Our motivating
-application is to the analysis of microbial time series, where an
-evolutionary tree relating microbes is available a priori. However, we
+application is to the analysis of bacterial time series, where an
+evolutionary tree relating bacteria is available a priori. However, we
 have identified common problem types where, if a tree is not directly
@@
 modeling to interactive exploration, and vice versa. Our code is open-source,
-and available at \url{xxxxxxxxx}\footnote{Link removed in blinded manuscript,
+and available at \url{https://github.com/krisrs1128/treelapse}\footnote{Link removed in blinded manuscript,
 but available to editor.}. We hope this package encourages data analysts to work
@@
 
-A microbiome is a community of microbes living in given environments,
+A microbiome is a community of bacteria living in given environments,
 for example, ocean water or the human gut [@human2012structure,
@@
 rapidly accelerated by the advent of genomic technologies, which
-enable detailed quantification of microbial ecological structure and
+enable detailed quantification of bacterial ecological structure and
 its influence in human and environmental health. Being concerned with
-both microbial community structure and human health, the field exists
+both bacterial community structure and human health, the field exists
 at the border between ecology and medicine; consequently, papers in
@@
 dynamics problems. In the differential abundance problem, we attempt
-to compare the abundances of individual microbes across experimental
+to compare the abundances of individual bacteria across experimental
 conditions -- for example, treatment vs. control or healthy
@@
 analysis in genomics [@anders2010differential]. We prepend the
-description "tree-structured" because, in practice, researchers
-generate interpretations about intermediate taxonomic orders of
-microbes -- it is more interesting to discover novel behavior
-taxonomic levels between high-order phyla and low-level
-species. Hence, we frame the tree-structured differential abundance
-problem as the question of identifying the largest taxonomic subtree
-whose associated microbes are differentially abundant.
+description "tree-structured" because, in practice, researchers generate
+interpretations about intermediate taxonomic orders -- it is more interesting to
+discover novel behavior taxonomic levels between high-order phyla and low-level
+species. Hence, we frame the tree-structured differential abundance problem as
+the question of identifying the largest taxonomic subtree whose associated
+bacteria are differentially abundant.
 
-In the tree structured microbial dynamics problem, the goal is to
-describe changes in microbial abundances in an environment over
+In the tree structured bacterial dynamics problem, the goal is to
+describe changes in bacterial abundances in an environment over
 time. As in the differential abundance problem, it is useful if these
@@
 flavor. For example, researchers are often interested in understanding
-how microbial populations respond to sudden or gradual environmental
-changes or how microbes fill, drop out from, or compete for
+how bacterial populations respond to sudden or gradual environmental
+changes or how species fill, drop out from, or compete for
 environmental niches. Medically, these questions are important for
@@
 
-To unify the tree-structured differential abundance and microbial
-dynamic problems, we identify the data with a collection of random
+To unify the tree-structured differential abundance and bacterial
+dynamics problems, we identify the data with a collection of random
 variables indexed by nodes in a prespecified tree structure. In the
@@
 a node exhibits differential abundance when the coordinates are drawn
-from different distributions. On the other hand, in the microbial
+from different distributions. On the other hand, in the bacterial
 dynamics problem, each random variable is a time series, living in
@@
 
-Our third display is directed at the microbial dynamics
+Our third display is directed at the bacterial dynamics
 question. Here, two panels are arranged one over the other; one
@@
 associated with nodes on distant subtrees. This display is also suited
-for the microbial dynamics problem. For example, by highlighting all
+for the bacterial dynamics problem. For example, by highlighting all
 nodes at one taxonomic level in the tree, we can easily summarize the
@@
 see how much correlation and competition there is between
-taxonomically similar microbes. As in the timebox trees display, a
+taxonomically similar bacteria. As in the timebox trees display, a
 search box and pan-zoom scented widget are provided.
@@
 
-### Microbial Dynamics of Antibiotics Time Courses ###
+### Bacterial Dynamics of Antibiotics Time Courses ###
 
 [@dethlefsen2008pervasive] investigates the effect of antibiotics on
-microbial community composition from an ecological perspective. The
+bacterial community composition from an ecological perspective. The
 study tracks the microbiome of three patients across ten months, with
 two five-day antibiotic time courses separated by 6 months. Discerning
-the variation in resilience across microbes is important, considering
-the the role of microbes in health and not just disease.
+the variation in resilience across bacteria is important, considering
+the the role of bacteria in health and not just disease.
 
@@
 transformation. An initial view, Figure \@ref(fig:antibioticoverview),
-reveals two dramatic drops in the overall microbial abundance time
+reveals two dramatic drops in the overall bacterial abundance time
 series during the antibiotics time courses. Two more subtle effects
@@
 Alternatively, we can summarize each node by the average across its
-descendants -- this brings attention to individual microbes that may
+descendants -- this brings attention to individual bacteria that may
 be underlying some of the broader taxonomic patterns we have noted
@@
 because there are not many leaves associated with this phylum, the sum
-was small. Interestingly, the abundance of these microbes seems to
+was small. Interestingly, the abundance of these bacteria seems to
 *increase* after the first antibiotics treatment.  Be cautious,
@@
 
-### Differential Microbial Abundance and Preterm Births ###
+### Differential Bacterial Abundance and Preterm Births ###
 
-The study [@digiulio2015temporal] tracked the abundance of microbes in
+The study [@digiulio2015temporal] tracked the abundance of bacteria in
 the vaginal microbiome during pregnancy in an effort to study
-relationships between microbial community composition and preterm
-birth. Ideally, it would be possible to develop clear microbial
+relationships between bacterial community composition and preterm
+birth. Ideally, it would be possible to develop clear bacterial
 signatures associated with preterm births.
@@
 non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
-a heatmap, with microbes ordered according to a hierarchical
+a heatmap, with bacteria ordered according to a hierarchical
 clustering.  By using the DOI sankey instead, we can interpret the
@@
 identifying associations between preterm births and CSTs -- presumably
-because testing individual microbes loses power -- we can compare
-microbe abundances between preterm and non-preterm samples along
+because testing individual bacteria loses power -- we can compare
+bacterial abundances between preterm and non-preterm samples along
 subtrees, without requiring CSTs as an intermediary.
@@
 their values along the subtree. Specifically, we took the average of
-all samples within each CST to define values at the (microbe-level)
+all samples within each CST to define values at the (species-level)
 leaf nodes, and then aggregated the averages up to the root. It is
@@
 diversity. Further, focusing on the Lactobacillaceae family, we note
-that the differential abundance of these microbes distinguishes the
+that the differential abundance of these bacteria distinguishes the
 remaining CSTs, see Figure \@ref(fig:pretermcstslacto).
@@
 
-```{r pretermpreterm, fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
+```{r pretermpreterm, out.width = "300px", fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
 include_graphics("figure/preterm_preterm.png")
@@
 comparing microbiomes across various ecological environments
-[@caporaso2011global], which is publicly accessible through phyloseq
-[@mcmurdie2013phyloseq].
+[@caporaso2011global], which is publicly accessible through the phyloseq R
+package [@mcmurdie2013phyloseq].
 
@@
 We hierarchically cluster these 26 samples based on the 500 most abundant taxa,
-using complete linkage on the UniFrac distance. Figure \@ref(gptimebox) displays
+using complete linkage on the UniFrac distance. Figure \@ref(fig:gptimebox) displays
 the resulting hierarchy together with a parallel coordinates view of the
@@
 
-```{r gptimebox, fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
+```{r gptimebox, out.width = "200px", fig.show = "hold", fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
 include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
@@
 In Figure \@ref(fig:gptimebox), we compare two subclusters from the hierarchical
-clustering tree, after zooming to a few of the microbes that distinguish between
+clustering tree, after zooming to a few of the bacteria that distinguish between
 the clusters. Upon revisiting the original data, it becomes clear that the
@@
 right come from soil and skin, and looking up taxonomic groups associated with
-the distinguishing microbes confirms this. For example, many of the microbes
+the distinguishing bacteria confirms this. For example, many of the species
 with high abundances in the left figure come from order Oceanospirillales.
@@
 linked brushing can be practically incorporated into a range of
-practical analysis workflows, from studying the impact of microbes on
+practical analysis workflows, from studying the impact of bacteria on
 human health to better allocating units in commuter bikesharing

### 2017-03-18 treelapse.Rmd

@@
 
-In the tree structured bacterial dynamics problem, the goal is to
+In the tree-structured bacterial dynamics problem, the goal is to
 describe changes in bacterial abundances in an environment over
@@
 
-```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa."}
-include_graphics("figure/antibiotic_overview.png")
+```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa.
+We have annotated the figure with its three main components."}
+include_graphics("figure/annotated_antibiotic_overview.png")
 ```
@@
 
-```{r zillowsf, fig.cap = "To study teh range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
+```{r zillowsf, fig.cap = "To study the range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
 include_graphics("figure/zillow_sf.png")
@@
 
-```{r gptimebox, out.width = "200px", fig.show = "hold", fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
+```{r gptimebox, out.width = "250px", fig.show = "hold", fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
 include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
@@
 have leveraged a link between R and D3 [@vaidyanathan2014htmlwidgets]
-to create exploratory visualizations while still deep in data
-analysis; in this way our work is a departure from the culture of
-polished, journalistic visualizations prioritized by the D3
-community. Finally, we have given a series of examples to demonstrate
-how the general visualization techniques of focus-plus-context and
-linked brushing can be practically incorporated into a range of
-practical analysis workflows, from studying the impact of bacteria on
+to create visualizations during the exploratory phase of data analysis; in this
+way our work is a departure from the culture of polished, journalistic
+visualizations prioritized by the D3 community. Finally, we have given a series
+of examples to demonstrate how the general visualization techniques of
+focus-plus-context and linked brushing can be practically incorporated into a
+range of practical analysis workflows, from studying the impact of bacteria on
 human health to better allocating units in commuter bikesharing

### 2017-03-23 treelapse.Rmd
Write description for structSSI

@@
 
-[@dethlefsen2008pervasive] investigates the effect of antibiotics on
+@dethlefsen2008pervasive investigated the effect of antibiotics on
 bacterial community composition from an ecological perspective. The
@@
 
-The study [@digiulio2015temporal] tracked the abundance of bacteria in
+@digiulio2015temporal tracked the abundance of bacteria in
 the vaginal microbiome during pregnancy in an effort to study
@@
 
+### Inspecting Confirmatory Analysis {#structssi}
+
+In addition to facilitating exploratory study, treelapse has potential value as
+a device for inspecting confirmatory analysis. We provide an illustration
+extending an example from [@callahan2016bioconductor], which formally tested
+bacterial species for association with age in a sample of mice. The testing
+approach advocated there is particularly well-suited to visualization with
+treelapse, since it sought to detect associations at multiple levels of
+phylogenetic resolution, using statistical tools developed in
+[@yekutieli2008hierarchical; @sankaran2014structssi].
+
+The data of interest in [@callahan2016bioconductor] are bacterial counts
+collected across old and young mice. After variance-stabilizing these counts
+using DESeq2 [@love2014moderated], a $t$-test was applied to each node in a
+phylogenetic tree, comparing abundances between old and young mice. To account
+for multiple testing, we employ the structSSI algorithm
+[@yekutieli2008hierarchical; @sankaran2014structssi] along with methods
+available in the `multtest` package [@pollard2005multiple].
+
+To interpret the results, we apply timebox trees. Our goals are to (1) identify
+subtrees with consistently elevated differential abundance across age groups
+and (2) compare alternative multiple testing adjustment procedures. Our approach
+is to display the negative-log raw and adjusted $p$-values for each node, with
+alternative methods compared via parallel coordinates. One view of the resulting
+display is captured in Figure \@ref(fig:structssi). First, we see that
+significant nodes tend to be significant across all methods -- the ordering
+between different series appears stable. Interestingly, the Sidak one-step and
+structSSI procedures seem to have lower power than the others, including
+conservative FWER-controlling methods, like the Bonferroni procedure. Further,
+in this application, FDR-controlling techniques do not seem to offer notably
+different adjusted $p$-values, relative to those controlling FWER. This suggests
+that, for this problem, bacteria are either strongly associated with age, or not
+associated at all, so that there is little gain from using more sensitive
+procedures.
+
+```{r structssi, fig.cap = "Viewing a tree of $p$-values across different methods highlights two subtrees with strong associations with mouse age, across several testing procedures."}
+include_graphics("figure/structssi.png")
+```
+
+Further, selecting series with strong association between abundance and age, two
+major subtrees are brought to the forefront. Separately querying the the
+taxonomic identities of these bacteria reveals that they are two subgroups of
+Clostridia, which is consistent with the analysis of
+[@callahan2016bioconductor]. More than this specific analysis outcome, this view
+demonstrates that interactive visual inspection of results from confirmatory
+analysis provides deeper insight than the standard practice of printing tables
+of (adjusted or unadjusted) $p$-values: the relationship between significant
+nodes is only clear upon visualization on the tree.
+
 ## Conclusion and Future Work {#conclusion}

### 2017-03-23 treelapse.tex

--- 2017-03-23 treelapse.Rmd
@@
----
-standalone: true
-documentclass: article
-fontsize: 12pt
-output:
-  bookdown::pdf_book:
-    keep_tex: yes
-    toc: false
-    includes:
-      in_header: preamble.tex
-      before_body: before_body.tex
-bibliography: ["bibliography.bib"]
-biblio-style: "agsm"
-link-citations: yes
----
+\documentclass[12pt]{article}
+\usepackage{hyperref}
 
-## Introduction ##
+\usepackage{amsmath}
+\usepackage{graphicx,psfrag,epsf}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL
 
-```{r set_chunk_opts, echo = FALSE}
-library("knitr")
-opts_chunk$set(
-  fig.align = "center",
-  out.width = "350px",
-  echo = FALSE,
-  cache = TRUE
-)
-```
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
+
+\begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+{
+  \title{\bf Interactive Visualization of Hierarchically Structured Data}
+  \author{Kris Sankaran\thanks{
+      This research was supported in part by an NIH training grant (5T32GM096982-03).}\hspace{.2cm}\\
+    Department of Statistics, Stanford University\\
+    and \\
+    Susan Holmes\thanks{
+       This research was supported in part by NIH grant R01 AI112401.
+      }\hspace{0.2cm}\\
+    Department of Statistics, Stanford University}
+  \maketitle
+} \fi
+
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Title}
+\end{center}
+  \medskip
+} \fi
+
+\bigskip
+\begin{abstract}
+We introduce methods for visualization of data structured along trees,
+especially hierarchically structured collections of time series.  To
+this end, we identify questions that often emerge when working with
+hierarchical data and provide an R package to simplify their
+investigation. Our key contribution is the adaptation of the
+visualization principles of focus-plus-context and linking to the
+study of tree-structured data.
+
+Our motivating application is to the analysis of microbial time
+series, where an evolutionary tree relating microbes is available a
+priori. However, we have identified common problem types where, if a
+tree is not directly available, it can be constructed from data and
+then studied using our techniques. We perform detailed case studies to
+describe the alternative use cases, interpretations, and utility of
+the proposed visualization methods.
+
+\end{abstract}
+
+\noindent%
+{\it Keywords:}  D3, focus-plus-context, linking, time-series, tree-structured, R
+\vfill
+
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
+
+\subsection*{Introduction}\label{introduction}
 
 We introduce methods for visualization of data structured along trees,
-especially hierarchically structured collections of time series. We
-hope both to characterize generically useful techniques for
-interactively visualizing hierarchical data and to offer practical
-tools for implementing such displays. To this end, we identify
-questions that often emerge when working with hierarchical data and
-provide an R package to simplify their investigation [@ihaka1996r].
+especially hierarchically structured collections of time series. We hope
+both to characterize generically useful techniques for interactively
+visualizing hierarchical data and to offer practical tools for
+implementing such displays. To this end, we identify questions that
+often emerge when working with hierarchical data and provide an R
+package to simplify their investigation \citep{ihaka1996r}.
 
@@
 of focus-plus-context and linking to the study of tree-structured data
-[@buja1996interactive, @becker1987brushing]. Our motivating
+\citep{buja1996interactive, becker1987brushing}. Our motivating
 application is to the analysis of bacterial time series, where an
@@
 
-We have implemented our visualizations in D3, but encapsulated in an R package,
-called treelapse, to facilitate rapid turnover from data preparation and
-modeling to interactive exploration, and vice versa. Our code is open-source,
-and available at \url{https://github.com/krisrs1128/treelapse}\footnote{Link removed in blinded manuscript,
-but available to editor.}. We hope this package encourages data analysts to work
-at the border between data modeling and visualization, and more generally
-empowers a wider audience to apply less widely known, but powerful,
-visualization ideas.
+We have implemented our visualizations in D3, but encapsulated in an R
+package, called treelapse, to facilitate rapid turnover from data
+preparation and modeling to interactive exploration, and vice versa. Our
+code is open-source, and available at
+\url{https://github.com/krisrs1128/treelapse}\footnote{Link removed in blinded manuscript,
+but available to editor.}. We hope this package encourages data analysts
+to work at the border between data modeling and visualization, and more
+generally empowers a wider audience to apply less widely known, but
+powerful, visualization ideas.
 
 The paper is organized as follows. First, we describe our motivating
-application to the microbiome and the associated generic analysis
-tasks. Next, we review the underlying visualization principles behind
-our contributions. Then we then connect these principles to analysis
-tasks we identified earlier, describing in detail the visualization
-methods we have implemented in treelapse. We close with several case
-studies using publicly available data across both microbiome and
-non-microbiome related applications.
+application to the microbiome and the associated generic analysis tasks.
+Next, we review the underlying visualization principles behind our
+contributions. Then we then connect these principles to analysis tasks
+we identified earlier, describing in detail the visualization methods we
+have implemented in treelapse. We close with several case studies using
+publicly available data across both microbiome and non-microbiome
+related applications.
 
-### Problem Motivation ###
+\subsubsection*{Problem Motivation}\label{problem-motivation}
 
 A microbiome is a community of bacteria living in given environments,
-for example, ocean water or the human gut [@human2012structure,
-@human2012structure, @cho2012human]. Progress in the field has been
-rapidly accelerated by the advent of genomic technologies, which
-enable detailed quantification of bacterial ecological structure and
-its influence in human and environmental health. Being concerned with
-both bacterial community structure and human health, the field exists
-at the border between ecology and medicine; consequently, papers in
-the area often apply a blend of exploratory data analysis and formal
-statistical inference.
+for example, ocean water or the human gut 
+\citep{human2012structure, cho2012human}. Progress in the field has
+been rapidly accelerated by the advent of genomic technologies, which
+enable detailed quantification of bacterial ecological structure and its
+influence in human and environmental health. Being concerned with both
+bacterial community structure and human health, the field exists at the
+border between ecology and medicine; consequently, papers in the area
+often apply a blend of exploratory data analysis and formal statistical
+inference.
 
 The two essential microbiome analysis problems that motivated our work
-are the tree-structured differential abundance and differential
-dynamics problems. In the differential abundance problem, we attempt
-to compare the abundances of individual bacteria across experimental
-conditions -- for example, treatment vs. control or healthy
-vs. diseased. This is the microbiome analog of differential expression
-analysis in genomics [@anders2010differential]. We prepend the
-description "tree-structured" because, in practice, researchers generate
-interpretations about intermediate taxonomic orders -- it is more interesting to
-discover novel behavior taxonomic levels between high-order phyla and low-level
-species. Hence, we frame the tree-structured differential abundance problem as
-the question of identifying the largest taxonomic subtree whose associated
-bacteria are differentially abundant.
+are the tree-structured differential abundance and differential dynamics
+problems. In the differential abundance problem, we attempt to compare
+the abundances of individual bacteria across experimental conditions --
+for example, treatment vs. control or healthy vs. diseased. This is the
+microbiome analog of differential expression analysis in genomics
+\citep{anders2010differential}.
+We prepend the description ``tree-structured'' because, in practice,
+researchers generate interpretations about intermediate taxonomic orders
+-- it is more interesting to discover novel behavior taxonomic levels
+between high-order phyla and low-level species. Hence, we frame the
+tree-structured differential abundance problem as the question of
+identifying the largest taxonomic subtree whose associated bacteria are
+differentially abundant.
 
 In the tree-structured bacterial dynamics problem, the goal is to
-describe changes in bacterial abundances in an environment over
-time. As in the differential abundance problem, it is useful if these
+describe changes in bacterial abundances in an environment over time. As
+in the differential abundance problem, it is useful if these
 descriptions can be given at the highest subtree at which the pattern
-appears. Specific questions of interest often have an ecological
-flavor. For example, researchers are often interested in understanding
-how bacterial populations respond to sudden or gradual environmental
-changes or how species fill, drop out from, or compete for
-environmental niches. Medically, these questions are important for
-illuminating the impact of antibiotic time courses or diet changes,
-for example.
+appears. Specific questions of interest often have an ecological flavor.
+For example, researchers are often interested in understanding how
+bacterial populations respond to sudden or gradual environmental changes
+or how species fill, drop out from, or compete for environmental niches.
+Medically, these questions are important for illuminating the impact of
+antibiotic time courses or diet changes, for example.
 
-### Problem Abstraction ###
+\subsubsection*{Problem Abstraction}\label{problem-abstraction}
 
@@
 differential abundance problem, each random variable lives in
-$\mathbb{R}^{G}$ where $G$ is the number of groups being
-compared. Each coordinate represents the abundance for that group, and
-a node exhibits differential abundance when the coordinates are drawn
-from different distributions. On the other hand, in the bacterial
-dynamics problem, each random variable is a time series, living in
-$\mathbb{R}^{T}$.
+\(\mathbb{R}^{G}\) where \(G\) is the number of groups being compared.
+Each coordinate represents the abundance for that group, and a node
+exhibits differential abundance when the coordinates are drawn from
+different distributions. On the other hand, in the bacterial dynamics
+problem, each random variable is a time series, living in
+\(\mathbb{R}^{T}\).
 
@@
 according to the value of the children nodes: we define the value at
-each node to be either the sum or average of all descendant
-tips. However, it is possible to imagine situations where the internal
-nodes are drawn from their own distribution, unconstrained by
-descendants. In general, analysis in this abstraction focuses on
-describing the distribution of these random variables as a function of
-their position across subsets of the tree. The essential difficulty in
-these problems is high-dimensionality -- there are many tree nodes,
-each holding a vector-valued random variable. Even simply navigating
-across the tree and comparing coordinates in the observed variables is
-a challenge; ideally we could construct a succinct representation of
-the essential covariation across subtrees and coordinates.
+each node to be either the sum or average of all descendant tips.
+However, it is possible to imagine situations where the internal nodes
+are drawn from their own distribution, unconstrained by descendants. In
+general, analysis in this abstraction focuses on describing the
+distribution of these random variables as a function of their position
+across subsets of the tree. The essential difficulty in these problems
+is high-dimensionality -- there are many tree nodes, each holding a
+vector-valued random variable. Even simply navigating across the tree
+and comparing coordinates in the observed variables is a challenge;
+ideally we could construct a succinct representation of the essential
+covariation across subtrees and coordinates.
 
@@
 which have a priori known tree structures. For example, collections of
-spatially-indexed time series are frequently encountered in practice
--- consider energy consumption, product sales, or high school dropout
-rates across regional districts. This type of data has an implicit
-tree structure -- at the top level are different states, while at the
-bottom are individual census tracts, say. Analysis here revolves
-around the question of how variation across time series is related to
-their geographic position.  A case study to this type of data is given
-in Section \@ref(zillow-study).
+spatially-indexed time series are frequently encountered in practice --
+consider energy consumption, product sales, or high school dropout rates
+across regional districts. This type of data has an implicit tree
+structure -- at the top level are different states, while at the bottom
+are individual census tracts, say. Analysis here revolves around the
+question of how variation across time series is related to their
+geographic position. A case study to this type of data is given in
+Section \ref{zillow-study}.
 
 Alternatively, if this type of hierarchical contextual information is
-not directly available, a tree structure can be learned from the
-data. This could be achieved by learning a hierarchical clustering on
-the original series. Further, if contextual information is available,
-but it is not hierarchical, it is possible to setup a supervised
-problem that uses context to predict features of the time series. We
-can construct a tree by applying a tree-based classifier
-[@breiman1984classification] or extracting a regression tree from a
-more complex supervised model [@boz2002extracting,
-@saito2002extracting]. Analysis then focuses on how different
-partitions of the contextual, covariate space relate to observed time
-series. This approach is described in Section \@ref(bikesharing-study).
+not directly available, a tree structure can be learned from the data.
+This could be achieved by learning a hierarchical clustering on the
+original series. Further, if contextual information is available, but it
+is not hierarchical, it is possible to setup a supervised problem that
+uses context to predict features of the time series. We can construct a tree by
+applying a tree-based classifier \citep{breiman1984classification} or extracting
+a regression tree from a more complex supervised model
+\citep{boz2002extracting,saito2002extracting}. Analysis then focuses on how
+different partitions of the contextual, covariate space relate to observed time
+series. This approach is described in Section \ref{bikesharing-study}.
 
 Finally, note that, while we have focused on time series valued nodes,
-all of this discussion could be translated to studying
-high-dimensional data via parallel coordinates
-[@inselberg1991parallel]. The usual parallel coordinates challenges
-remain, mainly selecting scales for and ordering across the different
-coordinates, but the linking and focus-plus-context can still be
-employed this setting.
+all of this discussion could be translated to studying high-dimensional
+data via parallel coordinates \citep{inselberg1991parallel}. The usual parallel
+coordinates challenges remain, mainly selecting scales for and ordering
+across the different coordinates, but the linking and focus-plus-context
+can still be employed this setting.
 
-## Background Literature and Solution Principles ##
+\subsection*{Background Literature and Solution
+Principles}\label{background-literature-and-solution-principles}
 
 Now that we have specified the essential questions of interest, we
-survey some ideas from the visualization literature that can be
-applied to answer them. As the core difficulty is high-dimensionality,
-so it should be no surprise that the techniques we adapt come from the
+survey some ideas from the visualization literature that can be applied
+to answer them. As the core difficulty is high-dimensionality, so it
+should be no surprise that the techniques we adapt come from the
 literature on high-dimensional data visualization, namely,
@@
 The focus-plus-context principle is that large collections of visual
-elements can be studied at multiple scales, by simultaneously
-focusing" on a few elements of interest and maintaining the "context"
-of the coarser-scale background. A simple example of this idea is to
-include a search box that highlights matching samples (focus) and
-mutes the rest (context). Two more sophisticated methods anchored in
-this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
-central to the proposals in treelapse [@hochheiser2004dynamic,
-@heer2004doitrees]. In timeboxes, a collection of time series are
-graphically queried using interactive brushes. Series that pass
-through all of the user-specified brushes are highlighted, and the
-rest are faded to the background. Hence, time series meeting the
-constraints imposed by the brushes are focused, while the remainder
-are de-emphasized, though they remain present as context. This method
-can be interpreted programmatically as the visual analog of a database
-query, or probabilistically as the conditional distribution for the
-full series, given it passes through certain bounds.
+elements can be studied at multiple scales, by simultaneously focusing"
+on a few elements of interest and maintaining the ``context'' of the
+coarser-scale background. A simple example of this idea is to include a
+search box that highlights matching samples (focus) and mutes the rest
+(context). Two more sophisticated methods anchored in this idea are
+timeboxes and Degree-of-Interest (DOI) trees; both are central to the proposals
+in treelapse \citep{hochheiser2004dynamic, heer2004doitrees}. In timeboxes, a
+collection of time series are graphically queried using interactive brushes.
+Series that pass through all of the user-specified brushes are highlighted, and
+the rest are faded to the background. Hence, time series meeting the constraints
+imposed by the brushes are focused, while the remainder are de-emphasized,
+though they remain present as context. This method can be interpreted
+programmatically as the visual analog of a database query, or probabilistically
+as the conditional distribution for the full series, given it passes through
+certain bounds.
 
@@
 on the fringes as context. The implementation is modularized into two
-tasks -- the determination of a DOI distribution over nodes in the
-tree and visual layout of a tree given DOI assignments. The DOI
-distribution used in [@heer2004doitrees] places maximal interest on
-the node that the user had most recently clicked, along with all
-ancestors. The DOI for all other nodes is defined as the distance to
-the closest maximal interest node. The layout step then trims
-low-interest subtrees until the remaining nodes fit within a given
-screen size. By adjusting the minimal DOI below which nodes are
-hidden, the user can transition between node-specific and full-tree
-scales.
+tasks -- the determination of a DOI distribution over nodes in the tree
+and visual layout of a tree given DOI assignments. The DOI distribution
+used in \citep{heer2004doitrees}
+places maximal interest on the node that the user had most recently
+clicked, along with all ancestors. The DOI for all other nodes is
+defined as the distance to the closest maximal interest node. The layout
+step then trims low-interest subtrees until the remaining nodes fit
+within a given screen size. By adjusting the minimal DOI below which
+nodes are hidden, the user can transition between node-specific and
+full-tree scales.
 
@@
 application is to linked scatterplot brushing 
-[@becker1987brushing]. Here, a scatterplot matrix gives the
-relationship between all pairs of variables.  Points brushed in one
-scatterplot are then highlighted in all others. For example, this
-helps the user determine whether an outlier in one dimension is an
-outlier in others. Another instance of this idea links the results of
-dimensionality reduction methods to displays of the raw data, as
-implemented by XGobi and Cranvas, for example [@xie2013cranvas,
-@swayne1998xgobi]. As in timeboxes, linking can be interpreted as
-database queries or conditional probabilities: given a subset of the
-series after conditioning on the values for one set of features, what
-are the values for a second set [@buja1996interactive]?
+\citep{becker1987brushing}. Here, a scatterplot
+matrix gives the relationship between all pairs of variables. ~Points
+brushed in one scatterplot are then highlighted in all others. For
+example, this helps the user determine whether an outlier in one
+dimension is an outlier in others. Another instance of this idea links
+the results of dimensionality reduction methods to displays of the raw
+data, as implemented by XGobi and Cranvas, for example 
+\citep{xie2013cranvas, swayne1998xgobi}. As in timeboxes,
+linking can be interpreted as database queries or conditional
+probabilities: given a subset of the series after conditioning on the
+values for one set of features, what are the values for a second set
+\citep{buja1996interactive}?
 
-Finally, unrelated to established visualization principles, we note
-that our work is deliberately grounded in the R software
-ecosystem. This connection is made using the htmlwidgets package
-[@vaidyanathan2014htmlwidgets]. Not only does linking R with D3 make
-these visualization methods more broadly accessible, we hope to
-facilitate exchange between data modeling and interactive
-visualization. Moreover, our tools are intentionally limited in scope
--- designed to facilitate this dialog for a specific class of
+Finally, unrelated to established visualization principles, we note that
+our work is deliberately grounded in the R software ecosystem. This
+connection is made using the htmlwidgets package 
+\citep{vaidyanathan2014htmlwidgets}. Not only
+does linking R with D3 make these visualization methods more broadly
+accessible, we hope to facilitate exchange between data modeling and
+interactive visualization. Moreover, our tools are intentionally limited
+in scope -- designed to facilitate this dialog for a specific class of
 problems, rather than providing a toolbox for generic types of
-visualization design. We believe that this narrow context within a
-broad ecosystem strikes a balance between problem-specificity and
-ease-of-use.
+visualization design. We believe that this narrow context within a broad
+ecosystem strikes a balance between problem-specificity and ease-of-use.
 
-## Specific Proposals ##
+\subsection*{Specific Proposals}\label{specific-proposals}
 
 Our first proposed visualization technique is a minor modification of
-the DOI tree. The standard DOI tree definition does not have any
-notion of data defined at nodes, it is only used a device for
-navigating tree structures. A trivial extension can encode scalar data
-at nodes: have the node radius reflect the associated scalar value. To
-reinforce this effect, we can adjust the width of the parent
-edge. When parent nodes have values equal to the sum of their
-children, this creates the effect of values "flowing" from the root to
-leaves. To help viewers make use of their domain knowledge, we have
-included a search box that highlights paths to nodes with matching
-terms. Edges are ordered from widest on the left to narrowest on the
-right. While this method can only represent a single scalar-value per
-tree node, it suggests an approach to the tree-structured differential
-abundance problem, which we call the DOI sankey.
+the DOI tree. The standard DOI tree definition does not have any notion
+of data defined at nodes, it is only used a device for navigating tree
+structures. A trivial extension can encode scalar data at nodes: have
+the node radius reflect the associated scalar value. To reinforce this
+effect, we can adjust the width of the parent edge. When parent nodes
+have values equal to the sum of their children, this creates the effect
+of values ``flowing'' from the root to leaves. To help viewers make use
+of their domain knowledge, we have included a search box that highlights
+paths to nodes with matching terms. Edges are ordered from widest on the
+left to narrowest on the right. While this method can only represent a
+single scalar-value per tree node, it suggests an approach to the
+tree-structured differential abundance problem, which we call the DOI
+sankey.
 
@@
 and control groups at each tree node. Every edge in the tree is split
-into two colors^[We use the colorbrewer palette to facilitate
-readability [@brewer2003colorbrewer]], with relative widths of the
-different colors reflecting differences in sizes for the two
-groups. The overall width of each edge represents the sum of values
+into two colors\footnote{We use the colorbrewer palette to facilitate
+  readability \citep{brewer2003colorbrewer}}, with relative
+widths of the different colors reflecting differences in sizes for the
+two groups. The overall width of each edge represents the sum of values
 across all groups.
@@
 node and a single group, first compute the average abundance at that
-node among all samples in that group. This will give the width for
-that group's color on the edge leading to the specified
-node. Differentially abundant subtrees then correspond to subtrees
-where some colors occupy more space than others. That is, this
-representation makes it easier to identify points where the "flows"
-for different groups diverge -- the colors begin to separate. The DOI
-principle assists navigation across the tree structure, allowing focus
-on individual flow structures without losing broader tree context.
+node among all samples in that group. This will give the width for that
+group's color on the edge leading to the specified node. Differentially
+abundant subtrees then correspond to subtrees where some colors occupy
+more space than others. That is, this representation makes it easier to
+identify points where the ``flows'' for different groups diverge -- the
+colors begin to separate. The DOI principle assists navigation across
+the tree structure, allowing focus on individual flow structures without
+losing broader tree context.
 
-Our third display is directed at the bacterial dynamics
-question. Here, two panels are arranged one over the other; one
-displays all time series, while the other displays all tree nodes,
-with node sizes reflecting the time series value at that node. For
-this reason, we call the display, Timebox Trees. In the time series
-panel, we have directly implemented the timeboxes idea. We then link
-the panels: when a set of series is highlighted by the timeboxes, the
-associated tree nodes are also highlighted. For example, timeboxes can
-be used to focus on a set of series with specific shape -- increased
-abundance after an ecological shock, for example -- and identify along
-what subtrees this pattern is present. To further focus on specific
-elements, a pan-zoom scented widget is provided
-[@willett2007scented]. The widget is a miniature version of the full
-time series panel, equipped with a single brush whose extent specifies
-the limits in the main time series panel. As in the DOI trees and
-sankeys, a search bar can be used to highlight those series of
-interest a priori.
+Our third display is directed at the bacterial dynamics question. Here,
+two panels are arranged one over the other; one displays all time
+series, while the other displays all tree nodes, with node sizes
+reflecting the time series value at that node. For this reason, we call
+the display, Timebox Trees. In the time series panel, we have directly
+implemented the timeboxes idea. We then link the panels: when a set of
+series is highlighted by the timeboxes, the associated tree nodes are
+also highlighted. For example, timeboxes can be used to focus on a set
+of series with specific shape -- increased abundance after an ecological
+shock, for example -- and identify along what subtrees this pattern is
+present. To further focus on specific elements, a pan-zoom scented
+widget is provided \citep{willett2007scented}. The widget is a miniature version
+of the full time series panel, equipped with a single brush whose extent
+specifies the limits in the main time series panel. As in the DOI trees and
+sankeys, a search bar can be used to highlight those series of interest a
+priori.
 
@@
 focusing on all the children below a single node makes it possible to
-see how much correlation and competition there is between
-taxonomically similar bacteria. As in the timebox trees display, a
-search box and pan-zoom scented widget are provided.
+see how much correlation and competition there is between taxonomically
+similar bacteria. As in the timebox trees display, a search box and
+pan-zoom scented widget are provided.
 
-## Case Studies ##
+\subsection*{Case Studies}\label{case-studies}
 
-We now delve into applications on real data. Our goals are to
-illustrate potential workflows that incorporate treelapse, describe
-the formulation of questions that can be naturally investigated with
-our methods, and provide example interpretations on treelapse
-output. Our examples are also chosen to reflect the range of problem
-domains to which the package can be applied -- though it was motivated
-by applications to the microbiome, it is not tied to it. More
-importantly, we argue that the visualization principles reviewed above
-can substantively improve the practice of data analysis in the class
-of problems to which we have limited ourselves.
+We now delve into applications on real data. Our goals are to illustrate
+potential workflows that incorporate treelapse, describe the formulation
+of questions that can be naturally investigated with our methods, and
+provide example interpretations on treelapse output. Our examples are
+also chosen to reflect the range of problem domains to which the package
+can be applied -- though it was motivated by applications to the
+microbiome, it is not tied to it. More importantly, we argue that the
+visualization principles reviewed above can substantively improve the
+practice of data analysis in the class of problems to which we have
+limited ourselves.
 
-### Bacterial Dynamics of Antibiotics Time Courses ###
+\subsubsection*{Bacterial Dynamics of Antibiotics Time
+Courses}\label{bacterial-dynamics-of-antibiotics-time-courses}
 
-@dethlefsen2008pervasive investigated the effect of antibiotics on
-bacterial community composition from an ecological perspective. The
-study tracks the microbiome of three patients across ten months, with
-two five-day antibiotic time courses separated by 6 months. Discerning
-the variation in resilience across bacteria is important, considering
-the the role of bacteria in health and not just disease.
+\citet{dethlefsen2008pervasive} investigated the
+effect of antibiotics on bacterial community composition from an
+ecological perspective. The study tracks the microbiome of three
+patients across ten months, with two five-day antibiotic time courses
+separated by 6 months. Discerning the variation in resilience across
+bacteria is important, considering the the role of bacteria in health
+and not just disease.
 
 We approach the data using the linked time and treebox views, after
-first filtering low variance taxa and taking an $\text{asinh}$
-transformation. An initial view, Figure \@ref(fig:antibioticoverview),
+first filtering low variance taxa and taking an \(\text{asinh}\)
+transformation. An initial view, Figure \ref{fig:antibioticoverview},
 reveals two dramatic drops in the overall bacterial abundance time
-series during the antibiotics time courses. Two more subtle effects
-are also suggested from this view,
+series during the antibiotics time courses. Two more subtle effects are
+also suggested from this view,
 
-* The second antibiotic treatment seems to have a more lasting effect, as the
-series take longer to return to their original values.
-* Some high level taxa appear relatively unaffected by the first
+\begin{itemize}
+\item
+  The second antibiotic treatment seems to have a more lasting effect,
+  as the series take longer to return to their original values.
+\item
+  Some high level taxa appear relatively unaffected by the first
   antibiotic treatment. By more closely inspecting the display, we are
   able to identify these as members of the Bacteroidetes phylum, see
-Figure \@ref(fig:antibioticbacteroidetes).
+  Figure \ref{fig:antibioticbacteroidetes}.
+\end{itemize}
 
-```{r antibioticoverview, fig.cap = "The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa.
-We have annotated the figure with its three main components."}
-include_graphics("figure/annotated_antibiotic_overview.png")
-```
+\begin{figure}
 
-```{r antibioticbacteroidetes, fig.cap = "Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics."}
-include_graphics("figure/antibiotic_bacteroidetes.png")
-```
+{\centering \includegraphics[width=350px]{figure/annotated_antibiotic_overview}
 
-Next, using the scented widget, we focus on the window around the
-second antibiotic treatment. We apply the treebox display to compare
-then behavior of different families of Firmicutes, Lachnospiraceae and
-Ruminoccocus. We suspect that these taxa are associated with the
-delayed recovery after the second time course. To investigate this, we
-input these family names in the search box to isolate their positions
-on the tree; then we apply brushes to highlight the series that
-contribute to these higher-level families. The resulting view is given
-by Figure \@ref(fig:antibioticfirmicutes)
+}
 
-```{r antibioticfirmicutes, fig.cap = "Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels."}
-include_graphics("figure/antibiotic_firmicutes.png")
-```
+\caption{The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa.
+We have annotated the figure with its three main components.}\label{fig:antibioticoverview}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/antibiotic_bacteroidetes}
+
+}
+
+\caption{Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics.}\label{fig:antibioticbacteroidetes}
+\end{figure}
+
+Next, using the scented widget, we focus on the window around the second
+antibiotic treatment. We apply the treebox display to compare then
+behavior of different families of Firmicutes, Lachnospiraceae and
+Ruminoccocus. We suspect that these taxa are associated with the delayed
+recovery after the second time course. To investigate this, we input
+these family names in the search box to isolate their positions on the
+tree; then we apply brushes to highlight the series that contribute to
+these higher-level families. The resulting view is given by Figure
+\ref{fig:antibioticfirmicutes}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/antibiotic_firmicutes}
+
+}
+
+\caption{Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels.}\label{fig:antibioticfirmicutes}
+\end{figure}
 
 Alternatively, we can summarize each node by the average across its
-descendants -- this brings attention to individual bacteria that may
-be underlying some of the broader taxonomic patterns we have noted
-when studying the subtree sums. For example, in Figure
-\@ref(fig:ruminococcus), we highlight all families below order
+descendants -- this brings attention to individual bacteria that may be
+underlying some of the broader taxonomic patterns we have noted when
+studying the subtree sums. For example, in Figure
+\ref{fig:ruminococcus}, we highlight all families below order
 Ruminococcus, suggesting that the decrease due to antibiotics occurs
@@
 
-```{r ruminococcus, fig.cap = "By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders."}
-include_graphics("figure/ruminococcus.png")
-```
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/ruminococcus}
+
+}
+
+\caption{By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders.}\label{fig:ruminococcus}
+\end{figure}
 
@@
 changed. This is because, at each branching point, we place the node
-with larger average value on the left. This places the
-Verrucomicrobiae at the far left, which seem to have large abundance
-over many time points. This phylum had been previously obfuscated --
-because there are not many leaves associated with this phylum, the sum
-was small. Interestingly, the abundance of these bacteria seems to
-*increase* after the first antibiotics treatment.  Be cautious,
-however, that the average over only a few Verrucomicrobiae species
-will be a more variable estimate than the averages over the more
-prevalent phyla.
+with larger average value on the left. This places the Verrucomicrobiae
+at the far left, which seem to have large abundance over many time
+points. This phylum had been previously obfuscated -- because there are
+not many leaves associated with this phylum, the sum was small.
+Interestingly, the abundance of these bacteria seems to \emph{increase}
+after the first antibiotics treatment. Be cautious, however, that the
+average over only a few Verrucomicrobiae species will be a more variable
+estimate than the averages over the more prevalent phyla.
 
-```{r verrucomicrobiae, fig.cap = "The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant.
-In particular, they seem to increase after the first antibiotic time course."}
-include_graphics("figure/verrucomicrobiae.png")
-```
+\begin{figure}
 
-### Differential Bacterial Abundance and Preterm Births ###
+{\centering \includegraphics[width=350px]{figure/verrucomicrobiae}
 
-@digiulio2015temporal tracked the abundance of bacteria in
-the vaginal microbiome during pregnancy in an effort to study
-relationships between bacterial community composition and preterm
-birth. Ideally, it would be possible to develop clear bacterial
-signatures associated with preterm births.
+}
+
+\caption{The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant.
+In particular, they seem to increase after the first antibiotic time course.}\label{fig:verrucomicrobiae}
+\end{figure}
+
+\subsubsection*{Differential Bacterial Abundance and Preterm
+Births}\label{differential-bacterial-abundance-and-preterm-births}
+
+\citet{digiulio2015temporal}
+tracked the abundance of bacteria in the vaginal microbiome during
+pregnancy in an effort to study relationships between bacterial
+community composition and preterm birth. Ideally, it would be possible
+to develop clear bacterial signatures associated with preterm births.
 
@@
 could average across all individuals, we will take our cue from
-[@digiulio2015temporal] and place each sample into one of five
-Community State Types (CSTs), identified via k-medoids.  In that
-study, a linear model identified one of these CSTs (CST 4) as
-significantly more diverse, further it appeared associated with
-preterm births. Here, we corroborate this finding using exploratory
-views.
+\citep{digiulio2015temporal}
+place each sample into one of five Community State Types (CSTs),
+identified via k-medoids. In that study, a linear model identified one
+of these CSTs (CST 4) as significantly more diverse, further it appeared
+associated with preterm births. Here, we corroborate this finding using
+exploratory views.
 
 Therefore, our focus here is on the differential abundance question,
-rather than dynamics.  We would like to provide visual representations
-of differential abundance across CSTs and also between preterm and
-non-preterm births. [@digiulio2015temporal] interpreted the CSTs using
-a heatmap, with bacteria ordered according to a hierarchical
-clustering.  By using the DOI sankey instead, we can interpret the
-CSTs in their taxonomic context and at multiple scales of taxonomic
-resolution. Further, while [@digiulio2015temporal] focused on
-identifying associations between preterm births and CSTs -- presumably
-because testing individual bacteria loses power -- we can compare
-bacterial abundances between preterm and non-preterm samples along
-subtrees, without requiring CSTs as an intermediary.
+rather than dynamics. We would like to provide visual representations of
+differential abundance across CSTs and also between preterm and non-preterm
+births. \citet{digiulio2015temporal} interpreted the CSTs using a heatmap, with
+bacteria ordered according to a hierarchical clustering. By using the DOI sankey
+instead, we can interpret the CSTs in their taxonomic context and at multiple
+scales of taxonomic resolution. Further, while 
+\citep{digiulio2015temporal} focused on identifying associations between
+preterm births and CSTs -- presumably because testing individual bacteria loses
+power -- we can compare bacterial abundances between preterm and non-preterm
+samples along subtrees, without requiring CSTs as an intermediary.
 
-In Figure \@ref(fig:pretermcsts), we compare the 5 CSTs according to
-their values along the subtree. Specifically, we took the average of
-all samples within each CST to define values at the (species-level)
-leaf nodes, and then aggregated the averages up to the root. It is
+In Figure \ref{fig:pretermcsts}, we compare the 5 CSTs according to
+their values along the subtree. Specifically, we took the average of all
+samples within each CST to define values at the (species-level) leaf
+nodes, and then aggregated the averages up to the root. It is
 immediately clear that samples from CST 4 have much more taxonomic
@@
 that the differential abundance of these bacteria distinguishes the
-remaining CSTs, see Figure \@ref(fig:pretermcstslacto).
+remaining CSTs, see Figure \ref{fig:pretermcstslacto}.
 
-Alternatively, in Figure \@ref(fig:pretermcsts), we avoid working with
+Alternatively, in Figure \ref{fig:pretermcsts}, we avoid working with
 CSTs, displaying instead averages among samples associated with either
@@
 
-```{r pretermcsts, fig.cap = "The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes."}
-include_graphics("figure/preterm_csts.png")
-```
+\begin{figure}
 
-```{r pretermcstslacto, fig.cap = "Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent."}
-include_graphics("figure/preterm_csts_lacto.png")
-```
+{\centering \includegraphics[width=350px]{figure/preterm_csts}
 
-```{r pretermpreterm, out.width = "300px", fig.cap = "Samples with high levels of phyla other than Firmicutes appear to be related to preterm births."}
-include_graphics("figure/preterm_preterm.png")
-```
+}
 
-### Dynamics in Housing Prices {#zillow-study}
+\caption{The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes.}\label{fig:pretermcsts}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/preterm_csts_lacto}
+
+}
+
+\caption{Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent.}\label{fig:pretermcstslacto}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=300px]{figure/preterm_preterm}
+
+}
+
+\caption{Samples with high levels of phyla other than Firmicutes appear to be related to preterm births.}\label{fig:pretermpreterm}
+\end{figure}
+
+\subsubsection*{Dynamics in Housing Prices}\label{zillow-study}
 
@@
 relatively clear hierarchical structure. Our data are downloaded from
-Zillow^[[http://www.zillow.com/research/data/](http://www.zillow.com/research/data/)],
-and give the Zillow Home Value Indexes at the neighborhood level,
-across the country, computed monthly between 1996 and 2016. In our
-display, we have taken the base 10 log of these indexes. As our
-hierarchical structure, we use each neighborhood's assignment to
-state, regional, county, and city levels. We represent each of these
-coarser spatial categories using the average of all neighborhoods
-contained in them. We have filtered down to the 890 neighborhoods in
-California; rendering more neighborhoods while keeping all 246
-timepoints causes the interface to lag^[See Section
-\@ref(conclusion) for potential optimizations, however.].
+Zillow\footnote{\url{http://www.zillow.com/research/data/}}, and give
+the Zillow Home Value Indexes at the neighborhood level, across the
+country, computed monthly between 1996 and 2016. In our display, we have
+taken the base 10 log of these indexes. As our hierarchical structure,
+we use each neighborhood's assignment to state, regional, county, and
+city levels. We represent each of these coarser spatial categories using
+the average of all neighborhoods contained in them. We have filtered
+down to the 890 neighborhoods in California; rendering more
+neighborhoods while keeping all 246 timepoints causes the interface to
+lag\footnote{See Section \ref{conclusion} for potential optimizations,
+  however.}.
 
-Our basic analysis revolves around geographic and temporal variation
-in home prices. We are especially interested in the effect of the 2008
+Our basic analysis revolves around geographic and temporal variation in
+home prices. We are especially interested in the effect of the 2008
 recession and any variation in the lead-up to or recovery from this
@@
 be highlighted using a single brush in treeboxes, since each spatial
-level is displayed at the same height in the tree. The resulting
-display is given in Figure \@ref(fig:zillowcities). From a high level,
-the housing bubble, decrease in prices due to the recession, and
-subsequent recovery are readily apparent. The city-level view makes it
-clear that not all neighborhoods were equally affected by the
-recession -- richer cities plateaued at their 2008 prices,
-middle-income cities saw moderate decreases, and poorer cities saw the
-most significant declines in home prices. These more significant
-declines tended to occur in cities that had recently seen rapid
-increases during the housing bubble. Finally, we note that the range
-in home prices seems to have increased after the recession, speaking
-to the differential long-term effect of the recession on prices.
+level is displayed at the same height in the tree. The resulting display
+is given in Figure \ref{fig:zillowcities}. From a high level, the
+housing bubble, decrease in prices due to the recession, and subsequent
+recovery are readily apparent. The city-level view makes it clear that
+not all neighborhoods were equally affected by the recession -- richer
+cities plateaued at their 2008 prices, middle-income cities saw moderate
+decreases, and poorer cities saw the most significant declines in home
+prices. These more significant declines tended to occur in cities that
+had recently seen rapid increases during the housing bubble. Finally, we
+note that the range in home prices seems to have increased after the
+recession, speaking to the differential long-term effect of the
+recession on prices.
 
@@
 recession. We generate the sequence of views in Figures
-\@ref(fig:zillowmiddlepre), \@ref(fig:zillowmiddleup), and
-\@ref(fig:zillowmiddledown) to this end. The first of these figures
-isolates neighborhoods with middle incomes before the recession, using
-a single timebox. Since there appears to be a divergence in
-trajectories after the recession, we introduce a second post-recession
-timebox, dragging it over series with higher and lower incomes during
-this second time period. This is the content of Figures
-\@ref(fig:zillowmiddleup) and \@ref(fig:zillowmiddledown). Inspecting
-the highlighted tree nodes associated with these series, we find that
-most of the middle-income series that increased after the recession
-are associated with middle-income neighborhoods within the coastal
-Southern California counties. In constrast, those middle income series
-that saw decreases were mostly located in Central California and
-Oakland.
+\ref{fig:zillowmiddlepre}, \ref{fig:zillowmiddleup}, and
+\ref{fig:zillowmiddledown} to this end. The first of these figures
+isolates neighborhoods with middle incomes before the recession, using a
+single timebox. Since there appears to be a divergence in trajectories
+after the recession, we introduce a second post-recession timebox,
+dragging it over series with higher and lower incomes during this second
+time period. This is the content of Figures \ref{fig:zillowmiddleup} and
+\ref{fig:zillowmiddledown}. Inspecting the highlighted tree nodes
+associated with these series, we find that most of the middle-income
+series that increased after the recession are associated with
+middle-income neighborhoods within the coastal Southern California
+counties. In constrast, those middle income series that saw decreases
+were mostly located in Central California and Oakland.
 
-The previous analysis highlights the fact that, within even narrow geographic
-regions, there can be substantial variation in prices. We can study this
-directly using treeboxes. In Figure \@ref(fig:zillowsf) we have highlighted all
-series in San Francisco County. We see that, in 2016, prices range from around
-$10^{5.6} \approx \$400,000$ to $10^{6.7} \approx \$5 \text{ million}$. So, while
-all these neighborhoods tend to be among the more expensive ones in California,
+The previous analysis highlights the fact that, within even narrow
+geographic regions, there can be substantial variation in prices. We can
+study this directly using treeboxes. In Figure \ref{fig:zillowsf} we
+have highlighted all series in San Francisco County. We see that, in
+2016, prices range from around \(10^{5.6} \approx \$400,000\) to
+\(10^{6.7} \approx \$5 \text{ million}\). So, while all these
+neighborhoods tend to be among the more expensive ones in California,
 prices can vary in a non-smooth way across geographic space.
@@
 
-```{r zillowcities, fig.cap = "California home prices at the city level, between 1996 and 2016.
-The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession."}
-include_graphics("figure/zillow_cities.png")
-```
+\begin{figure}
 
-```{r zillowmiddlepre, fig.cap = "Neighborhoods with mid-range home prices before the recession are selected here.
-Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points."}
-include_graphics("figure/zillow_middle_pre.png")
-```
+{\centering \includegraphics[width=350px]{figure/zillow_cities}
 
-```{r zillowmiddleup, fig.cap = "Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly.
-These appear to be located mainly in Los Angeles and San Diego counties."}
-include_graphics("figure/zillow_middle_up.png")
-```
+}
 
-```{r zillowmiddledown, fig.cap = "Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession.
-These seem mostly to be located in Central California and the East San Francisco Bay Area."}
-include_graphics("figure/zillow_middle_down.png")
-```
+\caption{California home prices at the city level, between 1996 and 2016.
+The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession.}\label{fig:zillowcities}
+\end{figure}
 
-```{r zillowsf, fig.cap = "To study the range in home prices within San Francisco County, we can select the associated nodes using a treebox."}
-include_graphics("figure/zillow_sf.png")
-```
+\begin{figure}
 
-### Sources of Variation in Bikesharing Demand {#bikesharing-study}
+{\centering \includegraphics[width=350px]{figure/zillow_middle_pre}
+
+}
+
+\caption{Neighborhoods with mid-range home prices before the recession are selected here.
+Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points.}\label{fig:zillowmiddlepre}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/zillow_middle_up}
+
+}
+
+\caption{Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly.
+These appear to be located mainly in Los Angeles and San Diego counties.}\label{fig:zillowmiddleup}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/zillow_middle_down}
+
+}
+
+\caption{Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession.
+These seem mostly to be located in Central California and the East San Francisco Bay Area.}\label{fig:zillowmiddledown}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/zillow_sf}
+
+}
+
+\caption{To study the range in home prices within San Francisco County, we can select the associated nodes using a treebox.}\label{fig:zillowsf}
+\end{figure}
+
+\subsubsection*{Sources of Variation in Bikesharing
+Demand}\label{bikesharing-study}
 
 Our next example is a study in bikesharing demand, included as an
-example of analyzing collections of time series when there is no
-obvious hierarchical structure a priori. The data are available at
-the UCI
-Repository^[[https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset](https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset)]
+example of analyzing collections of time series when there is no obvious
+hierarchical structure a priori. The data are available at the UCI
+Repository\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset}}
 and were originally collected by a Washington D.C.-based bikesharing
 system for use in a Kaggle prediction competition. The data are hourly
-measurements of bike demand, aggregated across all bikesharing
-stations, over two years, along with supplemental weather data. In the
+measurements of bike demand, aggregated across all bikesharing stations,
+over two years, along with supplemental weather data. In the
 competition, participants were asked to predict the hourly demand on a
-held-out test set. Here, we adopt a descriptive view instead,
-attempting to characterize factors associated with variation in
-bikesharing demand.
+held-out test set. Here, we adopt a descriptive view instead, attempting
+to characterize factors associated with variation in bikesharing demand.
 
-Like the Zillow home prices application, we study this problem as one
-of describing a large collection of related time series. Here, we
-consider the demand during a single day to be one time series; this is
-a natural choice considering the daily periodicity of bike demand. To
-arrange these daily series along an interpretable tree structure, we
-apply a regression tree relating the supplemental data to the
-bikesharing demand [@breiman1984classification]. In more detail, we
-built this tree by noting the “two table” structure of this problem:
-one describes bike demand, the other holds the supplemental data. In
-both, the rows index days, while the columns index either hours or
-supplemental features. Our tree is the trained regression tree after
-predicting demand at 8AM based on the supplemental data. We choose
-this response because (1) we need a univariate response in order to
-apply regression trees and (2) the more straightforwards reduction to
-daily-average-demand fails to distinguish between weekdays and
-weekends, whose series appear qualitatively very different from each
-other.
+Like the Zillow home prices application, we study this problem as one of
+describing a large collection of related time series. Here, we consider
+the demand during a single day to be one time series; this is a natural
+choice considering the daily periodicity of bike demand. To arrange
+these daily series along an interpretable tree structure, we apply a
+regression tree relating the supplemental data to the bikesharing demand
+\citep{breiman1984classification}. In more
+detail, we built this tree by noting the ``two table'' structure of this
+problem: one describes bike demand, the other holds the supplemental
+data. In both, the rows index days, while the columns index either hours
+or supplemental features. Our tree is the trained regression tree after
+predicting demand at 8AM based on the supplemental data. We choose this
+response because (1) we need a univariate response in order to apply
+regression trees and (2) the more straightforwards reduction to
+daily-average-demand fails to distinguish between weekdays and weekends,
+whose series appear qualitatively very different from each other.
 
@@
 (unsurprisingly) the difference between weekends and weekdays. This is
-emphasized in Figures 14 and 15, respectively; using timeboxes to
-isolate the two types of series highlight the left and right sides of
-the tree, respectively. For a more subtle effect, we select the
-internal nodes associated with the first split below the weekday
-vs. weekend split; these are given in Figures 16 and 17. This suggests
-that weekday demand increased during the second year.
+emphasized in Figures \ref{fig:working} and \ref{fig:weekend}, respectively;
+using timeboxes to isolate the two types of series highlight the left and right
+sides of the tree, respectively. For a more subtle effect, we select the
+internal nodes associated with the first split below the weekday vs. weekend
+split; these are given in Figures \ref{fig:weekday2011} and
+\ref{fig:weekday2012}. This suggests that weekday demand increased during the
+second year.
 
-In contrast to these general questions about daily demand, we could
-ask a more granular question about specific time windows. For example,
-what characterizes days on which there is larger than average demand
-after midnight? We can select these series after first zooming into
-this time window. Figure 18 reveals that the highlighted series are
-associated with the warm-weekend split, which seems quite reasonable
-in retrospect.
+In contrast to these general questions about daily demand, we could ask
+a more granular question about specific time windows. For example, what
+characterizes days on which there is larger than average demand after
+midnight? We can select these series after first zooming into this time window.
+Figure \ref{fig:warmweekend} reveals that the highlighted series are associated
+with the warm-weekend split, which seems quite reasonable in retrospect.
 
@@
 8AM demand and then count the abundance of the groups flowing down
-different branches. We find that the quintiles are each rather
-strongly separated after descending even a few steps down the
-regression tree – for example, Figures \@ref(fig:weekday2011) and
-\@ref(fig:weekday2012) focus on 2011 vs. 2012 split among weekday
-samples, showing that this split distinguishes between samples falling
-in the second and third quintiles of 8AM demand.
+different branches. We find that the quintiles are each rather strongly
+separated after descending even a few steps down the regression tree --
+for example, Figures \ref{fig:weekday2011} and \ref{fig:weekday2012}
+focus on 2011 vs. 2012 split among weekday samples, showing that this
+split distinguishes between samples falling in the second and third
+quintiles of 8AM demand.
 
-This interactive representation of regression trees is potentially
-more useful on larger trees that cannot be easily parsed in a single
-view; in this sense the bikesharing tree is relatively simple. In our
-ideal data analysis workflow, we imagine the analyst applying
-interactive visualization and modeling techniques in an iterative,
-nonlinear fashion.
+This interactive representation of regression trees is potentially more
+useful on larger trees that cannot be easily parsed in a single view; in
+this sense the bikesharing tree is relatively simple. In our ideal data
+analysis workflow, we imagine the analyst applying interactive
+visualization and modeling techniques in an iterative, nonlinear
+fashion.
 
-```{r working, fig.cap = "The two peaks at rush hour distinguish weekday series from the rest."}
-include_graphics("figure/working.png")
-```
+\begin{figure}
 
-```{r weekend, fig.cap = "Unlike weekday demand, weekend demand is unimodal.
+{\centering \includegraphics[width=350px]{figure/working}
+
+}
+
+\caption{The two peaks at rush hour distinguish weekday series from the rest.}\label{fig:working}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/weekend}
+
+}
+
+\caption{Unlike weekday demand, weekend demand is unimodal.
 The few weekday series with unimodal series seem to be associated with holidays.
-This is the case for New Years' Eve, which is currently hovered over in the tree."}
-include_graphics("figure/weekend.png")
-```
+This is the case for New Years' Eve, which is currently hovered over in the tree.}\label{fig:weekend}
+\end{figure}
 
-```{r weekday2011, fig.cap = "Weeday demand appears larger in 2012 than 2011 -- compare with the next figure."}
-include_graphics("figure/weekday_2011.png")
-```
+\begin{figure}
 
-```{r weekday2012, fig.cap = "Weekday demand increased in 2012 -- compare with the previous figure."}
-include_graphics("figure/weekday_2012.png")
-```
+{\centering \includegraphics[width=350px]{figure/weekday_2011}
 
-```{r warmweekend, fig.cap = "The samples with highest night demand tend to fall on warm weekends."}
-include_graphics("figure/warm_weekend.png")
-```
+}
 
-```{r bikesankey, fig.cap = "We can interpret the regression tree using an interactive DOI representation."}
-include_graphics("figure/bike_sankey.png")
-```
+\caption{Weeday demand appears larger in 2012 than 2011 -- compare with the next figure.}\label{fig:weekday2011}
+\end{figure}
 
-### Hierarchically Clustering the Global Patterns Data {#global_patterns}
+\begin{figure}
 
-Each of the timebox tree and treebox examples presented so far have focused on
-data with a clear time component. We note however that these techniques could
-alternatively be applied to high-dimensional data, via the use of parallel
-coordinates [@inselberg1991parallel]. The usual parallel coordinates challenges
-remain, namely selecting scales for and an ordering across the different
-coordinates, but the linking and focus-plus-context ideas can still be employed
-in this setting. Here we provide an implementation of this idea on a dataset
-comparing microbiomes across various ecological environments
-[@caporaso2011global], which is publicly accessible through the phyloseq R
-package [@mcmurdie2013phyloseq].
+{\centering \includegraphics[width=350px]{figure/weekday_2012}
 
-The original Global Patterns data consists of 26 samples across 9 environments
-(for example, freshwater and soil). In each site, there are counts across 19216
-taxa -- to simplify visualization, we filter to the 500 most abundant taxa.
+}
 
-We hierarchically cluster these 26 samples based on the 500 most abundant taxa,
-using complete linkage on the UniFrac distance. Figure \@ref(fig:gptimebox) displays
-the resulting hierarchy together with a parallel coordinates view of the
-$\text{asinh}$ transformed taxa.
+\caption{Weekday demand increased in 2012 -- compare with the previous figure.}\label{fig:weekday2012}
+\end{figure}
 
-```{r gptimebox, out.width = "250px", fig.show = "hold", fig.cap = "An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views."}
-include_graphics(c("figure/gp_cluster1.png", "figure/gp_cluster2.png"))
-```
+\begin{figure}
 
-In Figure \@ref(fig:gptimebox), we compare two subclusters from the hierarchical
-clustering tree, after zooming to a few of the bacteria that distinguish between
-the clusters. Upon revisiting the original data, it becomes clear that the
-samples highlighted on the left come from freshwater samples, while those on the
-right come from soil and skin, and looking up taxonomic groups associated with
-the distinguishing bacteria confirms this. For example, many of the species
-with high abundances in the left figure come from order Oceanospirillales.
+{\centering \includegraphics[width=350px]{figure/warm_weekend}
 
-### Inspecting Confirmatory Analysis {#structssi}
+}
 
-In addition to facilitating exploratory study, treelapse has potential value as
-a device for inspecting confirmatory analysis. We provide an illustration
-extending an example from [@callahan2016bioconductor], which formally tested
-bacterial species for association with age in a sample of mice. The testing
-approach advocated there is particularly well-suited to visualization with
-treelapse, since it sought to detect associations at multiple levels of
-phylogenetic resolution, using statistical tools developed in
-[@yekutieli2008hierarchical; @sankaran2014structssi].
+\caption{The samples with highest night demand tend to fall on warm weekends.}\label{fig:warmweekend}
+\end{figure}
 
-The data of interest in [@callahan2016bioconductor] are bacterial counts
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/bike_sankey}
+
+}
+
+\caption{We can interpret the regression tree using an interactive DOI representation.}\label{fig:bikesankey}
+\end{figure}
+
+\subsubsection*{Hierarchically Clustering the Global Patterns Data}\label{global_patterns}
+
+Each of the timebox tree and treebox examples presented so far have
+focused on data with a clear time component. We note however that these
+techniques could alternatively be applied to high-dimensional data, via
+the use of parallel coordinates \citep{inselberg1991parallel}. The usual
+parallel coordinates challenges remain, namely selecting scales for and an
+ordering across the different coordinates, but the linking and
+focus-plus-context ideas can still be employed in this setting. Here we
+provide an implementation of this idea on a dataset comparing microbiomes across
+various ecological environments \citep{caporaso2011global}, which is publicly
+accessible through the phyloseq R package \citep{mcmurdie2013phyloseq}.
+
+The original Global Patterns data consists of 26 samples across 9
+environments (for example, freshwater and soil). In each site, there are
+counts across 19216 taxa -- to simplify visualization, we filter to the
+500 most abundant taxa.
+
+We hierarchically cluster these 26 samples based on the 500 most
+abundant taxa, using complete linkage on the UniFrac distance. Figure
+\ref{fig:gptimebox} displays the resulting hierarchy together with a
+parallel coordinates view of the \(\text{asinh}\) transformed taxa.
+
+\begin{figure}
+  {
+    \centering
+    \includegraphics[width=225px]{figure/gp_cluster1}
+    \includegraphics[width=225px]{figure/gp_cluster2}
+}
+
+\caption{An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views.}\label{fig:gptimebox}
+\end{figure}
+
+In Figure \ref{fig:gptimebox}, we compare two subclusters from the
+hierarchical clustering tree, after zooming to a few of the bacteria
+that distinguish between the clusters. Upon revisiting the original
+data, it becomes clear that the samples highlighted on the left come
+from freshwater samples, while those on the right come from soil and
+skin, and looking up taxonomic groups associated with the distinguishing
+bacteria confirms this. For example, many of the species with high
+abundances in the left figure come from order Oceanospirillales.
+
+\subsubsection*{Inspecting Confirmatory Analysis}\label{structssi}
+
+In addition to facilitating exploratory study, treelapse has potential
+value as a device for inspecting confirmatory analysis. We provide an
+illustration extending an example from \citep{callahan2016bioconductor}, which
+formally tested bacterial species for association with age in a sample of mice.
+The testing approach advocated there is particularly well-suited to
+visualization with treelapse, since it sought to detect associations at multiple
+levels of phylogenetic resolution, using statistical tools developed in
+\citep{yekutieli2008hierarchical, sankaran2014structssi}.
+
+The data of interest in \citep{callahan2016bioconductor} are bacterial counts
 collected across old and young mice. After variance-stabilizing these counts
-using DESeq2 [@love2014moderated], a $t$-test was applied to each node in a
-phylogenetic tree, comparing abundances between old and young mice. To account
-for multiple testing, we employ the structSSI algorithm
-[@yekutieli2008hierarchical; @sankaran2014structssi] along with methods
-available in the `multtest` package [@pollard2005multiple].
+using DESeq2 \citep{love2014moderated}, a $t$-test was
+applied to each node in a phylogenetic tree, comparing abundances
+between old and young mice. To account for multiple testing, we employ
+the structSSI algorithm \citep{yekutieli2008hierarchical, sankaran2014structssi}
+along with methods available in the \texttt{multtest} package
+\citep{pollard2005multiple}.
 
-To interpret the results, we apply timebox trees. Our goals are to (1) identify
-subtrees with consistently elevated differential abundance across age groups
-and (2) compare alternative multiple testing adjustment procedures. Our approach
-is to display the negative-log raw and adjusted $p$-values for each node, with
-alternative methods compared via parallel coordinates. One view of the resulting
-display is captured in Figure \@ref(fig:structssi). First, we see that
-significant nodes tend to be significant across all methods -- the ordering
-between different series appears stable. Interestingly, the Sidak one-step and
+To interpret the results, we apply timebox trees. Our goals are to (1)
+identify subtrees with consistently elevated differential abundance
+across age groups and (2) compare alternative multiple testing
+adjustment procedures. Our approach is to display the negative-log raw
+and adjusted $p$-values for each node, with alternative methods
+compared via parallel coordinates. One view of the resulting display is
+captured in Figure \ref{fig:structssi}. First, we see that significant
+nodes tend to be significant across all methods -- the ordering between
+different series appears stable. Interestingly, the Sidak one-step and
 structSSI procedures seem to have lower power than the others, including
-conservative FWER-controlling methods, like the Bonferroni procedure. Further,
-in this application, FDR-controlling techniques do not seem to offer notably
-different adjusted $p$-values, relative to those controlling FWER. This suggests
-that, for this problem, bacteria are either strongly associated with age, or not
-associated at all, so that there is little gain from using more sensitive
-procedures.
+conservative FWER-controlling methods, like the Bonferroni procedure.
+Further, in this application, FDR-controlling techniques do not seem to
+offer notably different adjusted $p$-values, relative to those
+controlling FWER. This suggests that, for this problem, bacteria are
+either strongly associated with age, or not associated at all, so that
+there is little gain from using more sensitive procedures.
 
-```{r structssi, fig.cap = "Viewing a tree of $p$-values across different methods highlights two subtrees with strong associations with mouse age, across several testing procedures."}
-include_graphics("figure/structssi.png")
-```
+\begin{figure}
 
-Further, selecting series with strong association between abundance and age, two
-major subtrees are brought to the forefront. Separately querying the the
-taxonomic identities of these bacteria reveals that they are two subgroups of
-Clostridia, which is consistent with the analysis of
-[@callahan2016bioconductor]. More than this specific analysis outcome, this view
-demonstrates that interactive visual inspection of results from confirmatory
-analysis provides deeper insight than the standard practice of printing tables
-of (adjusted or unadjusted) $p$-values: the relationship between significant
-nodes is only clear upon visualization on the tree.
+{\centering \includegraphics[width=350px]{figure/structssi}
 
-## Conclusion and Future Work {#conclusion}
+}
 
-Here, we have reviewed some fundamental principles of data
-visualization and described their implementation in a new treelapse
-package. Further, we have provided examples of the practical
-usefulness of these principles in real-world data analysis situations.
+\caption{Viewing a tree of $p$-values across different methods highlights two subtrees with strong associations with mouse age, across several testing procedures.}\label{fig:structssi}
+\end{figure}
+
+Further, selecting series with strong association between abundance and
+age, two major subtrees are brought to the forefront. Separately
+querying the the taxonomic identities of these bacteria reveals that they are
+two subgroups of Clostridia, which is consistent with the analysis of
+\citep{callahan2016bioconductor}. More than this specific analysis outcome, this
+view demonstrates that interactive visual inspection of results from
+confirmatory analysis provides deeper insight than the standard practice of
+printing tables of (adjusted or unadjusted) $p$-values: the relationship between
+significant nodes is only clear upon visualization on the tree.
+
+\subsection*{Conclusion and Future Work}\label{conclusion}
+
+Here, we have reviewed some fundamental principles of data visualization
+and described their implementation in a new treelapse package. Further,
+we have provided examples of the practical usefulness of these
+principles in real-world data analysis situations.
 
 This package has only developed basic ideas, and there are a number of
-potentially useful extensions worth exploring. For example, we have
-not considered the principle of arrangement in our visualizations
-[@buja1996interactive], though many of our conclusions were based on
-comparing alternative selections of the same display. We could imagine
-faceting our displays across groups to make these types of comparisons
-more accessible. Further, we have only worked with the DOI
-distribution described in [@heer2004doitrees]. It would be interesting
-to define a more statistical notion of interest along nodes, based on
-cognostics, for example [@hafen2013trelliscope, @friedman2002john]. A
-simple extension could be to allow graph layouts instead of trees in
-time and treebox displays, for data that cannot be coerced into a
-hierarchical structure. Finally, if these ideas turn out to be useful
-in practice, it would be valuable to modularize the basic
-visualization layouts and relationships into a library, allowing the
-wider community to construct novel linked, interactive graphics with
-minimal effort.
+potentially useful extensions worth exploring. For example, we have not
+considered the principle of arrangement in our visualizations 
+\citep{buja1996interactive},
+though many of our conclusions were based on comparing alternative
+selections of the same display. We could imagine faceting our displays
+across groups to make these types of comparisons more accessible.
+Further, we have only worked with the DOI distribution described in
+\citep{heer2004doitrees}. It would be interesting to define a more statistical
+notion of interest along nodes, based on cognostics, for example
+\citep{hafen2013trelliscope, friedman2002john}. A simple extension could be to
+allow graph layouts instead of trees in time and
+treebox displays, for data that cannot be coerced into a hierarchical
+structure. Finally, if these ideas turn out to be useful in practice, it
+would be valuable to modularize the basic visualization layouts and
+relationships into a library, allowing the wider community to construct
+novel linked, interactive graphics with minimal effort.
 
 In summary, we have built an easily accessible R package for
-visualization techniques in a very specific methodology problem –
+visualization techniques in a very specific methodology problem --
 analysis of differential abundance and dynamics in hierarchically
-structured data – that appears in a variety of application domains. We
-have leveraged a link between R and D3 [@vaidyanathan2014htmlwidgets]
-to create visualizations during the exploratory phase of data analysis; in this
+structured data -- that appears in a variety of application domains. We
+have leveraged a link between R and D3 
+\citep{vaidyanathan2014htmlwidgets} to create
+visualizations during the exploratory phase of data analysis; in this
 way our work is a departure from the culture of polished, journalistic
-visualizations prioritized by the D3 community. Finally, we have given a series
-of examples to demonstrate how the general visualization techniques of
-focus-plus-context and linked brushing can be practically incorporated into a
-range of practical analysis workflows, from studying the impact of bacteria on
-human health to better allocating units in commuter bikesharing
-systems.
+visualizations prioritized by the D3 community. Finally, we have given a
+series of examples to demonstrate how the general visualization
+techniques of focus-plus-context and linked brushing can be practically
+incorporated into a range of practical analysis workflows, from studying
+the impact of bacteria on human health to better allocating units in
+commuter bikesharing systems.
 
-## References ##
+\bibliographystyle{plain}
+\bibliography{bibliography}
+
+\end{document}

### 2017-03-23 treelapse.Rmd
For the rest, edit in tex

@@
 link-citations: yes
+number-sections: false
 ---

### 2017-03-23 treelapse.tex
Fix some citation appearances

@@
 question of how variation across time series is related to their
-geographic position. A case study to this type of data is given in
-Section \ref{zillow-study}.
+geographic position.
 
@@
 different partitions of the contextual, covariate space relate to observed time
-series. This approach is described in Section \ref{bikesharing-study}.
+series.
 
@@
 reflecting the time series value at that node. For this reason, we call
-the display, Timebox Trees. In the time series panel, we have directly
+the display, timebox trees. In the time series panel, we have directly
 implemented the timeboxes idea. We then link the panels: when a set of
@@
 queries in terms of time series, it defines queries using nodes in the
-tree. For this reason, we call the display Treeboxes. Rather than
+tree. For this reason, we call the display treeboxes. Rather than
 focusing on the intersection of brushes, as in timebox trees, we focus
@@
 scales of taxonomic resolution. Further, while
-\citep{digiulio2015temporal} focused on identifying associations between
+\citet{digiulio2015temporal} focused on identifying associations between
 preterm births and CSTs -- presumably because testing individual bacteria loses
@@
 neighborhoods while keeping all 246 timepoints causes the interface to
-lag\footnote{See Section \ref{conclusion} for potential optimizations,
-  however.}.
+lag\footnote{See below for potential optimizations, however.}.
 
@@
 
-In contrast to these general questions about daily demand, we could ask
-a more granular question about specific time windows. For example, what
-characterizes days on which there is larger than average demand after
-midnight? We can select these series after first zooming into this time window.
-Figure \ref{fig:warmweekend} reveals that the highlighted series are associated
-with the warm-weekend split, which seems quite reasonable in retrospect.
-
-Finally, we can study the behavior of the regression tree itself using
-the DOI sankey. Here, we group samples according to their quintile of
-8AM demand and then count the abundance of the groups flowing down
-different branches. We find that the quintiles are each rather strongly
-separated after descending even a few steps down the regression tree --
-for example, Figures \ref{fig:weekday2011} and \ref{fig:weekday2012}
-focus on 2011 vs. 2012 split among weekday samples, showing that this
-split distinguishes between samples falling in the second and third
-quintiles of 8AM demand.
-
-This interactive representation of regression trees is potentially more
-useful on larger trees that cannot be easily parsed in a single view; in
-this sense the bikesharing tree is relatively simple. In our ideal data
-analysis workflow, we imagine the analyst applying interactive
-visualization and modeling techniques in an iterative, nonlinear
-fashion.
-
 \begin{figure}
@@
 
+In contrast to these general questions about daily demand, we could ask
+a more granular question about specific time windows. For example, what
+characterizes days on which there is larger than average demand after
+midnight? We can select these series after first zooming into this time window.
+Figure \ref{fig:warmweekend} reveals that the highlighted series are associated
+with the warm-weekend split, which seems quite reasonable in retrospect.
+
 \begin{figure}
@@
 
+Finally, we can study the behavior of the regression tree itself using
+the DOI sankey. Here, we group samples according to their quintile of
+8AM demand and then count the abundance of the groups flowing down
+different branches. We find that the quintiles are each rather strongly
+separated after descending even a few steps down the regression tree --
+for example, Figures \ref{fig:weekday2011} and \ref{fig:weekday2012}
+focus on 2011 vs. 2012 split among weekday samples, showing that this
+split distinguishes between samples falling in the second and third
+quintiles of 8AM demand.
+
 \begin{figure}
@@
 \end{figure}
+
+This interactive representation of regression trees is potentially more
+useful on larger trees that cannot be easily parsed in a single view; in
+this sense the bikesharing tree is relatively simple. In our ideal data
+analysis workflow, we imagine the analyst applying interactive
+visualization and modeling techniques in an iterative, nonlinear
+fashion.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 \bibliography{bibliography}

### 2017-03-24 treelapse.tex
Reduce structssi figure size
Update figure legend order (pregnancy)
Also, add weiland fellowship.

@@
   \author{Kris Sankaran\thanks{
-      This research was supported in part by an NIH training grant (5T32GM096982-03).}\hspace{.2cm}\\
+      This research was supported in part by an NIH training grant (5T32GM096982-03) and a Weiland Fellowship.}\hspace{.2cm}\\
     Department of Statistics, Stanford University\\
@@
 
-Our motivating application is to the analysis of microbial time
-series, where an evolutionary tree relating microbes is available a
+Our motivating application is to the analysis of bacterial time
+series, where an evolutionary tree relating bacteria is available a
 priori. However, we have identified common problem types where, if a
@@
 
-Alternatively, in Figure \ref{fig:pretermcsts}, we avoid working with
+Alternatively, in Figure \ref{fig:pretermpreterm}, we avoid working with
 CSTs, displaying instead averages among samples associated with either
-preterm or term births. The green edges are associated with preterm
+preterm or term births. The red edges are associated with preterm
 births -- we see that they contribute more weight to phyla outside the
@@
 
-{\centering \includegraphics[width=350px]{figure/structssi}
+{\centering \includegraphics[width=250px]{figure/structssi}
 

### 2017-07-10 reviewer_response.tex
begin reviewer response doc

\documentclass{article}
\usepackage{natbib}
\usepackage{graphicx}
\input{preamble.tex}

\begin{document}

We would like to thank the two reviewers and the associate editor for insightful
and constructive comments on our work. Below, we describe changes to the paper
text and software based on this feedback.

\subsection{Reviewer 1}

\textit{This small sampling of a large body of work points to the need for a
  cleaner conceptualization and clarifications of the contribution of the work
  presented in this paper relative to prior efforts.}

We have have modified our literature review to begin with a broader discussion
of tree and time series visualization methods, before proceeding to those
methods used more directly in our work. Further, we have added a bullet list
describing more concretely the contributions of this paper within the broader
body of knowledge.

\textit{My main concern regarding the presentation of the work lies in the third
  and fourth views - why are they presented as distinct? It would seem to me
  that all of the described functionality could be configured in that one view.
  In fact, I could imagine including the DOi in the same view as well - why are
  multiple views needed?}

Yes, it would be possible to develop a view that highlights series based on the
intersection of (1) the union of brushes over the trees and (2) the intersection
of brushes over the time series. We imagined that it would be easier for users
to interpret the display by explicitly modularizing the separate visual tasks.
(todo: we should try to experiment with this idea, to see if it is feasible)

\textit{Some detail about the scale of the tool would be useful. How many nodes
  can be included in the tree? How many time series, of what length? Rough
  estimates would help readers assess utility.}

We have added a description in Section ? describing the scales of different
inputs before the interface begins to lag.

\subsection{Reviewer 2}

\end{document}

### 2017-07-10 reviewer_response.tex

@@
 
+\textit{In the case studies, it would be interesting to see how the highlighted
+  nodes change when you shrink or pan the selecting area in the time series
+  plot.}
+
+This is visible online in our video demo. It is also possible to see this effect
+directly by interacting with precompiled versions in our compiled rmarkdown
+vignettes. We have further added a short supplementary note providing static
+figures describing this effect.
+
+\textit{Is it possible to combine the DOI sankey idea with the timebox trees, so
+  that the time series can be conditioned on the categorical variable?}
+
+It would be possible to colo series according to different properties, or
+alternatively faceting different series into separate panels. We have these
+extensions in mind for a future version of the package, though these are not
+possible in the package as it currently stands.
+
+\textit{As the authors mentioned in the conclusion part, faceting is an
+  important feature to be added in the interactive time series plot. Also,
+  zooming in $y$-scale is needed when the lines are very intensive, like the
+  bottom of Figure 1.}
+
+As mentioned above, faceting is a natural next step, but we have not implemented
+it yet. On zooming however, it is already possible to zoom the $y$-axis, using
+the pan-zoom widget.
+
+\textit{In the case studies, when there are two colors in the time series plot
+  (Figures 4, 9, 13, 16), the brown color is almost hidden by the blue. It would
+  be nicer to have the brown highlighted on the top, or add the transparency to
+  the blue.}
+
+It is already possible to customize the transparency of the blue selected lines
+by using the \texttt{ts\$opacity\_selected} argument in the \texttt{style\_opts}
+argument to the R plotting functions. However, we realize this is not the ideal
+approach to displaying the searched series, and since searching is an important
+function, we have revised our implementation and the relevant figures so that
+the searched series are always displayed on top.
+
+\textit{In the clustering example, it doesn't make sense to show the internal
+  nodes. It is better to display only the leaves in the time series plot.}
+
+We actually think that the nodes are useful to plot, because they can be
+interpreted as centroids of the associated subtree, which are a natural object
+of study.
+
+\textit{Page 7, Line 46, ``node sizes reflecting the time series value at that
+  node''. Does that mean the node size will change when the time changes?}
+
+Sorry, what we actually meant is that the node size reflects the
+\textit{average} value across all times. This has been clarified in the text.
+
+\textit{Page 11, lines 28-40. Should have referred to Figure 5.}
+
+We have added a reference to Figure 5 at the point that we start describing the
+series for Verrucomicrobiae.
+
+\textit{Page 13, Line 42. ``The red edges are associated with preterm births''.
+  Actually red represents term in Figure 8. Hence lines 45-47 are not correct.}
+
+We have fixed this in the text to refer to the green and yellow edges.
+
+\textit{Page 15, Line 46. How did you get the numbers of the power, 5.6 and 6.7?
+  In Figure 13 the $y$-axis goes from around 11 to 15, and I can't find any
+  number close to 6.}
+
+We had accidentally changed the base of the log between generating this figure
+and our text. The text and figure now both use base 10 logs.
+
+\textit{Some of the analysis from the figures are not grounded. For example,
+  page 14 lines 45-47 can not be concluded from Figure 9 easily. The analysis on
+  page 15 lines 37-39 are not directly from the plot, because Figures 11 and 12
+  themselves do not tell the county names.}
+
+We realize this is a difficulty of explaining inferences gathered from
+interactive visualizations through a sequence of static figures. In the text, we
+have now noted some of the interactive gestures the user would have to make to
+completely ground these analysis. For this issue of reproducibility, we have
+also made compiled interactive versions of these examples avaialble so that
+users can confirm the analysis for themselves, without having to install
+treelapse or rerun examples.
+
+\subsection{Associate Editor Review}
+
 \end{document}

### 2017-07-10 treelapse.tex
extra space

@@
 
-Our third display is directed at the bacterial dynamics question. Here,
-two panels are arranged one over the other; one displays all time
-series, while the other displays all tree nodes, with node sizes
-reflecting the time series value at that node. For this reason, we call
-the display, timebox trees. In the time series panel, we have directly
-implemented the timeboxes idea. We then link the panels: when a set of
-series is highlighted by the timeboxes, the associated tree nodes are
-also highlighted. For example, timeboxes can be used to focus on a set
-of series with specific shape -- increased abundance after an ecological
-shock, for example -- and identify along what subtrees this pattern is
-present. To further focus on specific elements, a pan-zoom scented
-widget is provided \citep{willett2007scented}. The widget is a miniature version
-of the full time series panel, equipped with a single brush whose extent
-specifies the limits in the main time series panel. As in the DOI trees and
-sankeys, a search bar can be used to highlight those series of interest a
-priori.
+Our third display is directed at the bacterial dynamics question. Here, two
+panels are arranged one over the other; one displays all time series, while the
+other displays all tree nodes, with node sizes reflecting the value at that node
+averaged across all time points. For this reason, we call the display, timebox
+trees. In the time series panel, we have directly implemented the timeboxes
+idea. We then link the panels: when a set of series is highlighted by the
+timeboxes, the associated tree nodes are also highlighted. For example,
+timeboxes can be used to focus on a set of series with specific shape --
+increased abundance after an ecological shock, for example -- and identify along
+what subtrees this pattern is present. To further focus on specific elements, a
+pan-zoom scented widget is provided \citep{willett2007scented}. The widget is a
+miniature version of the full time series panel, equipped with a single brush
+whose extent specifies the limits in the main time series panel. As in the DOI
+trees and sankeys, a search bar can be used to highlight those series of
+interest a priori.
 
@@
 
-Further, note that in this subtree averages view, the tree display has
-changed. This is because, at each branching point, we place the node
-with larger average value on the left. This places the Verrucomicrobiae
-at the far left, which seem to have large abundance over many time
-points. This phylum had been previously obfuscated -- because there are
-not many leaves associated with this phylum, the sum was small.
-Interestingly, the abundance of these bacteria seems to \emph{increase}
-after the first antibiotics treatment. Be cautious, however, that the
-average over only a few Verrucomicrobiae species will be a more variable
-estimate than the averages over the more prevalent phyla.
+Further, note that in this subtree averages view, the tree display has changed.
+This is because, at each branching point, we place the node with larger average
+value on the left. Figure \ref{fig:fig:verrucomicrobiae} notes that the nodes at
+the far left in the tree are associated with phylum Verrucomicrobiae,
+corresponding to a large average abundance across time points. This phylum had
+been previously obfuscated -- because there are not many leaves associated with
+this phylum, the sum was small. Interestingly, the abundance of these bacteria
+seems to \emph{increase} after the first antibiotics treatment. Be cautious,
+however, that the average over only a few Verrucomicrobiae species will be a
+more variable estimate than the averages over the more prevalent phyla.
 
@@
 
-\caption{The subtree averages aggregation brings attention to the Verrucomicrobiae, which though only present as a few species, are each rather abundant.
-In particular, they seem to increase after the first antibiotic time course.}\label{fig:verrucomicrobiae}
+\caption{The subtree averages aggregation brings attention to the
+  Verrucomicrobiae, which though only present as a few species, are each rather
+  abundant. In particular, they seem to increase after the first antibiotic time
+  course.}\label{fig:verrucomicrobiae}
 \end{figure}
@@
 
-Alternatively, in Figure \ref{fig:pretermpreterm}, we avoid working with
-CSTs, displaying instead averages among samples associated with either
-preterm or term births. The red edges are associated with preterm
-births -- we see that they contribute more weight to phyla outside the
-Firmicutes. This is consistent with the claim that CST 4, the most
-diverse of the CSTs, is associated with preterm births.
+Alternatively, in Figure \ref{fig:pretermpreterm}, we avoid working with CSTs,
+displaying instead averages among samples associated with either preterm or term
+births. The green and yellow edges are associated with preterm births -- we see
+that they contribute more weight to phyla outside the Firmicutes. This is
+consistent with the claim that CST 4, the most diverse of the CSTs, is
+associated with preterm births.
 

### 2017-07-11 reviewer_response.tex
Describe issues with combined figures

@@
 
-Yes, it would be possible to develop a view that highlights series based on the
-intersection of (1) the union of brushes over the trees and (2) the intersection
-of brushes over the time series. We imagined that it would be easier for users
-to interpret the display by explicitly modularizing the separate visual tasks.
-(todo: we should try to experiment with this idea, to see if it is feasible)
+The idea of combining these displays is interesting, and we have attempted to
+implement a version which would highlight the series / nodes that lay in both
+(1) the union of all treeboxes and (2) the intersection of all timeboxes.
+However, we found this idea difficult to implement, for technical rather than
+conceptual reasons -- when attempting to have both brush interactions for
+querying and mouseover interactions for providing detail, the events conflict.
+Indeed, we implemented the first and third recommendations described in
+\href{https://github.com/d3/d3/issues/1604}{this d3 issue}, but in both cases we
+found it necessary to introduce a lag between mousedown and the start of
+brushing in order for the event to properly register. This lag created a drag on
+responsiveness, and the value gained by unifying the displays did not seem worth
+the cost of less fluid interactivity.
 

### 2017-07-11 treelapse.tex
fix the section labeling

@@
 
-\subsection*{Introduction}\label{introduction}
+\section{Introduction}\label{introduction}
 
@@
 
-\subsubsection*{Problem Motivation}\label{problem-motivation}
+\subsection{Problem Motivation}\label{problem-motivation}
 
@@
 
-\subsubsection*{Problem Abstraction}\label{problem-abstraction}
+\subsection{Problem Abstraction}\label{problem-abstraction}
 
@@
 
-\subsection*{Background Literature and Solution
+\subsection{Background Literature and Solution
 Principles}\label{background-literature-and-solution-principles}
@@
 
-\subsection*{Specific Proposals}\label{specific-proposals}
+\subsection{Specific Proposals}\label{specific-proposals}
 
@@
 
-\subsection*{Case Studies}\label{case-studies}
+\section{Case Studies}\label{case-studies}
 
@@
 
-\subsubsection*{Bacterial Dynamics of Antibiotics Time
+\subsection{Bacterial Dynamics of Antibiotics Time
 Courses}\label{bacterial-dynamics-of-antibiotics-time-courses}
@@
 
-\subsubsection*{Differential Bacterial Abundance and Preterm
+\subsection{Differential Bacterial Abundance and Preterm
 Births}\label{differential-bacterial-abundance-and-preterm-births}
@@
 
-\subsubsection*{Dynamics in Housing Prices}\label{zillow-study}
+\subsection{Dynamics in Housing Prices}\label{zillow-study}
 
@@
 
-\subsubsection*{Sources of Variation in Bikesharing
+\subsection{Sources of Variation in Bikesharing
 Demand}\label{bikesharing-study}
@@
 
-\subsubsection*{Hierarchically Clustering the Global Patterns Data}\label{global_patterns}
+\subsection{Hierarchically Clustering the Global Patterns Data}\label{global_patterns}
 
@@
 
-\subsubsection*{Inspecting Confirmatory Analysis}\label{structssi}
+\subsection{Inspecting Confirmatory Analysis}\label{structssi}
 
@@
 
-\subsection*{Conclusion and Future Work}\label{conclusion}
+\section{Conclusion and Future Work}\label{conclusion}
 

### 2017-07-11 treelapse.tex
add description of scalability
(also, changed line lengths for a few sections)

@@
 We introduce methods for visualization of data structured along trees,
-especially hierarchically structured collections of time series. We hope
-both to characterize generically useful techniques for interactively
-visualizing hierarchical data and to offer practical tools for
-implementing such displays. To this end, we identify questions that
-often emerge when working with hierarchical data and provide an R
-package to simplify their investigation \citep{ihaka1996r}.
+especially hierarchically structured collections of time series. We hope both to
+characterize generically useful techniques for interactively visualizing
+hierarchical data and to offer practical tools for implementing such displays.
+To this end, we identify questions that often emerge when working with
+hierarchical data and provide an R package to simplify their investigation
+\citep{ihaka1996r}.
 
-Our key contribution is the adaptation of the visualization principles
-of focus-plus-context and linking to the study of tree-structured data
-\citep{buja1996interactive, becker1987brushing}. Our motivating
-application is to the analysis of bacterial time series, where an
-evolutionary tree relating bacteria is available a priori. However, we
-have identified common problem types where, if a tree is not directly
-available, it can be constructed from data and then studied using our
-techniques.
+Our key contribution is the adaptation of the visualization principles of
+focus-plus-context and linking to the study of tree-structured data
+\citep{buja1996interactive, becker1987brushing}. Our motivating application is
+to the analysis of bacterial time series, where an evolutionary tree relating
+bacteria is available a priori. However, we have identified common problem types
+where, if a tree is not directly available, it can be constructed from data and
+then studied using our techniques.
 
-We have implemented our visualizations in D3, but encapsulated in an R
-package, called treelapse, to facilitate rapid turnover from data
-preparation and modeling to interactive exploration, and vice versa. Our
-code is open-source, and available at
-\url{https://github.com/krisrs1128/treelapse}\footnote{Link removed in blinded manuscript,
-but available to editor.}. We hope this package encourages data analysts
-to work at the border between data modeling and visualization, and more
-generally empowers a wider audience to apply less widely known, but
-powerful, visualization ideas.
+We have implemented our visualizations in D3, but encapsulated in an R package,
+called treelapse, to facilitate rapid turnover from data preparation and
+modeling to interactive exploration, and vice versa. Our code is open-source,
+and available at \url{https://github.com/krisrs1128/treelapse}. We hope this
+package encourages data analysts to work at the border between data modeling and
+visualization, and more generally empowers a wider audience to apply less widely
+known, but powerful, visualization ideas.
 
-The paper is organized as follows. First, we describe our motivating
-application to the microbiome and the associated generic analysis tasks.
-Next, we review the underlying visualization principles behind our
-contributions. Then we then connect these principles to analysis tasks
-we identified earlier, describing in detail the visualization methods we
-have implemented in treelapse. We close with several case studies using
-publicly available data across both microbiome and non-microbiome
-related applications.
+The paper is organized as follows. First, we describe our motivating application
+to the microbiome and the associated generic analysis tasks. Next, we review the
+underlying visualization principles behind our contributions. Then we then
+connect these principles to analysis tasks we identified earlier, describing in
+detail the visualization methods we have implemented in treelapse. We close with
+several case studies using publicly available data across both microbiome and
+non-microbiome related applications.
 
@@
 
-To unify the tree-structured differential abundance and bacterial
-dynamics problems, we identify the data with a collection of random
-variables indexed by nodes in a prespecified tree structure. In the
-differential abundance problem, each random variable lives in
-\(\mathbb{R}^{G}\) where \(G\) is the number of groups being compared.
-Each coordinate represents the abundance for that group, and a node
-exhibits differential abundance when the coordinates are drawn from
-different distributions. On the other hand, in the bacterial dynamics
-problem, each random variable is a time series, living in
-\(\mathbb{R}^{T}\).
+To unify the tree-structured differential abundance and bacterial dynamics
+problems, we identify the data with a collection of random variables indexed by
+nodes in a prespecified tree structure. In the differential abundance problem,
+each random variable lives in \(\mathbb{R}^{G}\) where \(G\) is the number of
+groups being compared. Each coordinate represents the abundance for that group,
+and a node exhibits differential abundance when the coordinates are drawn from
+different distributions. On the other hand, in the bacterial dynamics problem,
+each random variable is a time series, living in \(\mathbb{R}^{T}\).
 
-In both of these applications, we constrain the values of parent nodes
-according to the value of the children nodes: we define the value at
-each node to be either the sum or average of all descendant tips.
-However, it is possible to imagine situations where the internal nodes
-are drawn from their own distribution, unconstrained by descendants. In
-general, analysis in this abstraction focuses on describing the
-distribution of these random variables as a function of their position
-across subsets of the tree. The essential difficulty in these problems
-is high-dimensionality -- there are many tree nodes, each holding a
-vector-valued random variable. Even simply navigating across the tree
-and comparing coordinates in the observed variables is a challenge;
-ideally we could construct a succinct representation of the essential
+In both of these applications, we constrain the values of parent nodes according
+to the value of the children nodes: we define the value at each node to be
+either the sum or average of all descendant tips. However, it is possible to
+imagine situations where the internal nodes are drawn from their own
+distribution, unconstrained by descendants. In general, analysis in this
+abstraction focuses on describing the distribution of these random variables as
+a function of their position across subsets of the tree. The essential
+difficulty in these problems is high-dimensionality -- there are many tree
+nodes, each holding a vector-valued random variable. Even simply navigating
+across the tree and comparing coordinates in the observed variables is a
+challenge; ideally we could construct a succinct representation of the essential
 covariation across subtrees and coordinates.
 
-This framework suggests other potential application areas, not all of
-which have a priori known tree structures. For example, collections of
-spatially-indexed time series are frequently encountered in practice --
-consider energy consumption, product sales, or high school dropout rates
-across regional districts. This type of data has an implicit tree
-structure -- at the top level are different states, while at the bottom
-are individual census tracts, say. Analysis here revolves around the
-question of how variation across time series is related to their
-geographic position.
+This framework suggests other potential application areas, not all of which have
+a priori known tree structures. For example, collections of spatially-indexed
+time series are frequently encountered in practice -- consider energy
+consumption, product sales, or high school dropout rates across regional
+districts. This type of data has an implicit tree structure -- at the top level
+are different states, while at the bottom are individual census tracts, say.
+Analysis here revolves around the question of how variation across time series
+is related to their geographic position.
 
-Alternatively, if this type of hierarchical contextual information is
-not directly available, a tree structure can be learned from the data.
-This could be achieved by learning a hierarchical clustering on the
-original series. Further, if contextual information is available, but it
-is not hierarchical, it is possible to setup a supervised problem that
-uses context to predict features of the time series. We can construct a tree by
-applying a tree-based classifier \citep{breiman1984classification} or extracting
-a regression tree from a more complex supervised model
-\citep{boz2002extracting,saito2002extracting}. Analysis then focuses on how
-different partitions of the contextual, covariate space relate to observed time
-series.
+Alternatively, if this type of hierarchical contextual information is not
+directly available, a tree structure can be learned from the data. This could be
+achieved by learning a hierarchical clustering on the original series. Further,
+if contextual information is available, but it is not hierarchical, it is
+possible to setup a supervised problem that uses context to predict features of
+the time series. We can construct a tree by applying a tree-based classifier
+\citep{breiman1984classification} or extracting a regression tree from a more
+complex supervised model \citep{boz2002extracting,saito2002extracting}. Analysis
+then focuses on how different partitions of the contextual, covariate space
+relate to observed time series.
 
@@
 
-Now that we have specified the essential questions of interest, we
-survey some ideas from the visualization literature that can be applied
-to answer them. As the core difficulty is high-dimensionality, so it
-should be no surprise that the techniques we adapt come from the
-literature on high-dimensional data visualization, namely,
-focus-plus-context and linking.
+Now that we have specified the essential questions of interest, we survey some
+ideas from the visualization literature that can be applied to answer them. As
+the core difficulty is high-dimensionality, so it should be no surprise that the
+techniques we adapt come from the literature on high-dimensional data
+visualization, namely, focus-plus-context and linking.
 
-The focus-plus-context principle is that large collections of visual
-elements can be studied at multiple scales, by simultaneously focusing"
-on a few elements of interest and maintaining the ``context'' of the
-coarser-scale background. A simple example of this idea is to include a
-search box that highlights matching samples (focus) and mutes the rest
-(context). Two more sophisticated methods anchored in this idea are
-timeboxes and Degree-of-Interest (DOI) trees; both are central to the proposals
-in treelapse \citep{hochheiser2004dynamic, heer2004doitrees}. In timeboxes, a
-collection of time series are graphically queried using interactive brushes.
-Series that pass through all of the user-specified brushes are highlighted, and
-the rest are faded to the background. Hence, time series meeting the constraints
-imposed by the brushes are focused, while the remainder are de-emphasized,
-though they remain present as context. This method can be interpreted
-programmatically as the visual analog of a database query, or probabilistically
-as the conditional distribution for the full series, given it passes through
-certain bounds.
+The focus-plus-context principle is that large collections of visual elements
+can be studied at multiple scales, by simultaneously focusing" on a few elements
+of interest and maintaining the ``context'' of the coarser-scale background. A
+simple example of this idea is to include a search box that highlights matching
+samples (focus) and mutes the rest (context). Two more sophisticated methods
+anchored in this idea are timeboxes and Degree-of-Interest (DOI) trees; both are
+central to the proposals in treelapse \citep{hochheiser2004dynamic,
+  heer2004doitrees}. In timeboxes, a collection of time series are graphically
+queried using interactive brushes. Series that pass through all of the
+user-specified brushes are highlighted, and the rest are faded to the
+background. Hence, time series meeting the constraints imposed by the brushes
+are focused, while the remainder are de-emphasized, though they remain present
+as context. This method can be interpreted programmatically as the visual analog
+of a database query, or probabilistically as the conditional distribution for
+the full series, given it passes through certain bounds.
 
-In DOI trees, the viewer's attention is focused on a collection of
-high-interest nodes, while the remaining lower-interest nodes are left
-on the fringes as context. The implementation is modularized into two
-tasks -- the determination of a DOI distribution over nodes in the tree
-and visual layout of a tree given DOI assignments. The DOI distribution
-used in \citep{heer2004doitrees}
-places maximal interest on the node that the user had most recently
-clicked, along with all ancestors. The DOI for all other nodes is
-defined as the distance to the closest maximal interest node. The layout
-step then trims low-interest subtrees until the remaining nodes fit
-within a given screen size. By adjusting the minimal DOI below which
-nodes are hidden, the user can transition between node-specific and
-full-tree scales.
+In DOI trees, the viewer's attention is focused on a collection of high-interest
+nodes, while the remaining lower-interest nodes are left on the fringes as
+context. The implementation is modularized into two tasks -- the determination
+of a DOI distribution over nodes in the tree and visual layout of a tree given
+DOI assignments. The DOI distribution used in \citep{heer2004doitrees} places
+maximal interest on the node that the user had most recently clicked, along with
+all ancestors. The DOI for all other nodes is defined as the distance to the
+closest maximal interest node. The layout step then trims low-interest subtrees
+until the remaining nodes fit within a given screen size. By adjusting the
+minimal DOI below which nodes are hidden, the user can transition between
+node-specific and full-tree scales.
 
@@
 side-by-side in order to display covariation across views. A canonical
-application is to linked scatterplot brushing
-\citep{becker1987brushing}. Here, a scatterplot
-matrix gives the relationship between all pairs of variables. ~Points
-brushed in one scatterplot are then highlighted in all others. For
-example, this helps the user determine whether an outlier in one
-dimension is an outlier in others. Another instance of this idea links
-the results of dimensionality reduction methods to displays of the raw
-data, as implemented by XGobi and Cranvas, for example
-\citep{xie2013cranvas, swayne1998xgobi}. As in timeboxes,
-linking can be interpreted as database queries or conditional
-probabilities: given a subset of the series after conditioning on the
-values for one set of features, what are the values for a second set
+application is to linked scatterplot brushing \citep{becker1987brushing}. Here,
+a scatterplot matrix gives the relationship between all pairs of variables.
+~Points brushed in one scatterplot are then highlighted in all others. For
+example, this helps the user determine whether an outlier in one dimension is an
+outlier in others. Another instance of this idea links the results of
+dimensionality reduction methods to displays of the raw data, as implemented by
+XGobi and Cranvas, for example \citep{xie2013cranvas, swayne1998xgobi}. As in
+timeboxes, linking can be interpreted as database queries or conditional
+probabilities: given a subset of the series after conditioning on the values for
+one set of features, what are the values for a second set
 \citep{buja1996interactive}?
 
-Finally, unrelated to established visualization principles, we note that
-our work is deliberately grounded in the R software ecosystem. This
-connection is made using the htmlwidgets package
-\citep{vaidyanathan2014htmlwidgets}. Not only
-does linking R with D3 make these visualization methods more broadly
-accessible, we hope to facilitate exchange between data modeling and
-interactive visualization. Moreover, our tools are intentionally limited
-in scope -- designed to facilitate this dialog for a specific class of
-problems, rather than providing a toolbox for generic types of
-visualization design. We believe that this narrow context within a broad
-ecosystem strikes a balance between problem-specificity and ease-of-use.
+Finally, unrelated to established visualization principles, we note that our
+work is deliberately grounded in the R software ecosystem. This connection is
+made using the htmlwidgets package \citep{vaidyanathan2014htmlwidgets}. Not only
+does linking R with D3 make these visualization methods more broadly accessible,
+we hope to facilitate exchange between data modeling and interactive
+visualization. Moreover, our tools are intentionally limited in scope --
+designed to facilitate this dialog for a specific class of problems, rather than
+providing a toolbox for generic types of visualization design. We believe that
+this narrow context within a broad ecosystem strikes a balance between
+problem-specificity and ease-of-use.
 
@@
 
-Our first proposed visualization technique is a minor modification of
-the DOI tree. The standard DOI tree definition does not have any notion
-of data defined at nodes, it is only used a device for navigating tree
-structures. A trivial extension can encode scalar data at nodes: have
-the node radius reflect the associated scalar value. To reinforce this
-effect, we can adjust the width of the parent edge. When parent nodes
-have values equal to the sum of their children, this creates the effect
-of values ``flowing'' from the root to leaves. To help viewers make use
-of their domain knowledge, we have included a search box that highlights
-paths to nodes with matching terms. Edges are ordered from widest on the
-left to narrowest on the right. While this method can only represent a
-single scalar-value per tree node, it suggests an approach to the
-tree-structured differential abundance problem, which we call the DOI
-sankey.
+Our first proposed visualization technique is a minor modification of the DOI
+tree. The standard DOI tree definition does not have any notion of data defined
+at nodes, it is only used a device for navigating tree structures. A trivial
+extension can encode scalar data at nodes: have the node radius reflect the
+associated scalar value. To reinforce this effect, we can adjust the width of
+the parent edge. When parent nodes have values equal to the sum of their
+children, this creates the effect of values ``flowing'' from the root to leaves.
+To help viewers make use of their domain knowledge, we have included a search
+box that highlights paths to nodes with matching terms. Edges are ordered from
+widest on the left to narrowest on the right. While this method can only
+represent a single scalar-value per tree node, it suggests an approach to the
+tree-structured differential abundance problem, which we call the DOI sankey.
 
-In the DOI sankey, we split each edge in the DOI Tree across different
-groups. For example, suppose we have the average counts for treatment
-and control groups at each tree node. Every edge in the tree is split
-into two colors\footnote{We use the colorbrewer palette to facilitate
-  readability \citep{brewer2003colorbrewer}}, with relative
-widths of the different colors reflecting differences in sizes for the
-two groups. The overall width of each edge represents the sum of values
-across all groups.
+In the DOI sankey, we split each edge in the DOI Tree across different groups.
+For example, suppose we have the average counts for treatment and control groups
+at each tree node. Every edge in the tree is split into two colors\footnote{We
+  use the colorbrewer palette to facilitate readability
+  \citep{brewer2003colorbrewer}}, with relative widths of the different colors
+reflecting differences in sizes for the two groups. The overall width of each
+edge represents the sum of values across all groups.
 
-This display is designed to facilitate investigation of the tree
-structured differential abundance question. For example, for a single
-node and a single group, first compute the average abundance at that
-node among all samples in that group. This will give the width for that
-group's color on the edge leading to the specified node. Differentially
-abundant subtrees then correspond to subtrees where some colors occupy
-more space than others. That is, this representation makes it easier to
-identify points where the ``flows'' for different groups diverge -- the
-colors begin to separate. The DOI principle assists navigation across
-the tree structure, allowing focus on individual flow structures without
-losing broader tree context.
+This display is designed to facilitate investigation of the tree structured
+differential abundance question. For example, for a single node and a single
+group, first compute the average abundance at that node among all samples in
+that group. This will give the width for that group's color on the edge leading
+to the specified node. Differentially abundant subtrees then correspond to
+subtrees where some colors occupy more space than others. That is, this
+representation makes it easier to identify points where the ``flows'' for
+different groups diverge -- the colors begin to separate. The DOI principle
+assists navigation across the tree structure, allowing focus on individual flow
+structures without losing broader tree context.
 
@@
 
-The final display currently implemented in the package is the natural
-converse of the timebox trees display. Rather than defining visual
-queries in terms of time series, it defines queries using nodes in the
-tree. For this reason, we call the display treeboxes. Rather than
-focusing on the intersection of brushes, as in timebox trees, we focus
-on the union of brushed over nodes. This allows us to highlight series
-associated with nodes on distant subtrees. This display is also suited
-for the bacterial dynamics problem. For example, by highlighting all
-nodes at one taxonomic level in the tree, we can easily summarize the
-time series pattern for all the taxa at that level. Alternatively,
-focusing on all the children below a single node makes it possible to
-see how much correlation and competition there is between taxonomically
-similar bacteria. As in the timebox trees display, a search box and
-pan-zoom scented widget are provided.
+The final display currently implemented in the package is the natural converse
+of the timebox trees display. Rather than defining visual queries in terms of
+time series, it defines queries using nodes in the tree. For this reason, we
+call the display treeboxes. Rather than focusing on the intersection of brushes,
+as in timebox trees, we focus on the union of brushed over nodes. This allows us
+to highlight series associated with nodes on distant subtrees. This display is
+also suited for the bacterial dynamics problem. For example, by highlighting all
+nodes at one taxonomic level in the tree, we can easily summarize the time
+series pattern for all the taxa at that level. Alternatively, focusing on all
+the children below a single node makes it possible to see how much correlation
+and competition there is between taxonomically similar bacteria. As in the
+timebox trees display, a search box and pan-zoom scented widget are provided.
+
+To be practically useful, the resulting visualizations must respond fluidly to
+user interaction. As the data increase in scale, this fluidity can deteriorate
+for two reasons. First, rendering many SVG elements in a framework like D3 is
+costly\citep{johnson2008scalability}. While it is possible to use alternatives
+-- HTML5's Canvas, for example -- it is often more challenging to implement
+complex interactive behavior through them. Second, some of the dynamic queries
+require a search over a many elements. These limitations are most pronounced in
+the timebox tree display, which must search through all timepoints among all
+time series whenever the brush is moved. The first, but not the second, concern
+applies to treeboxes, while neither applies to DOI trees. Nonetheless, we feel
+comfortable recommending timebox trees for data on the order of 500 tree tips
+and 50 timepoints We note the sizes of the data sets in each case study below,
+which each render fluidly, with the possible exception of the California housing
+data, where there is a noticeable lag in the timebox tree and treebox displays.
 
@@
 We now delve into applications on real data. Our goals are to illustrate
-potential workflows that incorporate treelapse, describe the formulation
-of questions that can be naturally investigated with our methods, and
-provide example interpretations on treelapse output. Our examples are
-also chosen to reflect the range of problem domains to which the package
-can be applied -- though it was motivated by applications to the
-microbiome, it is not tied to it. More importantly, we argue that the
-visualization principles reviewed above can substantively improve the
-practice of data analysis in the class of problems to which we have
-limited ourselves.
+potential workflows that incorporate treelapse, describe the formulation of
+questions that can be naturally investigated with our methods, and provide
+example interpretations on treelapse output. Our examples are also chosen to
+reflect the range of problem domains to which the package can be applied --
+though it was motivated by applications to the microbiome, it is not tied to it.
+More importantly, we argue that the visualization principles reviewed above can
+substantively improve the practice of data analysis in the class of problems to
+which we have limited ourselves.
 
@@
 
-\citet{dethlefsen2008pervasive} investigated the
-effect of antibiotics on bacterial community composition from an
-ecological perspective. The study tracks the microbiome of three
-patients across ten months, with two five-day antibiotic time courses
-separated by 6 months. Discerning the variation in resilience across
-bacteria is important, considering the the role of bacteria in health
-and not just disease.
+\citet{dethlefsen2008pervasive} investigated the effect of antibiotics on
+bacterial community composition from an ecological perspective. The study tracks
+the microbiome of three patients across ten months, with two five-day antibiotic
+time courses separated by 6 months. Discerning the variation in resilience
+across bacteria is important, considering the the role of bacteria in health and
+not just disease.
 
-We approach the data using the linked time and treebox views, after
-first filtering low variance taxa and taking an \(\text{asinh}\)
-transformation. An initial view, Figure \ref{fig:antibioticoverview},
-reveals two dramatic drops in the overall bacterial abundance time
-series during the antibiotics time courses. Two more subtle effects are
-also suggested from this view,
+We approach the data using the linked time and treebox views, after first
+filtering low variance taxa and taking an \(\text{asinh}\) transformation. An
+initial view, Figure \ref{fig:antibioticoverview}, reveals two dramatic drops in
+the overall bacterial abundance time series during the antibiotics time courses.
+Two more subtle effects are also suggested from this view,
 
@@
 Next, using the scented widget, we focus on the window around the second
-antibiotic treatment. We apply the treebox display to compare then
-behavior of different families of Firmicutes, Lachnospiraceae and
-Ruminoccocus. We suspect that these taxa are associated with the delayed
-recovery after the second time course. To investigate this, we input
-these family names in the search box to isolate their positions on the
-tree; then we apply brushes to highlight the series that contribute to
-these higher-level families. The resulting view is given by Figure
-\ref{fig:antibioticfirmicutes}
+antibiotic treatment. We apply the treebox display to compare then behavior of
+different families of Firmicutes, Lachnospiraceae and Ruminoccocus. We suspect
+that these taxa are associated with the delayed recovery after the second time
+course. To investigate this, we input these family names in the search box to
+isolate their positions on the tree; then we apply brushes to highlight the
+series that contribute to these higher-level families. The resulting view is
+given by Figure \ref{fig:antibioticfirmicutes}
 
 \begin{figure}
-
 {\centering \includegraphics[width=350px]{figure/antibiotic_firmicutes}
@@
 
-We next consider an application unrelated to the microbiome, but with
-relatively clear hierarchical structure. Our data are downloaded from
-Zillow\footnote{\url{http://www.zillow.com/research/data/}}, and give
-the Zillow Home Value Indexes at the neighborhood level, across the
-country, computed monthly between 1996 and 2016. In our display, we have
-taken the base 10 log of these indexes. As our hierarchical structure,
-we use each neighborhood's assignment to state, regional, county, and
-city levels. We represent each of these coarser spatial categories using
-the average of all neighborhoods contained in them. We have filtered
-down to the 890 neighborhoods in California; rendering more
-neighborhoods while keeping all 246 timepoints causes the interface to
-lag\footnote{See below for potential optimizations, however.}.
+We next consider an application unrelated to the microbiome, but with relatively
+clear hierarchical structure. Our data are downloaded from
+Zillow\footnote{\url{http://www.zillow.com/research/data/}}, and give the Zillow
+Home Value Indexes at the neighborhood level, across the country, computed
+monthly between 1996 and 2016. In our display, we have taken the base 10 log of
+these indexes. As our hierarchical structure, we use each neighborhood's
+assignment to state, regional, county, and city levels. We represent each of
+these coarser spatial categories using the average of all neighborhoods
+contained in them. We have filtered down to the 890 neighborhoods in California;
+rendering more neighborhoods while keeping all 246 timepoints causes the
+interface to lag\footnote{See below for potential optimizations, however.}.
 
-Our basic analysis revolves around geographic and temporal variation in
-home prices. We are especially interested in the effect of the 2008
-recession and any variation in the lead-up to or recovery from this
-event. These questions can be naturally framed using timebox trees and
-treeboxes.
+Our basic analysis revolves around geographic and temporal variation in home
+prices. We are especially interested in the effect of the 2008 recession and any
+variation in the lead-up to or recovery from this event. These questions can be
+naturally framed using timebox trees and treeboxes.
 
-We first study variation in prices at the city level. These series can
-be highlighted using a single brush in treeboxes, since each spatial
-level is displayed at the same height in the tree. The resulting display
-is given in Figure \ref{fig:zillowcities}. From a high level, the
-housing bubble, decrease in prices due to the recession, and subsequent
-recovery are readily apparent. The city-level view makes it clear that
-not all neighborhoods were equally affected by the recession -- richer
-cities plateaued at their 2008 prices, middle-income cities saw moderate
-decreases, and poorer cities saw the most significant declines in home
-prices. These more significant declines tended to occur in cities that
-had recently seen rapid increases during the housing bubble. Finally, we
-note that the range in home prices seems to have increased after the
-recession, speaking to the differential long-term effect of the
+We first study variation in prices at the city level. These series can be
+highlighted using a single brush in treeboxes, since each spatial level is
+displayed at the same height in the tree. The resulting display is given in
+Figure \ref{fig:zillowcities}. From a high level, the housing bubble, decrease
+in prices due to the recession, and subsequent recovery are readily apparent.
+The city-level view makes it clear that not all neighborhoods were equally
+affected by the recession -- richer cities plateaued at their 2008 prices,
+middle-income cities saw moderate decreases, and poorer cities saw the most
+significant declines in home prices. These more significant declines tended to
+occur in cities that had recently seen rapid increases during the housing
+bubble. Finally, we note that the range in home prices seems to have increased
+after the recession, speaking to the differential long-term effect of the
 recession on prices.
 
-Alternatively, we can study the trajectories of home prices among
-neighborhoods, conditional on their being middle-income before the
-recession. We generate the sequence of views in Figures
-\ref{fig:zillowmiddlepre}, \ref{fig:zillowmiddleup}, and
-\ref{fig:zillowmiddledown} to this end. The first of these figures
-isolates neighborhoods with middle incomes before the recession, using a
-single timebox. Since there appears to be a divergence in trajectories
-after the recession, we introduce a second post-recession timebox,
-dragging it over series with higher and lower incomes during this second
-time period. This is the content of Figures \ref{fig:zillowmiddleup} and
-\ref{fig:zillowmiddledown}. Inspecting the highlighted tree nodes
-associated with these series, we find that most of the middle-income
-series that increased after the recession are associated with
-middle-income neighborhoods within the coastal Southern California
-counties. In constrast, those middle income series that saw decreases
-were mostly located in Central California and Oakland.
+Alternatively, we can study the trajectories of home prices among neighborhoods,
+conditional on their being middle-income before the recession. We generate the
+sequence of views in Figures \ref{fig:zillowmiddlepre},
+\ref{fig:zillowmiddleup}, and \ref{fig:zillowmiddledown} to this end. The first
+of these figures isolates neighborhoods with middle incomes before the
+recession, using a single timebox. Since there appears to be a divergence in
+trajectories after the recession, we introduce a second post-recession timebox,
+dragging it over series with higher and lower incomes during this second time
+period. This is the content of Figures \ref{fig:zillowmiddleup} and
+\ref{fig:zillowmiddledown}. Inspecting the highlighted tree nodes associated
+with these series, we find that most of the middle-income series that increased
+after the recession are associated with middle-income neighborhoods within the
+coastal Southern California counties. In constrast, those middle income series
+that saw decreases were mostly located in Central California and Oakland.
 
-The previous analysis highlights the fact that, within even narrow
-geographic regions, there can be substantial variation in prices. We can
-study this directly using treeboxes. In Figure \ref{fig:zillowsf} we
-have highlighted all series in San Francisco County. We see that, in
-2016, prices range from around \(10^{5.6} \approx \$400,000\) to
-\(10^{6.7} \approx \$5 \text{ million}\). So, while all these
-neighborhoods tend to be among the more expensive ones in California,
-prices can vary in a non-smooth way across geographic space.
+The previous analysis highlights the fact that, within even narrow geographic
+regions, there can be substantial variation in prices. We can study this
+directly using treeboxes. In Figure \ref{fig:zillowsf} we have highlighted all
+series in San Francisco County. We see that, in 2016, prices range from around
+\(10^{5.6} \approx \$400,000\) to \(10^{6.7} \approx \$5 \text{ million}\). So,
+while all these neighborhoods tend to be among the more expensive ones in
+California, prices can vary in a non-smooth way across geographic space.
 
 We conclude this example with a caveat that the Zillow data are not
-representative of all neighborhoods in California, only those with
-enough listings on the site, so should be supplemented by other data
-sources for any substantial decision-making.
+representative of all neighborhoods in California, only those with enough
+listings on the site, so should be supplemented by other data sources for any
+substantial decision-making.
 
@@
 collected across old and young mice. After variance-stabilizing these counts
-using DESeq2 \citep{love2014moderated}, a $t$-test was
-applied to each node in a phylogenetic tree, comparing abundances
-between old and young mice. To account for multiple testing, we employ
-the structSSI algorithm \citep{yekutieli2008hierarchical, sankaran2014structssi}
-along with methods available in the \texttt{multtest} package
-\citep{pollard2005multiple}.
+using DESeq2 \citep{love2014moderated}, a $t$-test was applied to each node in a
+phylogenetic tree, comparing abundances between old and young mice. To account
+for multiple testing, we employ the structSSI algorithm
+\citep{yekutieli2008hierarchical, sankaran2014structssi} along with methods
+available in the \texttt{multtest} package \citep{pollard2005multiple}.
 
-To interpret the results, we apply timebox trees. Our goals are to (1)
-identify subtrees with consistently elevated differential abundance
-across age groups and (2) compare alternative multiple testing
-adjustment procedures. Our approach is to display the negative-log raw
-and adjusted $p$-values for each node, with alternative methods
-compared via parallel coordinates. One view of the resulting display is
-captured in Figure \ref{fig:structssi}. First, we see that significant
-nodes tend to be significant across all methods -- the ordering between
-different series appears stable. Interestingly, the Sidak one-step and
+To interpret the results, we apply timebox trees. Our goals are to (1) identify
+subtrees with consistently elevated differential abundance across age groups and
+(2) compare alternative multiple testing adjustment procedures. Our approach is
+to display the negative-log raw and adjusted $p$-values for each node, with
+alternative methods compared via parallel coordinates. One view of the resulting
+display is captured in Figure \ref{fig:structssi}. First, we see that
+significant nodes tend to be significant across all methods -- the ordering
+between different series appears stable. Interestingly, the Sidak one-step and
 structSSI procedures seem to have lower power than the others, including
-conservative FWER-controlling methods, like the Bonferroni procedure.
-Further, in this application, FDR-controlling techniques do not seem to
-offer notably different adjusted $p$-values, relative to those
-controlling FWER. This suggests that, for this problem, bacteria are
-either strongly associated with age, or not associated at all, so that
-there is little gain from using more sensitive procedures.
+conservative FWER-controlling methods, like the Bonferroni procedure. Further,
+in this application, FDR-controlling techniques do not seem to offer notably
+different adjusted $p$-values, relative to those controlling FWER. This suggests
+that, for this problem, bacteria are either strongly associated with age, or not
+associated at all, so that there is little gain from using more sensitive
+procedures.
 
@@
 
-Further, selecting series with strong association between abundance and
-age, two major subtrees are brought to the forefront. Separately
-querying the the taxonomic identities of these bacteria reveals that they are
-two subgroups of Clostridia, which is consistent with the analysis of
+Further, selecting series with strong association between abundance and age, two
+major subtrees are brought to the forefront. Separately querying the the
+taxonomic identities of these bacteria reveals that they are two subgroups of
+Clostridia, which is consistent with the analysis of
 \citep{callahan2016bioconductor}. More than this specific analysis outcome, this
@@
 
-Here, we have reviewed some fundamental principles of data visualization
-and described their implementation in a new treelapse package. Further,
-we have provided examples of the practical usefulness of these
-principles in real-world data analysis situations.
+Here, we have reviewed some fundamental principles of data visualization and
+described their implementation in a new treelapse package. Further, we have
+provided examples of the practical usefulness of these principles in real-world
+data analysis situations.
 
@@
 considered the principle of arrangement in our visualizations
-\citep{buja1996interactive},
-though many of our conclusions were based on comparing alternative
-selections of the same display. We could imagine faceting our displays
-across groups to make these types of comparisons more accessible.
+\citep{buja1996interactive}, though many of our conclusions were based on
+comparing alternative selections of the same display. We could imagine faceting
+our displays across groups to make these types of comparisons more accessible.
 Further, we have only worked with the DOI distribution described in
@@
 \citep{hafen2013trelliscope, friedman2002john}. A simple extension could be to
-allow graph layouts instead of trees in time and
-treebox displays, for data that cannot be coerced into a hierarchical
-structure. Finally, if these ideas turn out to be useful in practice, it
-would be valuable to modularize the basic visualization layouts and
-relationships into a library, allowing the wider community to construct
-novel linked, interactive graphics with minimal effort.
+allow graph layouts instead of trees in time and treebox displays, for data that
+cannot be coerced into a hierarchical structure. Finally, if these ideas turn
+out to be useful in practice, it would be valuable to modularize the basic
+visualization layouts and relationships into a library, allowing the wider
+community to construct novel linked, interactive graphics with minimal effort.
 
-In summary, we have built an easily accessible R package for
-visualization techniques in a very specific methodology problem --
-analysis of differential abundance and dynamics in hierarchically
-structured data -- that appears in a variety of application domains. We
-have leveraged a link between R and D3
-\citep{vaidyanathan2014htmlwidgets} to create
-visualizations during the exploratory phase of data analysis; in this
-way our work is a departure from the culture of polished, journalistic
-visualizations prioritized by the D3 community. Finally, we have given a
-series of examples to demonstrate how the general visualization
-techniques of focus-plus-context and linked brushing can be practically
-incorporated into a range of practical analysis workflows, from studying
-the impact of bacteria on human health to better allocating units in
+In summary, we have built an easily accessible R package for visualization
+techniques in a very specific methodology problem -- analysis of differential
+abundance and dynamics in hierarchically structured data -- that appears in a
+variety of application domains. We have leveraged a link between R and D3
+\citep{vaidyanathan2014htmlwidgets} to create visualizations during the
+exploratory phase of data analysis; in this way our work is a departure from the
+culture of polished, journalistic visualizations prioritized by the D3
+community. Finally, we have given a series of examples to demonstrate how the
+general visualization techniques of focus-plus-context and linked brushing can
+be practically incorporated into a range of practical analysis workflows, from
+studying the impact of bacteria on human health to better allocating units in
 commuter bikesharing systems.

### 2017-07-12 treelapse.tex
add table of problem sizes

@@
 nodes in a prespecified tree structure. In the differential abundance problem,
-each random variable lives in \(\mathbb{R}^{G}\) where \(G\) is the number of
+each random variable lives in $\mathbb{R}^{G}$ where $G$ is the number of
 groups being compared. Each coordinate represents the abundance for that group,
@@
 different distributions. On the other hand, in the bacterial dynamics problem,
-each random variable is a time series, living in \(\mathbb{R}^{T}\).
+each random variable is a time series, living in $\mathbb{R}^{T}$.
 
@@
 data, where there is a noticeable lag in the timebox tree and treebox displays.
+In problems of larger size, we recommend a preliminary filtering or aggregation
+step across nodes or, if the time series is smooth, across neighboring
+timepoints, in order to avoid these potential scaling difficulties.
 
@@
 which we have limited ourselves.
+
+\begin{table}
+\centering
+\begin{tabular}{lll}
+Data            & Number of timepoints & Number of nodes \\
+Antibiotics     & 56                 & 386             \\
+Preterm births  & 216                & 318             \\
+Housing prices  & 254                & 944             \\
+Bikesharing     & 24                 & 819             \\
+Global patterns & 500                & 51             
+\end{tabular}
+\caption{Problem dimensions for each of the case studies. For problems with
+  dimensions larger than that in the housing prices problem, we recommend an
+  initial summarization or filtering step to prevent performance
+  issues. \label{problem-scaling}}
+\end{table}
 
@@
 
-\caption{Introducing a second box into the timebox display identifies the Bacteroidaceae as a taxon only midly impacted by antibiotics.}\label{fig:antibioticbacteroidetes}
+\caption{Introducing a second box into the timebox display identifies the
+  Bacteroidaceae as a taxon only midly impacted by
+  antibiotics.}\label{fig:antibioticbacteroidetes}
 \end{figure}
@@
 
-\caption{Zooming into the second antibiotic timecourse and highlighting the Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time to recover to pre-treatment levels.}\label{fig:antibioticfirmicutes}
+\caption{Zooming into the second antibiotic timecourse and highlighting the
+  Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
+  to recover to pre-treatment levels.}\label{fig:antibioticfirmicutes}
 \end{figure}
@@
 
-\caption{By hovering over the Ruminoccocus branches, we see that there is a prolonged effect of the antibiotics time courses more or less uniformly across the lower taxonomic orders.}\label{fig:ruminococcus}
+\caption{By hovering over the Ruminoccocus branches, we see that there is a
+  prolonged effect of the antibiotics time courses more or less uniformly across
+  the lower taxonomic orders.}\label{fig:ruminococcus}
 \end{figure}

### 2017-07-12 treelapse.tex
Add description of contributions / context

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath, amssymb, amsfonts}
 \usepackage{graphicx,psfrag,epsf}
@@
 
-Our key contribution is the adaptation of the visualization principles of
-focus-plus-context and linking to the study of tree-structured data
-\citep{buja1996interactive, becker1987brushing}. Our motivating application is
-to the analysis of bacterial time series, where an evolutionary tree relating
-bacteria is available a priori. However, we have identified common problem types
-where, if a tree is not directly available, it can be constructed from data and
-then studied using our techniques.
+In particular, we adapt the visualization principles of focus-plus-context and
+linking to the study of tree-structured data \citep{buja1996interactive,
+  becker1987brushing}. Our motivating application is to the analysis of
+bacterial time series, where an evolutionary tree relating bacteria is available
+a priori. However, we have identified common problem types where, if a tree is
+not directly available, it can be constructed from data and then studied using
+our techniques.
 
@@
 known, but powerful, visualization ideas.
+
+In summary, our key contributions are,
+\begin{itemize}
+\item Proposals for visualizing hierarchically structured data, based on
+  principles from the data visualization community.
+\item The implementation of these proposals in a publicly available R package.
+\item The illustration of the wide reach of hierarchical data visualization,
+  through case studies in both scientific and societal contexts.
+\end{itemize}
 
@@
 
-A microbiome is a community of bacteria living in given environments,
-for example, ocean water or the human gut
-\citep{human2012structure, cho2012human}. Progress in the field has
-been rapidly accelerated by the advent of genomic technologies, which
-enable detailed quantification of bacterial ecological structure and its
-influence in human and environmental health. Being concerned with both
-bacterial community structure and human health, the field exists at the
-border between ecology and medicine; consequently, papers in the area
-often apply a blend of exploratory data analysis and formal statistical
-inference.
+A microbiome is a community of bacteria living in given environments, for
+example, ocean water or the human gut \citep{human2012structure, cho2012human}.
+Progress in the field has been rapidly accelerated by the advent of genomic
+technologies, which enable detailed quantification of bacterial ecological
+structure and its influence in human and environmental health. Being concerned
+with both bacterial community structure and human health, the field exists at
+the border between ecology and medicine; consequently, papers in the area often
+apply a blend of exploratory data analysis and formal statistical inference.
 
-The two essential microbiome analysis problems that motivated our work
-are the tree-structured differential abundance and differential dynamics
-problems. In the differential abundance problem, we attempt to compare
-the abundances of individual bacteria across experimental conditions --
-for example, treatment vs. control or healthy vs. diseased. This is the
-microbiome analog of differential expression analysis in genomics
-\citep{anders2010differential}.
-We prepend the description ``tree-structured'' because, in practice,
-researchers generate interpretations about intermediate taxonomic orders
--- it is more interesting to discover novel behavior taxonomic levels
-between high-order phyla and low-level species. Hence, we frame the
-tree-structured differential abundance problem as the question of
-identifying the largest taxonomic subtree whose associated bacteria are
-differentially abundant.
+The two essential microbiome analysis problems that motivated our work are the
+tree-structured differential abundance and differential dynamics problems. In
+the differential abundance problem, we attempt to compare the abundances of
+individual bacteria across experimental conditions -- for example, treatment vs.
+control or healthy vs. diseased. This is the microbiome analog of differential
+expression analysis in genomics \citep{anders2010differential}. We prepend the
+description ``tree-structured'' because, in practice, researchers generate
+interpretations about intermediate taxonomic orders -- it is more interesting to
+discover novel behavior taxonomic levels between high-order phyla and low-level
+species. Hence, we frame the tree-structured differential abundance problem as
+the question of identifying the largest taxonomic subtree whose associated
+bacteria are differentially abundant.
 
-In the tree-structured bacterial dynamics problem, the goal is to
-describe changes in bacterial abundances in an environment over time. As
-in the differential abundance problem, it is useful if these
-descriptions can be given at the highest subtree at which the pattern
-appears. Specific questions of interest often have an ecological flavor.
-For example, researchers are often interested in understanding how
-bacterial populations respond to sudden or gradual environmental changes
-or how species fill, drop out from, or compete for environmental niches.
-Medically, these questions are important for illuminating the impact of
-antibiotic time courses or diet changes, for example.
+In the tree-structured bacterial dynamics problem, the goal is to describe
+changes in bacterial abundances in an environment over time. As in the
+differential abundance problem, it is useful if these descriptions can be given
+at the highest subtree at which the pattern appears. Specific questions of
+interest often have an ecological flavor. For example, researchers are often
+interested in understanding how bacterial populations respond to sudden or
+gradual environmental changes or how species fill, drop out from, or compete for
+environmental niches. Medically, these questions are important for illuminating
+the impact of antibiotic time courses or diet changes, for example.
 
@@
 ideas from the visualization literature that can be applied to answer them. As
-the core difficulty is high-dimensionality, so it should be no surprise that the
+the core difficulty is high-dimensionality, it should be no surprise that the
 techniques we adapt come from the literature on high-dimensional data
-visualization, namely, focus-plus-context and linking.
+visualization, which has enjoyed rapid progress in the last 25 years. Modern
+research in this domain develops abstractions and taxonomies for guiding
+visualization designs so that they most effectively communicates properties of
+the data to their intended audience. A major push in this body of work explores
+the potential for interactivity to improve many stages of the data analysis
+process, from preliminary data preparation, to refinement and navigation across
+views, to final sharing and annotation of results \citep{heer2012taxonomy}.
+Further work has attempted to bridge the gap between statistical analysis and
+data visualization methodology, both of which provide techniques for learning
+from high-dimensional data \citep{de2003visual}.
 
-The focus-plus-context principle is that large collections of visual elements
-can be studied at multiple scales, by simultaneously focusing" on a few elements
-of interest and maintaining the ``context'' of the coarser-scale background. A
+The problem structures most relevant to our study are tree and temporal
+structure, and the visualization community has various ways of reasoning about
+these data, see \citep{graham2010survey, aigner2011visualization} for detailed
+surveys. From this literature, our approach is most directly informed by the
+focus-plus context and linking principles, which we briefly review here. The
+focus-plus-context principle is that large collections of visual elements can be
+studied at multiple scales, by simultaneously focusing" on a few elements of
+interest and maintaining the ``context'' of the coarser-scale background. A
 simple example of this idea is to include a search box that highlights matching
@@
 
-\caption{The two antibiotic time courses are readily apparent even when only highlighting the most abundant taxa.
-We have annotated the figure with its three main components.}\label{fig:antibioticoverview}
+\caption{The two antibiotic time courses are readily apparent even when only
+  highlighting the most abundant taxa. We have annotated the figure with its
+  three main components.}\label{fig:antibioticoverview}
 \end{figure}
@@
 This is because, at each branching point, we place the node with larger average
-value on the left. Figure \ref{fig:fig:verrucomicrobiae} notes that the nodes at
+value on the left. Figure \ref{fig:verrucomicrobiae} notes that the nodes at
 the far left in the tree are associated with phylum Verrucomicrobiae,
@@
 
-\caption{The increased diversity among samples in CST 4 is represented by the relatively larger contribution of red edges to branches outside of the Firmicutes.}\label{fig:pretermcsts}
+\caption{The increased diversity among samples in CST 4 is represented by the
+  relatively larger contribution of red edges to branches outside of the
+  Firmicutes.}\label{fig:pretermcsts}
 \end{figure}
@@
 
-\caption{Zooming into the Lactobacillaceae family, we notice that the difference between the remaining four CSTs is related to which types of Lactobacillus are most prominent.}\label{fig:pretermcstslacto}
+\caption{Zooming into the Lactobacillaceae family, we notice that the difference
+  between the remaining four CSTs is related to which types of Lactobacillus are
+  most prominent.}\label{fig:pretermcstslacto}
 \end{figure}
@@
 
-\caption{Samples with high levels of phyla other than Firmicutes appear to be related to preterm births.}\label{fig:pretermpreterm}
+\caption{Samples with high levels of phyla other than Firmicutes appear to be
+  related to preterm births.}\label{fig:pretermpreterm}
 \end{figure}
@@
 
-\caption{California home prices at the city level, between 1996 and 2016.
-The effect of the 2008 recession is clear, and we have hovered over the San Bernardino series, to display the identity of one of the cities most strongly affected by the recession.}\label{fig:zillowcities}
+\caption{California home prices at the city level, between 1996 and 2016. The
+  effect of the 2008 recession is clear, and we have hovered over the San
+  Bernardino series, to display the identity of one of the cities most strongly
+  affected by the recession.}\label{fig:zillowcities}
 \end{figure}
@@
 
-\caption{Neighborhoods with mid-range home prices before the recession are selected here.
-Note that the collection of series seems to widen after 2008 -- we are interested in whether there are reliable predictors of these alternate trajectories, given their similar starting points.}\label{fig:zillowmiddlepre}
+\caption{Neighborhoods with mid-range home prices before the recession are
+  selected here. Note that the collection of series seems to widen after 2008 --
+  we are interested in whether there are reliable predictors of these alternate
+  trajectories, given their similar starting points.}\label{fig:zillowmiddlepre}
 \end{figure}
@@
 
-\caption{Among those neighborhoods with mid-range prices before the recession, we have selected those that recovered more rapidly.
-These appear to be located mainly in Los Angeles and San Diego counties.}\label{fig:zillowmiddleup}
+\caption{Among those neighborhoods with mid-range prices before the recession,
+  we have selected those that recovered more rapidly. These appear to be located
+  mainly in Los Angeles and San Diego counties.}\label{fig:zillowmiddleup}
 \end{figure}
@@
 
-\caption{Rather than selecting mid-range series that recovered quickly, we can isolate those whose prices remained depressed after the recession.
-These seem mostly to be located in Central California and the East San Francisco Bay Area.}\label{fig:zillowmiddledown}
+\caption{Rather than selecting mid-range series that recovered quickly, we can
+  isolate those whose prices remained depressed after the recession. These seem
+  mostly to be located in Central California and the East San Francisco Bay
+  Area.}\label{fig:zillowmiddledown}
 \end{figure}
@@
 
-\caption{To study the range in home prices within San Francisco County, we can select the associated nodes using a treebox.}\label{fig:zillowsf}
+\caption{To study the range in home prices within San Francisco County, we can
+  select the associated nodes using a treebox.}\label{fig:zillowsf}
 \end{figure}
@@
 
-\caption{The two peaks at rush hour distinguish weekday series from the rest.}\label{fig:working}
+\caption{The two peaks at rush hour distinguish weekday series from the
+  rest.}\label{fig:working}
 \end{figure}
@@
 
-\caption{Unlike weekday demand, weekend demand is unimodal.
-The few weekday series with unimodal series seem to be associated with holidays.
-This is the case for New Years' Eve, which is currently hovered over in the tree.}\label{fig:weekend}
+\caption{Unlike weekday demand, weekend demand is unimodal. The few weekday
+  series with unimodal series seem to be associated with holidays. This is the
+  case for New Years' Eve, which is currently hovered over in the
+  tree.}\label{fig:weekend}
 \end{figure}
@@
 
-\caption{Weeday demand appears larger in 2012 than 2011 -- compare with the next figure.}\label{fig:weekday2011}
+\caption{Weeday demand appears larger in 2012 than 2011 -- compare with the next
+  figure.}\label{fig:weekday2011}
 \end{figure}
@@
 
-\caption{Weekday demand increased in 2012 -- compare with the previous figure.}\label{fig:weekday2012}
+\caption{Weekday demand increased in 2012 -- compare with the previous
+  figure.}\label{fig:weekday2012}
 \end{figure}
@@
 
-\caption{The samples with highest night demand tend to fall on warm weekends.}\label{fig:warmweekend}
+\caption{The samples with highest night demand tend to fall on warm
+  weekends.}\label{fig:warmweekend}
 \end{figure}
@@
 
-\caption{We can interpret the regression tree using an interactive DOI representation.}\label{fig:bikesankey}
+\caption{We can interpret the regression tree using an interactive DOI
+  representation.}\label{fig:bikesankey}
 \end{figure}
@@
 
-\caption{An application to the Global Patterns demonstrates how linking in treelapse can be applied to combine hierarchical clustering and parallel coordinates views.}\label{fig:gptimebox}
+\caption{An application to the Global Patterns demonstrates how linking in
+  treelapse can be applied to combine hierarchical clustering and parallel
+  coordinates views.}\label{fig:gptimebox}
 \end{figure}
@@
 
-\caption{Viewing a tree of $p$-values across different methods highlights two subtrees with strong associations with mouse age, across several testing procedures.}\label{fig:structssi}
+\caption{Viewing a tree of $p$-values across different methods highlights two
+  subtrees with strong associations with mouse age, across several testing
+  procedures.}\label{fig:structssi}
 \end{figure}
@@
 exploratory phase of data analysis; in this way our work is a departure from the
-culture of polished, journalistic visualizations prioritized by the D3
-community. Finally, we have given a series of examples to demonstrate how the
-general visualization techniques of focus-plus-context and linked brushing can
-be practically incorporated into a range of practical analysis workflows, from
+culture of polished, journalistic visualizations prioritized by the D3 community
+and is more closely aligned with the vision in \citep{de2003visual} of more
+tightly integrating data visualization and statistical analysis techniques.
+Finally, we have given a series of examples to demonstrate how the general
+visualization techniques of focus-plus-context and linked brushing can be
+practically incorporated into a range of practical analysis workflows, from
 studying the impact of bacteria on human health to better allocating units in

### 2017-07-12 reviewer_response.tex
update response to reviewers

@@
 
-We have have modified our literature review to begin with a broader discussion
-of tree and time series visualization methods, before proceeding to those
-methods used more directly in our work. Further, we have added a bullet list
-describing more concretely the contributions of this paper within the broader
-body of knowledge.
+We have added material at the beginning of the literature review (Section 1.3) that
+describes the larger body of work within which our study lives. We explain some of
+the main ideas in the surveys you mentioned before proceeding to the more narrow
+set of ideas that are directly adapted in our work. Further, we have provided a
+bullet list in the introduction that explicitly describes the contribution of
+this paper.
 
@@
 Indeed, we implemented the first and third recommendations described in
-\href{https://github.com/d3/d3/issues/1604}{this d3 issue}, but in both cases we
+\href{https://github.com/d3/d3/issues/1604}{this D3 issue}, but in both cases we
 found it necessary to introduce a lag between mousedown and the start of
@@
 
-We have added a description in Section ? describing the scales of different
-inputs before the interface begins to lag.
+We have added a discussion of scalability at the end of Section 1.4. We also
+explicitly list the problem dimensions in Table 1. Also, all examples (and more)
+are available as pre-compiled Rmarkdown vignettes linked from our package
+webpage, so users can assess utility in this examples easily without having to
+even install the package.
 

### 2017-07-12 treelapse.tex
update captions in all the microbiome examples

@@
 \centering
-\begin{tabular}{lll}
+\begin{tabular}{|l|l|l|}
+  \hline
   Data            & Number of timepoints & Number of nodes \\
+  \hline
   Antibiotics     & 56                 & 386             \\
@@
   Bikesharing     & 24                 & 819             \\
-Global patterns & 500                & 51             
+  Global patterns & 500                & 51             \\
+  \hline
 \end{tabular}
@@
 
-\caption{The two antibiotic time courses are readily apparent even when only
-  highlighting the most abundant taxa. We have annotated the figure with its
-  three main components.}\label{fig:antibioticoverview}
+\caption{Here we display the primary timebox tree view of the antibiotics data
+  set from \citep{dethlefsen2008}, annotated with the main components of the
+  visualization. The tree at the top is a taxonomic tree of all the bacteria
+  contained within the sample, and it is visually linked to the time series at
+  the bottom: each node in the tree corresponds to a path among the time series.
+  The selection brush is used to focus attention on the time series that go
+  through it -- these are highlighted in blue -- and other brushes can be added
+  using a button not displayed here. The pan-zoom widget at the top right is
+  used to update the scales of the main time series display so that only
+  particular time windows and $y$-axis ranges are visible. This view is the
+  basis for all the timebox tree and treebox displays that appear
+  below. \label{antibioticoverview}}
 \end{figure}
@@
 \caption{Introducing a second box into the timebox display identifies the
-  Bacteroidaceae as a taxon only midly impacted by
-  antibiotics.}\label{fig:antibioticbacteroidetes}
+  Bacteroidetes as a taxon only midly impacted by antibiotics. The layout is
+  identical to Figure 1, except two small brushes are placed over the time
+  series between 10 and 20 days, and now only those time series and
+  corresponding nodes in the tree are highlighted in blue. Further, the user has
+  hovered over the top blue node in the tree, revealing the taxonomic identity
+  of these series. Hence, brushing the time series and linking with the tree can
+  be used to discover and characterize notable variation within the
+  data.\label{fig:antibioticbacteroidetes}}
 \end{figure}
@@
   Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
-  to recover to pre-treatment levels.}\label{fig:antibioticfirmicutes}
-\end{figure}
-
-Alternatively, we can summarize each node by the average across its
-descendants -- this brings attention to individual bacteria that may be
-underlying some of the broader taxonomic patterns we have noted when
-studying the subtree sums. For example, in Figure
-\ref{fig:ruminococcus}, we highlight all families below order
-Ruminococcus, suggesting that the decrease due to antibiotics occurs
-uniformly across almost all families. A point that was not evident in
-the earlier sum-across-descendants view is that, after the second
-treatment of antibiotics, a few of the Ruminoccocus families recover
-more rapidly than the rest, for example the Unc095d3 (highlighted in
-red) are only briefly affected. In contrast, most families seem to
-recover in unison after the first treatment.
+  to recover to pre-treatment levels. Here, the red lines and nodes are those
+  that match the text search provided by the user in a search box just below the
+  figure (not displayed here). Hovering the mouse over these lines provides
+  their identities -- the top red line is Lachnospiraceae, and the bottom red
+  line is Ruminoccocus. Note that the brush in the treebox display is located
+  over the tree, rather than over the time series. In particular, the search box
+  and interactive brushing can be combined to interrogate hypotheses of apriori
+  interest. \label{fig:antibioticfirmicutes}}
+\end{figure}
+
+Alternatively, we can summarize each node by the average across its descendants
+-- this brings attention to individual bacteria that may be underlying some of
+the broader taxonomic patterns we have noted when studying the subtree sums. For
+example, in Figure \ref{fig:ruminococcus}, we highlight all families below order
+Ruminococcus, suggesting that the decrease due to antibiotics occurs uniformly
+across almost all families. A point that was not evident in the earlier
+sum-across-descendants view is that, after the second treatment of antibiotics,
+a few of the Ruminoccocus families recover more rapidly than the rest, for
+example the Unc095d3 (highlighted in red) are only briefly affected. In
+contrast, most families seem to recover in unison after the first treatment.
 
@@
   prolonged effect of the antibiotics time courses more or less uniformly across
-  the lower taxonomic orders.}\label{fig:ruminococcus}
+  the lower taxonomic orders. The graphical elements are the same as before,
+  except the user has searched for Ruminoccocus and species Unc095d3, which has
+  the highest average abundance within this taxon. By displaying averages rather
+  than sums, we see that the effect of antibiotics visible at higher taxonomic
+  orders is not created by a single dominant species becoming less abundant, but
+  rather the decline in populations across all descendant species. The same
+  display applied to different data can yield different
+  insights.}\label{fig:ruminococcus}
 \end{figure}
@@
   abundant. In particular, they seem to increase after the first antibiotic time
-  course.}\label{fig:verrucomicrobiae}
+  course, which occurs between days 15 and 20. This view was generated by
+  placing a brush over the branch on the far left, which has those nodes with
+  the largest averages across all timepoints. The user's mouse is over the blue
+  series, which brings up the associated taxonomic label. The determination of
+  species whose abundances increase during antibiotics, which would require many
+  hypothesis tests using a more standard approach, becomes quickly apparent
+  via interactive visualization.}\label{fig:verrucomicrobiae}
 \end{figure}
@@
 
-\caption{The increased diversity among samples in CST 4 is represented by the
-  relatively larger contribution of red edges to branches outside of the
-  Firmicutes.}\label{fig:pretermcsts}
+\caption{The increased diversity among samples in community state type (CST) 4
+  is represented by the relatively larger contribution of red edges to branches
+  outside of the Firmicutes. This display shows the top of the DOI sankey
+  visualization of the preterm birth data studied in
+  \citep{digiulio2015temporal}. The root of the tree is the taxonomic kingdom
+  Bacteria, its children are labeled according to their phylum names. Each color
+  within a branch is associated with CST, and the width of the associated color
+  corresponds to the average abundance of that taxonomic group among all samples
+  belonging to the corresponding CST, as indicated by the legend at the left.
+  Phyla are sorted from most to least abundant. This initial display of the DOI
+  sankey provides a summary of overall abundances across taxonomic groups and
+  CSTs, and suggests subtrees to navigate into, to extract more detailed
+  abundance information.
+}\label{fig:pretermcsts}
 \end{figure}
@@
   between the remaining four CSTs is related to which types of Lactobacillus are
-  most prominent.}\label{fig:pretermcstslacto}
+  most prominent. The DOI sankey refocuses the tree around the last group that
+  was clicked on, in this case showing more detail about order Lactobacillales
+  and its descendants. The overview display can be recovered by navigating back
+  up to higher-level taxa. Hence, it is possible to navigate between broad
+  overview and detailed displays in a way that facilitates interpretation of
+  results from statistical analysis.}\label{fig:pretermcstslacto}
 \end{figure}
@@
 \caption{Samples with high levels of phyla other than Firmicutes appear to be
-  related to preterm births.}\label{fig:pretermpreterm}
+  related to preterm births. Here, we again display the preterm birth data with
+  a DOI sankey, but instead of grouping samples according to statistically
+  generated CSTs, we directly assign samples to preterm vs. not preterm
+  according to whether the mother eventually had a preterm birth. These new
+  states are visible in the updated legend. Through interactivity, it becomes
+  possible to develop meaningful visual summaries even before calculating formal
+  statistical ones.}
+}\label{fig:pretermpreterm}
 \end{figure}

### 2017-07-12 treelapse.tex

@@
 time courses separated by 6 months. Discerning the variation in resilience
-across bacteria is important, considering the the role of bacteria in health and
+across bacteria is important, considering the role of bacteria in health and
 not just disease.
@@
 \caption{Here we display the primary timebox tree view of the antibiotics data
-  set from \citep{dethlefsen2008}, annotated with the main components of the
+  set from \citep{dethlefsen2008pervasive}, annotated with the main components of the
   visualization. The tree at the top is a taxonomic tree of all the bacteria
@@
   basis for all the timebox tree and treebox displays that appear
-  below. \label{antibioticoverview}}
+  below. \label{fig:antibioticoverview}}
 \end{figure}
@@
 \caption{Introducing a second box into the timebox display identifies the
-  Bacteroidetes as a taxon only midly impacted by antibiotics. The layout is
+  Bacteroidetes as a taxon only mildly impacted by antibiotics. The layout is
   identical to Figure 1, except two small brushes are placed over the time
@@
 
-\caption{Zooming into the second antibiotic timecourse and highlighting the
+\caption{Zooming into the second antibiotic time course and highlighting the
   Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
@@
   over the tree, rather than over the time series. In particular, the search box
-  and interactive brushing can be combined to interrogate hypotheses of apriori
+  and interactive brushing can be combined to interrogate hypotheses of a priori
   interest. \label{fig:antibioticfirmicutes}}
@@
 
-{\centering \includegraphics[width=300px]{figure/preterm_preterm}
-
-}
+{\centering \includegraphics[width=300px]{figure/preterm_preterm}}
 
@@
   statistical ones.}
-}\label{fig:pretermpreterm}
+\label{fig:pretermpreterm}
 \end{figure}
@@
 after the recession are associated with middle-income neighborhoods within the
-coastal Southern California counties. In constrast, those middle income series
+coastal Southern California counties. In contrast, those middle income series
 that saw decreases were mostly located in Central California and Oakland.
@@
 
-\caption{California home prices at the city level, between 1996 and 2016. The
-  effect of the 2008 recession is clear, and we have hovered over the San
-  Bernardino series, to display the identity of one of the cities most strongly
-  affected by the recession.}\label{fig:zillowcities}
+\caption{California home prices at the city level, between 1996 and 2016. Here,
+  we provide a treebox display, as in Figure \ref{fig:antibioticfirmicutes}, but
+  of the California home prices data. In this case, the tree corresponds to a
+  geographic hierarchy, with regions at the top and neighborhoods at the bottom, while
+  time series correspond to home prices collected weekly between 1996 and 2016.
+  A brush is drawn over all nodes at the city level. The effect of the 2008
+  recession is clear, and we have hovered over the San Bernardino series, to
+  display the identity of one of the cities most strongly affected by the
+  recession. This provides an example of a dataset with meaningful (spatial)
+  hierarchical structure that might not initially be studied using trees.
+\label{fig:zillowcities}}
 \end{figure}
@@
 \caption{Neighborhoods with mid-range home prices before the recession are
-  selected here. Note that the collection of series seems to widen after 2008 --
-  we are interested in whether there are reliable predictors of these alternate
-  trajectories, given their similar starting points.}\label{fig:zillowmiddlepre}
+  brushed here, using a timebox tree version of Figure \ref{fig:zillowcities}.
+  The associated tree nodes are highlighted at the top. Note that the collection
+  of series seems to widen after 2008 -- we are interested in whether there are
+  reliable predictors of these alternate trajectories, given their similar
+  starting points. This serves as a baseline with which to compare Figure
+  \ref{fig:zillowmiddleup} -- these views are easy to transition between
+  interactively.}\label{fig:zillowmiddlepre}
 \end{figure}
@@
 \caption{Among those neighborhoods with mid-range prices before the recession,
-  we have selected those that recovered more rapidly. These appear to be located
-  mainly in Los Angeles and San Diego counties.}\label{fig:zillowmiddleup}
+  displayed in Figure \ref{fig:zillowmiddlepre}, we have selected those that
+  recovered more rapidly by introducing a brush at the top right of the time
+  series panel. By hovering a brush over a collection of tree nodes that all
+  seem to be highlighted, we infer that the associated neighborhoods are located
+  mainly in Los Angeles and San Diego counties. This follow-up view is
+  interesting to contrast with Figure \ref{fig:zillowmiddledown}.
+}\label{fig:zillowmiddleup}
 \end{figure}
@@
 
-\caption{Rather than selecting mid-range series that recovered quickly, we can
-  isolate those whose prices remained depressed after the recession. These seem
-  mostly to be located in Central California and the East San Francisco Bay
-  Area.}\label{fig:zillowmiddledown}
+\caption{In contrast to Figure \ref{fig:zillowmiddleup}, we can follow-up the
+  selection in Figure \ref{fig:zillowmiddlepre} by isolating those neighborhoods
+  where prices remained depressed after the recession. This is accomplished by
+  moving the brush on the right down towards lower prices. Hovering over the
+  associated highlighted nodes in the tree to reveal the associated locations,
+  we see that most of these series correspond to neighborhoods in Central
+  California and the East San Francisco Bay Area. The ability to sketch the
+  overall shapes of series using brushes and interpret the associated selections
+  using a tree simplifies what might otherwise be complex
+  comparisons.}\label{fig:zillowmiddledown}
 \end{figure}
@@
 
-\caption{To study the range in home prices within San Francisco County, we can
-  select the associated nodes using a treebox.}\label{fig:zillowsf}
+\caption{In contrast to Figures \ref{fig:zillowmiddlepre} to
+  \ref{fig:zillowmiddledown}, we can study the variation in series associated
+  with a subset of the tree by using treeboxes. To study the range in home
+  prices within San Francisco County, we can brush over the associated nodes in
+  the tree. The red circle indicates that the user has searched for ``San
+  Francisco,'' which guides the user to the appropriate subtree. The mouse is
+  hovering over one of the more expensive neighborhoods. Hence, though the
+  timebox tree and treebox views are similar, they are directed towards
+  different types of visual comparisons.}\label{fig:zillowsf}
 \end{figure}
@@
 
-\caption{The two peaks at rush hour distinguish weekday series from the
-  rest.}\label{fig:working}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=350px]{figure/weekend}
-
-}
-
-\caption{Unlike weekday demand, weekend demand is unimodal. The few weekday
-  series with unimodal series seem to be associated with holidays. This is the
-  case for New Years' Eve, which is currently hovered over in the
-  tree.}\label{fig:weekend}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=350px]{figure/weekday_2011}
-
-}
-
-\caption{Weeday demand appears larger in 2012 than 2011 -- compare with the next
-  figure.}\label{fig:weekday2011}
+\caption{The two peaks at rush hour distinguish weekday series from the rest.
+  through the timebox tree view. The display is the same type of timebox tree
+  view introduced in Figure \ref{fig:antibioticoverview}, but applied to the
+  bikesharing data, where the time axis represents the time of day and the
+  $y$-axis gives bikesharing demand. Each series is the bikesharing demand for a
+  single day, over the course of two years. The tree now corresponds to the
+  regression tree generated by predicting demand at 8am using supplementary
+  data. Two brushes are introduced to highlight the double peaks corresponding
+  to rush hours on weekdays. We see that although hierarchical structure was not
+  present immediately in the bikesharing data, it is useful to introduce and
+  interpret such structure by combining regression and visualization
+  methodology.
+}\label{fig:working}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=350px]{figure/weekend}}
+
+\caption{By adjusting the two brushes in Figure \ref{fig:working}, we see that
+  unlike weekday demand, weekend demand is unimodal. The few weekday series with
+  unimodal series seem to be associated with holidays. This is the case for New
+  Years' Eve, which is currently hovered over in the tree. The ease of
+  transitioning from Figure \ref{fig:working} to this display indicates the
+  importances of brushing in interaction.}\label{fig:weekend}
+\end{figure}
+
+\begin{figure}
+{\centering \includegraphics[width=350px]{figure/weekday_2011}}
+
+\caption{Weekday demand appears larger in 2012 than 2011 -- compare with Figure 
+  \ref{fig:weekday2012}. Here, a brushes are introduced over the tree to see the
+  series associated with a particular split point. The red nodes are the results
+  of searches over the two nodes that are children of this split point. The fact
+  that the \texttt{yr>=0.5} selected line is larger than the \texttt{yr<0.5}
+  line means that demand was larger in 2012. In combination with searching and
+  treeboxes, it is possible to interpret more subtle split points in the decision
+  tree.}
+\label{fig:weekday2011}
 \end{figure}
@@
 
-\caption{Weekday demand increased in 2012 -- compare with the previous
-  figure.}\label{fig:weekday2012}
+\caption{Weekday demand increased in 2012 -- compare with 
+  Figure \ref{fig:weekday2011}. The search terms are the same as in that figure,
+  but the subtree associated with the 2012 split point is highlighted, using the
+  union of two boxes. Note that unlike timebox trees, which highlighted series
+  lying through the intersection of brushes, treeboxes highlight series within the
+  union of brushes.}\label{fig:weekday2012}
 \end{figure}
@@
 \caption{The samples with highest night demand tend to fall on warm
-  weekends.}\label{fig:warmweekend}
+  weekends. Here, the pan-zoom widget has been used to adjust both time and
+  demand axes to narrow on a specific window of interest. Brushes are drawn over
+  the larger among these series, and the corresponding tree nodes are located
+  close to one another, in the part of the tree corresponding to the warm / cold
+  average temperature split. More generally, panning and zooming allows
+  navigation between focus and context.}\label{fig:warmweekend}
 \end{figure}
@@
 
-\caption{We can interpret the regression tree using an interactive DOI
-  representation.}\label{fig:bikesankey}
+\caption{So far, we have focused on the timebox tree and treebox representations
+  of the bikesharing data -- a complementary view is provided by the DOI sankey.
+  Here, the tree is the result of the regression tree procedure, while the
+  colors represent particular quantiles of 8am demand. This allows the
+  determination of which split descendants are associated with low or high
+  demand.}\label{fig:bikesankey}
 \end{figure}
@@
 visualization and modeling techniques in an iterative, nonlinear
-fashion.
+fashion, in the spirit of \citep{de2003visual}.
 
@@
   treelapse can be applied to combine hierarchical clustering and parallel
-  coordinates views.}\label{fig:gptimebox}
+  coordinates views. Each panel represents a different subtree cluster within
+  this data set, as indicated by the different locations for the tree brushes.
+  The paths in the lower halves of each display represent the average value
+  across different bacteria rather than timepoints, as in all previous figures.
+  Though the samples originally do not include any hierarchical structure,
+  hierarchical clustering provides such a structure which can then be
+  interpreted using treeboxes. }\label{fig:gptimebox}
 \end{figure}
@@
   subtrees with strong associations with mouse age, across several testing
-  procedures.}\label{fig:structssi}
+  procedures. The tree represents the taxonomy of bacteria, and the series
+  provides the negative log $p$-values associated with nodes as computed by
+  different tests, listed along the $x$-axis as in parallel coordinates. By
+  selecting series with larger values for a test, we see the associated subtrees
+  of significant $p$-values. Hence, hierarchical views can be useful even in the
+  confirmatory testing settings which typically study results from individual
+  tests in isolation from each other.\label{fig:structssi}}
 \end{figure}
@@
 Further, selecting series with strong association between abundance and age, two
-major subtrees are brought to the forefront. Separately querying the the
+major subtrees are brought to the forefront. Separately querying the
 taxonomic identities of these bacteria reveals that they are two subgroups of

### 2017-07-13 treelapse.tex
Add links + update figures
(red search lines now appear above blue select lines)

@@
 modeling to interactive exploration, and vice versa. Our code is open-source,
-and available at \url{https://github.com/krisrs1128/treelapse}. We hope this
-package encourages data analysts to work at the border between data modeling and
-visualization, and more generally empowers a wider audience to apply less widely
-known, but powerful, visualization ideas.
+and is linked in Section \ref{sec:supp_links} of the supplementary materials. We
+hope this package encourages data analysts to work at the border between data
+modeling and visualization, and more generally empowers a wider audience to
+apply less widely known, but powerful, visualization ideas.
 
@@
 We next consider an application unrelated to the microbiome, but with relatively
-clear hierarchical structure. Our data are downloaded from
-Zillow\footnote{\url{http://www.zillow.com/research/data/}}, and give the Zillow
-Home Value Indexes at the neighborhood level, across the country, computed
-monthly between 1996 and 2016. In our display, we have taken the base 10 log of
-these indexes. As our hierarchical structure, we use each neighborhood's
-assignment to state, regional, county, and city levels. We represent each of
-these coarser spatial categories using the average of all neighborhoods
-contained in them. We have filtered down to the 890 neighborhoods in California;
-rendering more neighborhoods while keeping all 246 timepoints causes the
-interface to lag\footnote{See below for potential optimizations, however.}.
+clear hierarchical structure. Our data are downloaded from Zillow and give the
+Zillow Home Value Indexes at the neighborhood level, across the country,
+computed monthly between 1996 and 2016. A link to the data source is provided in
+Section \ref{sec:supp_links} of the supplementary materials. In our display, we
+have taken the base 10 log of these indexes. As our hierarchical structure, we
+use each neighborhood's assignment to state, regional, county, and city levels.
+We represent each of these coarser spatial categories using the average of all
+neighborhoods contained in them. We have filtered down to the 890 neighborhoods
+in California; rendering more neighborhoods while keeping all 246 timepoints
+causes a severe lag in the interface.
 
@@
 
-Our next example is a study in bikesharing demand, included as an
-example of analyzing collections of time series when there is no obvious
-hierarchical structure a priori. The data are available at the UCI
-Repository\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset}}
-and were originally collected by a Washington D.C.-based bikesharing
-system for use in a Kaggle prediction competition. The data are hourly
-measurements of bike demand, aggregated across all bikesharing stations,
-over two years, along with supplemental weather data. In the
-competition, participants were asked to predict the hourly demand on a
-held-out test set. Here, we adopt a descriptive view instead, attempting
-to characterize factors associated with variation in bikesharing demand.
+Our next example is a study in bikesharing demand, included as an example of
+analyzing collections of time series when there is no obvious hierarchical
+structure a priori. The data are available at the UCI Repository and were
+originally collected by a Washington D.C.-based bikesharing system for use in a
+Kaggle prediction competition. A link is provided in Section
+\ref{sec:supp_links} of the supplementary materials. The data are hourly
+measurements of bike demand, aggregated across all bikesharing stations, over
+two years, along with supplemental weather data. In the competition,
+participants were asked to predict the hourly demand on a held-out test set.
+Here, we adopt a descriptive view instead, attempting to characterize factors
+associated with variation in bikesharing demand.
 
@@
 \begin{figure}
-
-{\centering \includegraphics[width=350px]{figure/weekend}}
-
+  \centering
+  \includegraphics[width=350px]{figure/weekend}
     \caption{By adjusting the two brushes in Figure \ref{fig:working}, we see that
@@
 \begin{figure}
-{\centering \includegraphics[width=350px]{figure/weekday_2011}}
-
+  \centering
+  \includegraphics[width=350px]{figure/weekday_2011}
     \caption{Weekday demand appears larger in 2012 than 2011 -- compare with Figure 
@@
       treeboxes, it is possible to interpret more subtle split points in the decision
-  tree.}
-\label{fig:weekday2011}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=350px]{figure/weekday_2012}
-
-}
-
+      tree.} \label{fig:weekday2011}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=350px]{figure/weekday_2012}
   \caption{Weekday demand increased in 2012 -- compare with 
@@
 
+\section{Supplemental Materials}
+
+\subsection{Links}
+\label{sec:supp_links}
+
+\begin{itemize}
+\item treelapse home page: \url{https://krisrs1128.github.io/treelapse/}
+\item treelapse development page: \url{https://github.com/krisrs1128/treelapse}
+\item Home prices data: \url{http://www.zillow.com/research/data/}
+\item Bikesharing data: \url{https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset}
+\end{itemize}
+
 \end{document}

### 2017-07-13 reviewer_response.tex
complete reviewer response

@@
 \usepackage{natbib}
+\usepackage{setspace}
 \usepackage{graphicx}
+\usepackage{hyperref}
+\usepackage{color}
 \input{preamble.tex}
 
+\title{Response to Reviews: ``Interactive Visualization of Hierarchically Structured Data''}
+\author{Kris Sankaran and Susan P. Holmes}
+\linespread{1.25}
+
 \begin{document}
+\maketitle
 
-We would like to thank the two reviewers and the associate editor for insightful
-and constructive comments on our work. Below, we describe changes to the paper
-text and software based on this feedback.
+We would like to thank the two reviewers and the editor and associate editor for
+insightful and constructive comments on our work. Below, we document changes to
+the paper text and software based on this feedback. The original reviewer
+comments are printed in blue, and our responses are in black.
 
@@
 
-\textit{This small sampling of a large body of work points to the need for a
-  cleaner conceptualization and clarifications of the contribution of the work
-  presented in this paper relative to prior efforts.}
+\color{blue}
+
+Although I have a few comments on the implementation and reporting detail (given
+below), my main complaint with this paper lies in the lack of context. Although
+the authors make some reference to DOI Trees and Timeboxes, there is very little
+discussion of the wide range of relevant approaches that have been explored in
+25 years of tree visualization efforts. Review papers such as Heer \&
+Shneiderman Interactive Dynamics for Visual Analysis
+(https://doi.org/10.1145/2133416.2146416), and Oliveir and Levkwoitz From visual
+data exploration to visual data mining: a survey (10.1109/TVCG.2003.1207445)
+provide general background, while Graham and Kennedy A Survey of Multiple Tree
+Visualisation (http://journals.sagepub.com/doi/10.1057/ivs.2009.29) covers
+details of tree comparison and Aigner, et al. Visualization of Time-Oriented
+Data provides an overview of temporal data. This small sampling of a large body
+of work points to the need for a clearer conceptualization and clarification of
+the contribution of the work presented in this paper relative to prior efforts.
+
+\color{black}
 
@@
 set of ideas that are directly adapted in our work. Further, we have provided a
-bullet list in the introduction that explicitly describes the contribution of
+bullet list in the introduction that explicitly describes the contributions of
 this paper.
 
-\textit{My main concern regarding the presentation of the work lies in the third
-  and fourth views - why are they presented as distinct? It would seem to me
-  that all of the described functionality could be configured in that one view.
-  In fact, I could imagine including the DOi in the same view as well - why are
-  multiple views needed?}
+\color{blue}
 
+My main concern regarding the presentation of the work lies in the third and
+fourth views - why are they presented as distinct? It would seem to me that all
+of the described functionality could be configured in that one view. In fact, I
+could imagine including the DOi in the same view as well - why are multiple
+views needed?
+
+\color{black}
 The idea of combining these displays is interesting, and we have attempted to
@@
 querying and mouseover interactions for providing detail, the events conflict.
-Indeed, we implemented the first and third recommendations described in
-\href{https://github.com/d3/d3/issues/1604}{this D3 issue}, but in both cases we
+Indeed, we implemented the first and third recommendations described in this D3
+issue: https://github.com/d3/d3/issues/1604. However, in both approaches we
 found it necessary to introduce a lag between mousedown and the start of
@@
 
-\textit{Some detail about the scale of the tool would be useful. How many nodes
+\color{blue} Some detail about the scale of the tool would be useful. How many nodes
   can be included in the tree? How many time series, of what length? Rough
-  estimates would help readers assess utility.}
+  estimates would help readers assess utility.
 
+\color{black}
 We have added a discussion of scalability at the end of Section 1.4. We also
-explicitly list the problem dimensions in Table 1. Also, all examples (and more)
-are available as pre-compiled Rmarkdown vignettes linked from our package
-webpage, so users can assess utility in this examples easily without having to
-even install the package.
+explicitly list the problem dimensions in Table 1. Also, all case studies (as
+well as other examples) are available as pre-compiled Rmarkdown vignettes linked
+from our package webpage, so users can assess utility in this examples easily
+without having to even install the package.
+
+\color{blue} Finally, given the difficulty of building effective and usable
+  visualizations, the authors might note that the case studies are not
+  equivalent to an empirical validation of the tool through a user study. I see
+  no reason to believe that this tool wouldn't be highly usable, but
+  demonstrating that value is often harder than one might think.
+
+\color{black}
+We agree that a user study would provide data to improve design in the future,
+and have added a note to that end in the conclusion. However, we believe that
+such a study is out of scope for this work.
 
@@
 
-\textit{In the case studies, it would be interesting to see how the highlighted
+\color{blue} 1. In the case studies, it would be interesting to see how the highlighted
   nodes change when you shrink or pan the selecting area in the time series
-  plot.}
+  plot.
 
-This is visible online in our video demo. It is also possible to see this effect
-directly by interacting with precompiled versions in our compiled rmarkdown
-vignettes. We have further added a short supplementary note providing static
-figures describing this effect.
+\color{black}
+This is visible online in our video demo, which is now linked in the
+supplementary materials. It is also possible to see this effect directly by
+interacting with precompiled versions in our Rmarkdown vignettes, available on
+the treelapse home page.
 
-\textit{Is it possible to combine the DOI sankey idea with the timebox trees, so
-  that the time series can be conditioned on the categorical variable?}
+\color{blue} 2. Is it possible to combine the DOI sankey idea with the timebox trees,
+  so that the time series can be conditioned on the categorical variable?
 
-It would be possible to colo series according to different properties, or
+\color{black}
+It would be possible to color series according to different properties, or
 alternatively faceting different series into separate panels. We have these
-extensions in mind for a future version of the package, though these are not
+extensions in mind for a future version of the package, but they are not
 possible in the package as it currently stands.
 
-\textit{As the authors mentioned in the conclusion part, faceting is an
+\color{blue} 3. As the authors mentioned in the conclusion part, faceting is an
   important feature to be added in the interactive time series plot. Also,
   zooming in $y$-scale is needed when the lines are very intensive, like the
-  bottom of Figure 1.}
+  bottom of Figure 1.
 
-As mentioned above, faceting is a natural next step, but we have not implemented
-it yet. On zooming however, it is already possible to zoom the $y$-axis, using
-the pan-zoom widget.
+\color{black}
+We agree, faceting is a natural next step, but we have not implemented it yet.
+Regarding zooming however, it is already possible to zoom the $y$-axis, using
+the pan-zoom widget. This is visible in Figures 3, 18, and 20 of the revised
+draft, for example.
 
-\textit{In the case studies, when there are two colors in the time series plot
+\color{blue} 4. In the case studies, when there are two colors in the time series plot
   (Figures 4, 9, 13, 16), the brown color is almost hidden by the blue. It would
   be nicer to have the brown highlighted on the top, or add the transparency to
-  the blue.}
+  the blue.
 
+\color{black}
 It is already possible to customize the transparency of the blue selected lines
 by using the \texttt{ts\$opacity\_selected} argument in the \texttt{style\_opts}
-argument to the R plotting functions. However, we realize this is not the ideal
-approach to displaying the searched series, and since searching is an important
-function, we have revised our implementation and the relevant figures so that
-the searched series are always displayed on top.
+argument to the \texttt{timebox\_tree} and \texttt{treebox} functions. However,
+we realize this is not the ideal approach to displaying the searched series, and
+as suggested by the reviewer, we have revised our implementation and the updated
+figures so that the searched lines are always displayed on top.
 
-\textit{In the clustering example, it doesn't make sense to show the internal
-  nodes. It is better to display only the leaves in the time series plot.}
+\color{blue}
 
-We actually think that the nodes are useful to plot, because they can be
+5. Can you brush two blocks of nodes and compare the time series from two
+selected areas?
+
+\color{black}
+Yes, you can, this is displayed in Figure 17, for example. As it is, the colors
+of the series are the same regardless of which brush was applied to it (you
+could imagine a version with different colors for different brushes, but this is
+not currently implemented). However, it is relatively straightforwards to see
+which series are associated with a given brush when working interactively,
+because when moving one brush, the series associated with all others stay
+constant.
+
+\color{blue}
+
+6. In the clustering example, it doesn't make sense to show the internal nodes.
+It is better to display only the leaves in the time series plot.
+
+\color{black}
+We actually think that the nodes are useful to display, because they can be
 interpreted as centroids of the associated subtree, which are a natural object
-of study.
+of interest.
 
-\textit{Page 7, Line 46, ``node sizes reflecting the time series value at that
-  node''. Does that mean the node size will change when the time changes?}
+\color{blue}
 
-Sorry, what we actually meant is that the node size reflects the
-\textit{average} value across all times. This has been clarified in the text.
+In addition, there are some issues in the article:
 
-\textit{Page 11, lines 28-40. Should have referred to Figure 5.}
+\color{blue}
 
+1. Page 7, Line 46, ``node sizes reflecting the time series value at that
+node''. Does that mean the node size will change when the time changes?
+
+\color{black}
+What we actually meant is that the node size reflects the \textit{average} value
+across all times. This has been clarified in the text.
+
+\color{blue} 2. Page 11, lines 28-40. Should have referred to Figure 5.
+
+\color{black}
 We have added a reference to Figure 5 at the point that we start describing the
@@
 
-\textit{Page 13, Line 42. ``The red edges are associated with preterm births''.
-  Actually red represents term in Figure 8. Hence lines 45-47 are not correct.}
+\color{blue} 3. Page 13, Line 42. ``The red edges are associated with preterm
+  births''. Actually red represents term in Figure 8. Hence lines 45-47 are not
+  correct.
 
-We have fixed this in the text to refer to the green and yellow edges.
+\color{black}
+We have fixed this in the text so that we refer to the green and yellow edges.
 
-\textit{Page 15, Line 46. How did you get the numbers of the power, 5.6 and 6.7?
-  In Figure 13 the $y$-axis goes from around 11 to 15, and I can't find any
-  number close to 6.}
+\color{blue}
 
+4. Page 15, Line 46. How did you get the numbers of the power, 5.6 and 6.7? In
+Figure 13 the $y$-axis goes from around 11 to 15, and I can't find any number
+close to 6.
+
+\color{black}
 We had accidentally changed the base of the log between generating this figure
-and our text. The text and figure now both use base 10 logs.
+and writing our text. The text and figure now both use natural logs, and the
+short calculation for the price ranges has been updated according ot the new
+$y$-axis values.
 
-\textit{Some of the analysis from the figures are not grounded. For example,
-  page 14 lines 45-47 can not be concluded from Figure 9 easily. The analysis on
-  page 15 lines 37-39 are not directly from the plot, because Figures 11 and 12
-  themselves do not tell the county names.}
+\color{blue}
+5. Some of the analysis from the figures are not grounded. For example, page 14
+lines 45-47 can not be concluded from Figure 9 easily. The analysis on page 15
+lines 37-39 are not directly from the plot, because Figures 11 and 12 themselves
+do not tell the county names.
 
-We realize this is a difficulty of explaining inferences gathered from
-interactive visualizations through a sequence of static figures. In the text, we
-have now noted some of the interactive gestures the user would have to make to
-completely ground these analysis. For this issue of reproducibility, we have
-also made compiled interactive versions of these examples avaialble so that
-users can confirm the analysis for themselves, without having to install
-treelapse or rerun examples.
+\color{black}
+We have removed lines 45 - 47 on page 14. That conclusion is apparent when
+brushing up and down along the time series within that display, but we agree
+that it is not obvious from the single static image. If that finding is
+considered particularly interesting, we could create a sequence of images
+showing this effect. However, since we have another sequence with an analogous
+story in the subsequent figures, we have chosen to instead remove this
+discussion.
 
-\subsection{Associate Editor Review}
+For the analysis related to Figures 11 and 12, we have clarified that these
+county names are only visible upon hovering over the associated tree nodes. Further,
+we have added sentences explaining that the mouse is currently hovered over
+nodes of regions among the set that we claim have the effect, so that our
+statement is at least partially validated even from the static image alone.
+
+\subsection{Editor's comments to Author}
+
+\color{blue}
+
+Unorthodox structure: sections need numbering, and various pieces at the
+beginning would all be part of an introduction.
+
+\color{black}
+We have added numbering to the sections. Also, the subsections ``problem
+motivation'' through ``specific proposals'' are all contained within the
+introduction now.
+
+\color{blue}
+
+Captions need three pieces: (1) what the plot is about, (2) specific mapping of
+variables to graphical elements, like color, (3) a sentence that tells the
+reader what they should learn by looking at the plot.
+
+\color{black}
+We have added more detail to each caption, and structured them so that they
+include these three components.
+
+\color{blue}
+
+With interactive graphics, I would recommend that you add video illustrations.
+You can do this by loading the videos to a standard site like vimeo or youtube
+and providing the link in the paper pdf.
+
+\color{black}
+We have created a youtube video, which is now provided in the Supplementary
+Material. It describes the basic views available in treelapse, and walks through
+the antibiotics case study.
+
+We have also made precompiled vignettes available on the treelapse home page
+(linked in the Supplementary Materials), so that users can play with the actual
+interactive figures in this paper without installing the package.
+
+\color{blue} You have links embedded in the text. Make a supplementary materials
+  section, with a list of links instead.
+
+\color{black}
+We have added a supplementary materials section with a links subsection, which
+is not referred to from within the text.
+
+\color{blue}
+You have blinded the paper. I would suggest that this really doesn't work for
+you to get an adequate review. At this point the reviewers have favorably viewed
+the work, so the paper has passed the first stage of evaluation. The reviewers
+will need to be able to play with the software in the next review to give you
+useful feedback.
+
+\color{black}
+The revised version is no longer blinded, and we provided links to all the
+software and case studies. We look forwards to feedback on the actual software.
+
+\subsection{Associate Editor comments to the author}
+
+\color{blue}
+
+Please address the ``lack of context'' issues pointed out by reviewer 1 and the
+comments from both reviewers that relate to clarifying explanations in the text
+and/or figures.
+
+\color{black}
+We have added a more general literature review and revised our discussion of the
+figures to ensure analysis conclusions are visible from the figure (or have
+clarified within the text which other interactive steps are necessary), as
+detailed in the response to Reviewer 2's $5^{th}$ issue.
+
+\color{blue}
+
+The figures are not particularly attractive. You might consider making them a
+little larger so that individual features are more visible. It is also difficult
+to convey interactivity via static images; have you considered producing short
+video demonstrations for each case study?
+
+\color{black}
+We have increased the sizes of figures. Most had been 350px across, and are now
+375px wide.
+
+We have also created a video demo, linked in the supplementary material, but the
+focus is on the antibiotics case study. If the editors think it is necessary to
+have similar videos for the other case studies, we could create create them,
+though we hope it would be enough to include one demo video and links to the
+rest of the compiled vignettes.
 

### 2017-07-13 treelapse.tex
increase figure sizes

@@
 
-{\centering \includegraphics[width=350px]{figure/annotated_antibiotic_overview}
+{\centering \includegraphics[width=375px]{figure/annotated_antibiotic_overview}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/antibiotic_bacteroidetes}
+{\centering \includegraphics[width=375px]{figure/antibiotic_bacteroidetes}
 
@@
 \begin{figure}
-{\centering \includegraphics[width=350px]{figure/antibiotic_firmicutes}
+{\centering \includegraphics[width=375px]{figure/antibiotic_firmicutes}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/ruminococcus}
+{\centering \includegraphics[width=375px]{figure/ruminococcus}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/verrucomicrobiae}
+{\centering \includegraphics[width=375px]{figure/verrucomicrobiae}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/preterm_csts}
+{\centering \includegraphics[width=375px]{figure/preterm_csts}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/preterm_csts_lacto}
+{\centering \includegraphics[width=375px]{figure/preterm_csts_lacto}
 
@@
 Section \ref{sec:supp_links} of the supplementary materials. In our display, we
-have taken the base 10 log of these indexes. As our hierarchical structure, we
+have taken the natural log of these indexes. As our hierarchical structure, we
 use each neighborhood's assignment to state, regional, county, and city levels.
@@
 
-We first study variation in prices at the city level. These series can be
-highlighted using a single brush in treeboxes, since each spatial level is
-displayed at the same height in the tree. The resulting display is given in
-Figure \ref{fig:zillowcities}. From a high level, the housing bubble, decrease
-in prices due to the recession, and subsequent recovery are readily apparent.
-The city-level view makes it clear that not all neighborhoods were equally
-affected by the recession -- richer cities plateaued at their 2008 prices,
-middle-income cities saw moderate decreases, and poorer cities saw the most
-significant declines in home prices. These more significant declines tended to
-occur in cities that had recently seen rapid increases during the housing
-bubble. Finally, we note that the range in home prices seems to have increased
-after the recession, speaking to the differential long-term effect of the
-recession on prices.
-
-Alternatively, we can study the trajectories of home prices among neighborhoods,
+For example, we can study the trajectories of home prices among neighborhoods,
 conditional on their being middle-income before the recession. We generate the
@@
 period. This is the content of Figures \ref{fig:zillowmiddleup} and
-\ref{fig:zillowmiddledown}. Inspecting the highlighted tree nodes associated
-with these series, we find that most of the middle-income series that increased
+\ref{fig:zillowmiddledown}. Though not directly visible from the static figures,
+hovering the mouse over the highlighted tree nodes provides the geographic
+identities, and we find that most of the middle-income series that increased
 after the recession are associated with middle-income neighborhoods within the
-coastal Southern California counties. In contrast, those middle income series
-that saw decreases were mostly located in Central California and Oakland.
+coastal Southern California counties. For example, the mouse is currently over a
+subtree with many highlighted nodes, which is shown to be the Los Angeles-Long
+Beach-Anaheim metropolitan area. In contrast, hovering over nodes associated
+with those middle income neighborhoods that saw decreases indicates that they
+were mostly located in Central California and Oakland. In Figure
+\ref{fig:zillowmiddledown}, the mouse is positioned over the Sacramento (which
+is located in Central California) subtree, and seems enriched for this subset of
+strongly recession-affected series.
 
@@
 series in San Francisco County. We see that, in 2016, prices range from around
-\(10^{5.6} \approx \$400,000\) to \(10^{6.7} \approx \$5 \text{ million}\). So,
+$e^{13} \approx \$440,000$ to $e^{14.5} \approx 2 \text{ million}$. So,
 while all these neighborhoods tend to be among the more expensive ones in
@@
 
-{\centering \includegraphics[width=350px]{figure/zillow_cities}
-
-}
-
-\caption{California home prices at the city level, between 1996 and 2016. Here,
-  we provide a treebox display, as in Figure \ref{fig:antibioticfirmicutes}, but
-  of the California home prices data. In this case, the tree corresponds to a
-  geographic hierarchy, with regions at the top and neighborhoods at the bottom, while
-  time series correspond to home prices collected weekly between 1996 and 2016.
-  A brush is drawn over all nodes at the city level. The effect of the 2008
-  recession is clear, and we have hovered over the San Bernardino series, to
-  display the identity of one of the cities most strongly affected by the
-  recession. This provides an example of a dataset with meaningful (spatial)
-  hierarchical structure that might not initially be studied using trees.
-\label{fig:zillowcities}}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=350px]{figure/zillow_middle_pre}
-
-}
-
-\caption{Neighborhoods with mid-range home prices before the recession are
-  brushed here, using a timebox tree version of Figure \ref{fig:zillowcities}.
-  The associated tree nodes are highlighted at the top. Note that the collection
-  of series seems to widen after 2008 -- we are interested in whether there are
-  reliable predictors of these alternate trajectories, given their similar
-  starting points. This serves as a baseline with which to compare Figure
+{\centering \includegraphics[width=375px]{figure/zillow_middle_pre}
+
+}
+
+\caption{The time series here represent California neighborhood home prices
+  between 1996 and 2016, and the tree corresponds to a geographic hierarchy,
+  with regions at the top and neighborhoods at the bottom. We have brushed the
+  neighborhoods with mid-range home prices before the recession. The associated
+  tree nodes are highlighted at the top. Note that the collection of series
+  seems to widen after 2008 -- we are interested in whether there are reliable
+  predictors of these alternate trajectories, given their similar starting
+  points. This serves as a baseline with which to compare Figure
   \ref{fig:zillowmiddleup} -- these views are easy to transition between
@@
 
-{\centering \includegraphics[width=350px]{figure/zillow_middle_up}
+{\centering \includegraphics[width=375px]{figure/zillow_middle_up}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/zillow_middle_down}
+{\centering \includegraphics[width=375px]{figure/zillow_middle_down}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/zillow_sf}
+{\centering \includegraphics[width=375px]{figure/zillow_sf}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/working}
+{\centering \includegraphics[width=375px]{figure/working}
 
@@
   \centering
-  \includegraphics[width=350px]{figure/weekend}
+  \includegraphics[width=375px]{figure/weekend}
     \caption{By adjusting the two brushes in Figure \ref{fig:working}, we see that
@@
   \centering
-  \includegraphics[width=350px]{figure/weekday_2011}
+  \includegraphics[width=375px]{figure/weekday_2011}
     \caption{Weekday demand appears larger in 2012 than 2011 -- compare with Figure 
@@
   \centering
-  \includegraphics[width=350px]{figure/weekday_2012}
+  \includegraphics[width=375px]{figure/weekday_2012}
   \caption{Weekday demand increased in 2012 -- compare with 
@@
 
-{\centering \includegraphics[width=350px]{figure/warm_weekend}
+{\centering \includegraphics[width=375px]{figure/warm_weekend}
 
@@
 
-{\centering \includegraphics[width=350px]{figure/bike_sankey}
+{\centering \includegraphics[width=375px]{figure/bike_sankey}
 
@@
 
-{\centering \includegraphics[width=250px]{figure/structssi}
+{\centering \includegraphics[width=325px]{figure/structssi}
 
@@
 allow graph layouts instead of trees in time and treebox displays, for data that
-cannot be coerced into a hierarchical structure. Finally, if these ideas turn
+cannot be coerced into a hierarchical structure. Further, if these ideas turn
 out to be useful in practice, it would be valuable to modularize the basic
@@
 community to construct novel linked, interactive graphics with minimal effort.
+Finally, formal quantitative assessments of interface design through a user
+study could guide changes that improve the experience of practitioners.
 
@@
 \item treelapse development page: \url{https://github.com/krisrs1128/treelapse}
+\item treelapse video demo: \url{https://www.youtube.com/watch?v=EcmYBRMVMbI&t=14s}
 \item Home prices data: \url{http://www.zillow.com/research/data/}

### 2017-09-22 treelapse.tex
some minor changes before resending

@@
 \begin{figure}
-
-{\centering \includegraphics[width=300px]{figure/preterm_preterm}}
-
+\centering
+\includegraphics[width=300px]{figure/preterm_preterm}
 \caption{Samples with high levels of phyla other than Firmicutes appear to be
@@
   hierarchical clustering provides such a structure which can then be
-  interpreted using treeboxes. }\label{fig:gptimebox}
+  interpreted using treeboxes.}\label{fig:gptimebox}
 \end{figure}
@@
 hierarchical clustering tree, after zooming to a few of the bacteria
-that distinguish between the clusters. Upon revisiting the original
-data, it becomes clear that the samples highlighted on the left come
-from freshwater samples, while those on the right come from soil and
-skin, and looking up taxonomic groups associated with the distinguishing
+that distinguish between the clusters. In contrast to the figures displayed to
+this point, we only print time series associated with observed samples -- the
+leaves of the hierarchical clustering tree. This reduces visual artifacts that
+can be created by plotting many similar internal nodes, and which can overwhelm
+patterns occuring in the leaves, which are those of central interest. Upon
+revisiting the original data, it becomes clear that the samples highlighted on
+the left come from freshwater samples, while those on the right come from soil
+and skin, and looking up taxonomic groups associated with the distinguishing
 bacteria confirms this. For example, many of the species with high

### 2017-09-22 reviewer_response_ii.tex
setup second round of reviewer responses

\documentclass{article}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\linespread{1.25}

\title{Response to Reviews: JCGS-17-033.R2}
\begin{document}
\maketitle

We would like to thank the reviewers for their encouragement and comments
leading to valuable improvements of our software and paper. The original
reviewer comments are printed in blue, and our responses are in black.

\subsection{Reviewer 1}
\label{subsec:reviewer_1}

\color{blue}
Comments to the Author

Thanks to the authors for having addressed concerns from the previous review.
The revisions made in response to these comments have, in my opinion,
strengthened the paper.

I als want to commend the details presented in the video and on the GitHub
pages. Very nicely done.

I have a few minor suggestions for a few additional revisions that might help
the paper.

1. Regarding the third and fourth view, your response to earlier comments
suggests that implementation details interfered with attempts to combine these
views. Discussing these concerns in the paper (or on the documentation website)
might help others either avoid such difficulties or identify appropriate
work-arounds. Please consider adding relevant text to the paper.

\color{black}

\color{blue}
2. On page 7, the description of the tree reads: "Every edge in the tree is
split into two colors1, with relative widths of the different colors reflecting
differences in sizes for the two groups. The overall width of each edge
represents the sum of values across all groups." I was a bit confused by these
sentences, as they seem redundant. Doesn't the second sentence follow directly
from the first? If not, some additional clarification is needed.

\color{black}

\color{blue}
3. Regarding the tree ordering, I found myself wondering about the reordering of
the trees based on the size of the subtree. As this seems to be to be a tradeoff
between information clarity and consistency, I can imagine some cases where this
behavior might be useful, and others where it might be distracting. Have you
considered the possibility of animating the transition, so was to help users
avoid disorientation? Alternatively, a toggle to enable/disable this behavior
might help.

\color{black}

\subsection{Reviewer 2}
\label{subsec:reviewer_2}

\color{blue}
Comments to the Author

The authors have made substantial changes to the
manuscript according to the comments from reviewers and editors.

However, I still have one concern about the display of hierarchical clustering.
The authors did answer the question about this example from the second reviewer:
"...the (internal) nodes are useful to display, because they can be interpreted
as centroids of the associated subtree..." I agree that in other examples and
some scenarios where the internal nodes serve as meaningful group dividers, like
the California home prices data, they are useful to display. But hierarchical
clustering is a different story. All the internal nodes are binary, and the
centroids of subtrees are not of interest especially for those lower-level
subtrees. Too many time series lines for the centroids may mislead or increase
the difficulty of user's understanding about the original data. For the Global
Patterns data, Figure 19 (left) highlights 11 nodes of which only 6 are the
leaves, i.e., the original samples. So there are almost half of the time series
lines are additional. Things will get worse when the subtree is unbalanced,
because the centroids will emphasize the larger subgroups and weaken the smaller
subgroups by drawing more time series lines for the larger subgroups.

To solve this problem, it would be good to provide the option to highlight
different levels of nodes. For example, if the user selects the lowest level,
only the original data points will appear.

\color{black}

\end{document}

### 2017-09-22 reviewer_response_ii.tex
response about only displaying leaves

@@
 
-The authors have made substantial changes to the
-manuscript according to the comments from reviewers and editors.
+The authors have made substantial changes to the manuscript according to the
+comments from reviewers and editors.
 
@@
 \color{black}
+We appreciate the reviewers elaboration of the problems that can arise when
+plotting series associated with internal nodes, and have taken the recommended
+steps to resolve the issue. In particular, for both timebox trees and treeboxes,
+we have added functionality to either (1) only display time series associated
+with leaves (original data samples) or (2) only display time series within a
+certain window of depths on the tree.
+
+We have updated Figure 19 (the hierarchical clustering example), so that only
+series associated with leaves are shown. The vignette for this example has also
+been updated, so that users can see how to access this functionality.
 

### 2017-09-22 treelapse.tex
responding to reviewer 1

--- 2017-07-13 treelapse.tex
@@
 
+In principle, it would be desirable to combine the timebox and treebox displays,
+so that highlighted nodes and series could be determined through brushes on both
+the tree and series. For example, it would be useful to highlight the samples
+that lie in the intersection of all timeboxes and union of all treeboxes.
+However, the authors have found practical implementation of this idea
+challenging.
+
 To be practically useful, the resulting visualizations must respond fluidly to

### 2017-09-23 reviewer_response_ii.tex
responding to reviewer 1
respond to rest of review

@@
 \color{black}
+We have added a brief discussion of the possibility of combining the third and
+fourth views in a paragraph following their introduction (``In principle, it
+would be desirable...''). We describe the difficulties we found, and provide a
+link to a github branch with the partial solution that we found unsatisfactory.
 
@@
 \color{black}
+It's true, those sentences are redundant. We realize now that this can be
+confusing, so we have removed the second sentence.
 
@@
 \color{black}
+We had not previously considered using orderings other than D3's default.
+However, we were able to incorporate the reviewer's suggestion by adding a
+``toggle layout'' button, which smoothly transitions between D3's default ordering
+(descending height and then ascending ID) and an ordering based on the number of
+total descendants. The online vignettes have been updated to include this new
+functionality.
 

### 2017-09-23 treelapse.tex
remove supplemental material

@@
 modeling to interactive exploration, and vice versa. Our code is open-source,
-and is linked in Section \ref{sec:supp_links} of the supplementary materials. We
-hope this package encourages data analysts to work at the border between data
-modeling and visualization, and more generally empowers a wider audience to
-apply less widely known, but powerful, visualization ideas.
+and is linked in the supplementary materials. We hope this package encourages
+data analysts to work at the border between data modeling and visualization, and
+more generally empowers a wider audience to apply less widely known, but
+powerful, visualization ideas.
 
@@
   \citep{brewer2003colorbrewer}}, with relative widths of the different colors
-reflecting differences in sizes for the two groups. The overall width of each
-edge represents the sum of values across all groups.
+reflecting differences in sizes for the two groups.
 
@@
 the tree and series. For example, it would be useful to highlight the samples
-that lie in the intersection of all timeboxes and union of all treeboxes.
-However, the authors have found practical implementation of this idea
-challenging.
+that lie in the intersection of all timeboxes and union of all treeboxes. This
+could allow more complex queries than are currently available. However, while
+conceptually appealing, the authors encountered obstacles in practical
+implementation: brush and mouseover events are required to occupy the same space
+in this combined view. Properly distinguishing these events can be challenging,
+and a solution based on the introduction of a lag between mousedown and the
+manipulation of a brush led to a deteriorated user experience\footnote{However,
+  the code for this approach is available publicly, in a separate branch of
+  treelapse:
+  \url{https://github.com/krisrs1128/treelapse/tree/combined-brushes}.}.
 
@@
 computed monthly between 1996 and 2016. A link to the data source is provided in
-Section \ref{sec:supp_links} of the supplementary materials. In our display, we
-have taken the natural log of these indexes. As our hierarchical structure, we
-use each neighborhood's assignment to state, regional, county, and city levels.
-We represent each of these coarser spatial categories using the average of all
-neighborhoods contained in them. We have filtered down to the 890 neighborhoods
-in California; rendering more neighborhoods while keeping all 246 timepoints
-causes a severe lag in the interface.
+the supplementary materials. In our display, we have taken the natural log of
+these indexes. As our hierarchical structure, we use each neighborhood's
+assignment to state, regional, county, and city levels. We represent each of
+these coarser spatial categories using the average of all neighborhoods
+contained in them. We have filtered down to the 890 neighborhoods in California;
+rendering more neighborhoods while keeping all 246 timepoints causes a severe
+lag in the interface.
 
@@
 originally collected by a Washington D.C.-based bikesharing system for use in a
-Kaggle prediction competition. A link is provided in Section
-\ref{sec:supp_links} of the supplementary materials. The data are hourly
-measurements of bike demand, aggregated across all bikesharing stations, over
-two years, along with supplemental weather data. In the competition,
-participants were asked to predict the hourly demand on a held-out test set.
-Here, we adopt a descriptive view instead, attempting to characterize factors
-associated with variation in bikesharing demand.
+Kaggle prediction competition. A link is provided in the supplementary
+materials. The data are hourly measurements of bike demand, aggregated across
+all bikesharing stations, over two years, along with supplemental weather data.
+In the competition, participants were asked to predict the hourly demand on a
+held-out test set. Here, we adopt a descriptive view instead, attempting to
+characterize factors associated with variation in bikesharing demand.
 
@@
 
-\section{Supplemental Materials}
-
-\subsection{Links}
-\label{sec:supp_links}
-
-\begin{itemize}
-\item treelapse home page: \url{https://krisrs1128.github.io/treelapse/}
-\item treelapse development page: \url{https://github.com/krisrs1128/treelapse}
-\item treelapse video demo: \url{https://www.youtube.com/watch?v=EcmYBRMVMbI&t=14s}
-\item Home prices data: \url{http://www.zillow.com/research/data/}
-\item Bikesharing data: \url{https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset}
-\end{itemize}
-
 \end{document}

### 2017-09-23 supplement.tex

\documentclass{article}
\usepackage{url}

\title{Supplemental Materials}
\author{Kris Sankaran}

\begin{document}
\maketitle

\section{Supplemental Materials}

\subsection{Links}
\label{sec:supp_links}

\begin{itemize}
\item treelapse home page: \url{https://krisrs1128.github.io/treelapse/}
\item treelapse development page: \url{https://github.com/krisrs1128/treelapse}
\item treelapse video demo: \url{https://www.youtube.com/watch?v=EcmYBRMVMbI&t=14s}
\item Home prices data: \url{http://www.zillow.com/research/data/}
\item Bikesharing data: \url{https://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset}
\end{itemize}

\end{document}
